### 2020-04-29 reports/article.Rmd
first draft discussed with NH

---
title: "Title"
author: X^1,2,3^, Y^2,4^, ... Z^5^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

```{r setup, include = FALSE}
options(
  digits         = 2,
  knitr.kable.NA = '',
  repos          = list(CRAN = "https://cran.rstudio.com/")
)

knitr::opts_chunk$set(
  echo       = FALSE,
  out.width  = "100%", 
  out.height = "100%"
)

library(tidyverse)
```

## Affiliations

1. Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
2. The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
3. Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
4. ...
5. ...

##### PAGE BREAK

# Abstract

**SUBTITLE 1** Paragraph 1 ...

**SUBTITLE 2** Paragraph 2 ...

**SUBTITLE 3** Paragraph 3 ...

##### PAGE BREAK

# Introduction

# Patients and Methods

We used data from two national quality registers 2008-2015: The Swedish hip arthroplasty register (SHAR), with a completeness of 96-98%,[@Karrholm2018] for model development and internal validation, and XXX for external validation.

SHAR was linked using Swedish personal identity numbers[@Ludvigsson2009] to enhance the variable set with education and civil status from Statistics Sweden,[@Ludvigsson2019] comorbidity and adverse events from the national patient register,[@Ludvigsson2011] and prescribed medications from the medical prescription register.[@Cnudde2016]
We excluded patients with hip arthroplasty due to fractures, tumours, unspecified, or unknown reasons. We included patients with either unilateral THA, or with their second staged bilateral hip.[@Bulow2020] Patients younger than 18 or older than 100 years were excluded, as were patients with a body mass index (BMI) above 50, or with missing data on BMI, ASA class, education, type of hospital, and cementation (Figure \@ref(fig:flowchart)).

Comorbidity, during one year before surgery, was based on the Swedish version of the 10th revsin of the international classification of diseases (ICD-10-SE), as recorded at any hospital physician appointment prior to surgery. Individual codes were then combined using Charlson and Elixhauser comorbidity groups.[@Quan2005] A similar procedure was performed for codes of the anatomical therapeutic chemical (ATC) classification, grouped by Rx Risk V.[@Pratt2018] Similar conditions were then combined to a broader classification scheme based on clinical relevance, and those categories were used as possible predictors (Table \@ref(tab:tabcategorization)).

Infections within 90 days and 2 years respectively, were identified by relevant ICD-10-SE codes, or by codes from the Classification of Surgical Procedures (NCSP) from the Nordic Medico-Statistical Committee (NOMESCO) recorded in the national patient register. An infection was also identified if causing a reoperation recorded to SHAR.

## Statistical analysis for model development

Infections within 90 days or 2 years were modelled with logistic regression, ignoring possi

We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA. Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales. We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed. Potential predictors with variable importance above this breaking point were considered important and kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison. This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]

## Statistical analysis for model validation

Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation). Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples. We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally. An AUC with 95 % confidence interval was calculated for the model as-is. Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales. An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.

## Statistical tools

We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC.

## Ethical approval

Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13). Informed consent was not mandatory according to the Swedish patient data law from 2009.

# Results

## Study participants

`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel). `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that. The unadjusted risk of 90-day mortality was therefore `r surv90d`. Further characteristics of the study population are presented in Table \@ref(tab:tab1). `r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser. The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.

In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel). Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.

## Model development and internal validation

```{r}
spell <- function(x)
  c("one", "two", "three", "four", "five", "six", 
    "seven", "eight", "nine", "ten", "eleven", "twelve")[x]
```

There were `r spell(nrow(comb_lgl))` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`. Those variables were therefore excluded as potential predictors prior to any statistical model derivation. The derived main model included `r coefs_print`. The reduced model, with covariates included at least 33 out of 100 times, was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).

There were no differences between models including age as a main effect, compared to those were age was modeled by restricted cubic splines with either two or three knots. We therefore focused on the more parsimonious models with age as a main effect. Similarly, the correction for optimism only affected the third decimals of the AUC confidence intervals and was therefore omitted.

```{r}
half_paran <- function(x) 
  gsub(
    "(\\(AUC = )(0.\\d*)(, )(95 % CI: 0.\\d{2} to 0.\\d{2}\\))", 
    "\\2 \\(\\4", 
    get_auc(x), 
    perl = TRUE
  )
```

The main and reduced models were no different regarding discriminative power `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`. We therefore considered the reduced model as superior due to its simplicity. Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocs) left panel and Figure \@ref(fig:aucci)).

The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:sep). Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 %, although with predicted probabilities usually higher than observed proportions (Figure \@ref(fig:calibration)).

Estimated model coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).

Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3 %, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.

## External validation

# Discussion

## Conclusion

##### PAGE BREAK

# Contribution of authors

# Acknowledgement

##### PAGE BREAK

```{r, child = "figs.Rmd"}
```

```{r, child = "tabs.Rmd"}
```

##### PAGE BREAK

# Bibliography

### 2020-04-29 reports/description.Rmd
first draft discussed with NH

---
title: "Infection and dislocation"
author: "Erik Bulow"
date: '2020-02-19'
output: 
  html_document:
     toc: true
     toc_float: true
     number_sections: true
     df_print: "kable"
     code_folding: "hide"
bibliography: H:/zotero_lib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source("../lib/clean_names.R")
model_data <- readRDS("../cache/model_data.rds")
```

# Note!

I will present both outcomes since it is easier for now. I will make separate figures and tables later for each publication separately.

# Data

Linked data from:

- SHAR: THA-related
- Census (folkbokforing): death
- National Patient Register: co-morbidity
- LISA: education, civil status.
- Prescription register: ATC codes for Rx Risk V

# Inclusion

```{r}
knitr::include_graphics("../graphs/flowchart.png")
```

# Comorbidity

- Charlson: 1 year prioir to surgery. ICD-10-codes as defined by @Quan2005 with index weights by @Charlson1987.
- Elixhauser: 1 year prioir to surgery. ICD-10-codes as defined by @Quan2005 with unweighted sum as index.
- Rx Risk V: 1 year prioir to surgery. ATC-codes and index as defined by @Pratt2018.

Individual groups from all above grouped as:

```{r, message=FALSE}
load("../cache/tab_categorization.RData")
mutate_all(tab_categorization, coalesce, "")
```

# Outcomes

Infection and dislocation as identified at reoperation in SHAR or from later recordings in NPR (ICD-10 or KVA).

Within 2 years and 90 das.

# Rare conditions

We only kept potential predictors with at least 10 patients with and without respective outcome.
Dropped variables are listed in the table. Most conditions were dropped due to too few patients with the condition who experienced the event ("event_cond"). Some conditions were so rare that there was not enough patients even without the event ("no event_cond").

```{r}

model_data %>% 
  select(outcome, time, lgl) %>% 
  unnest("lgl") %>% 
  group_by(outcome) %>% 
  mutate(time = replace(time, duplicated(time), "")) %>% 
  ungroup() %>% 
  mutate(outcome = replace(outcome, duplicated(outcome), ""))

```

We did the same check for individual levels of categorical variables:

```{r}

model_data %>% select(outcome, time, fct)
```

Hence, patients with childhood disease were excluded for prediction of infection within 90 days.
This, this model was trained on `r  model_data %>% filter(!is.na(fct)) %>% select(n_df0) %>% pluck(1)` patients.

# Summary tables

We make those for outcome within 2 years, but this could be modified!

## Infection

```{r}
load("../cache/table1_infection.RData")
table1_infection$table1
```

## Dislocation

```{r}
load("../cache/table1_dislocation.RData")
table1_dislocation$table1
```

# Models

The variable selection process is rerun 100 times. Some variables are always included (n = 100), some are less common.

```{r}
model_data %>% 
  select(outcome, time, coefs_selected) %>% 
  mutate(coefs_selected = map(coefs_selected, "brlasso_tbl_selected")) %>% 
  unnest("coefs_selected") %>% 
  group_by(outcome) %>% 
  mutate(time = replace(time, duplicated(time), "")) %>% 
  ungroup() %>% 
  mutate(outcome = replace(outcome, duplicated(outcome), ""))
```

We use all variables for the "main models" but only variables selected at least 75 times out of 100 are included in the reduced models.

It might be relevant however to decrease the cut-of to a lower number, perhaps 20? The reason is that the full model is superior to the reduced model but that it feels a little unnecessary to include variables that were seldom selected.

# Internal validation

## ROC-curves

We can see that the main model with age as RCS is superioir for all outcomes, although the main model with age as a main effect is not far behind for infection within 2 years.

```{r}
knitr::include_graphics("../graphs/roc.png")
```

## AUC

Unfourtenately, no model had AUC above 0.7. The main models seem preferable although the added value of age as RCS is not significant. Main effects model might be preffered due to simplicity?

```{r}
knitr::include_graphics("../graphs/auc_ci.png")
```

## Separation

There is not much separation between patients with and without each outcome (main models with age as main effect).

```{r}
knitr::include_graphics("../graphs/separation.png")
```

## Calibration

The 90 day infection model is not well calibrated, the 2 year model is!

```{r}
knitr::include_graphics("../graphs/calibration.png")
```

# Model parameters

Note that we have here included all levels from variables included with any level in the models. Hence all diagnoses, although some are not differnet from baseline. This should probably be changed to another baseline and without "others" etc. After that, we might merge some of the levels etc.

I include all models here since they can be looked at separately (compared to the graphs which would be very messy).

```{r}
load("../cache/tbl_coefs.RData")

tbl_coefs %>% 
  select(-Model, -math) %>% 
  group_by(outcome) %>%
  mutate(time = replace(as.character(time), duplicated(time), "")) %>% 
  ungroup() %>%
  mutate(
    outcome = replace(outcome, duplicated(outcome), ""),
    beta = round(beta, 2)
  )

```

# Bibliography

### 2020-04-29 reports/diagrupp.Rmd
first draft discussed with NH

---
title: "Diagnosgrupp"
author: "Erik Bulow"
date: '2020-04-27'
output: 
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
con <- shar_linkage_con()
```

# Källa

Jag lyckas tyvärr inte hitta någon definition av hur grupperingen av enskilda diagnoser går till.
Denna gruppering är redan gjord i det dataset jag har tillgång till. Jag kan dock se vilka kombinationer som förekommer rent empiriskt:

# Översättningar

Grupperingen nedan utgår från den svenska variabelnamn. Den översättning som användes är följande (behöver kanske korrigeras). Observera att endast inkluderade diagnosgrupper översatts och att det alltså finns fler i tabellen nedan:

```{r}
tibble(
  se = c(
       "Primär artros",
       "Annan sekundär artros",
       "Följdtillstånd efter barnsjukdom i höftleden",
       "Idiopatisk nekros",
       "Inflammatorisk ledsjukdom",
       "Komplikation eller följdtillstånd efter fraktur el annat trauma"
    ),
  en = c(
       "Primary osteoarthritis",
       "Secondary osteoarthritis",
       "Childhood disease",
       "Idiopathic necrosis",
       "Inflammatory joint disease",
       "Sequelae"
     )
  ) %>% 
  arrange(se)
```

# Gruppering

```{r}
diags_tmp <- 
  tbl(con, "primary") %>% 
  select(P_DiaGrp, P_Diagnosis) %>% 
  count(P_DiaGrp, P_Diagnosis, sort = TRUE) %>% 
  collect()

diags <- 
  diags_tmp %>% 
  ungroup() %>% 
  mutate_all(na_if, "NA") %>% 
  filter(!is.na(P_DiaGrp)) %>%
  arrange(P_DiaGrp) %>% 
  mutate(P_DiaGrp = replace(P_DiaGrp, duplicated(P_DiaGrp), ""))

diags
```

### 2020-04-29 reports/figs.Rmd
first draft discussed with NH

<!-- FIGURES -->

```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation (left). Data from XXX were used for external validation (right)."}
knitr::include_graphics("../graphs/flowchart.png")
```

##### PAGE BREAK

```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves."}
knitr::include_graphics("../graphs/roc.png")
```

##### PAGE BREAK

```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals. AUC above 0.7 were considered good."}

knitr::include_graphics("../graphs/auc_ci.png")
```

##### PAGE BREAK

```{r sep, fig.cap = "The vast majority of patients did not have an infection after THA, and therefore the blue bars dominate the histogram (upper panel; note the scales). Normalized density plots however reveals that patients who died within 90 days were, on average, estimated to have a higher probability to do so (lower panel)."}
knitr::include_graphics("../graphs/separation.png")
```

##### PAGE BREAK

```{r calibration, fig.cap = "This figure illustrates calibration between observed proportions and predicted probabilities with 95 % confidence intervals."}
knitr::include_graphics("../graphs/calibration.png")
```

### 2020-04-29 reports/tabs.Rmd
first draft discussed with NH

##### PAGE BREAK

```{r tabcategorization}
tab_categorization %>%
  ungroup() %>% 
  mutate(
    `Comorbidities by groups` = 
      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
  ) %>% 
  knitr::kable(
    caption = "Categorization of co-morbidities based on diagnostic groups according to Charlson and Elixhauser. (CNS = central nervous system.)")

```

##### PAGE BREAK

```{r tab1}

load("../cache/table1_infection.RData")

knitr::kable(
  mutate(table1_infection$table1, what = gsub("Heart infarct", "Myocardial infarction", what)), 
  caption = "Characteristics of the study population in the model derivation cohort (SHAR). Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years). BMI = Body mass index. ASA = American Society for Anesthesiologists classification. CNS = central nervous system.")
```

##### PAGE BREAK

```{r brlprop}
model_data <- readRDS("../cache/model_data.rds")

model_data %>% 
  select(outcome, time, coefs_selected) %>% 
  mutate(coefs_selected = map(coefs_selected, "brlasso_tbl_selected")) %>% 
  unnest("coefs_selected") %>%
  filter(outcome == "infection") %>% 
  select(-outcome) %>% 
  pivot_wider(names_from = "time", values_from = "n") %>% 
  mutate(
    variable = clean_names(variable) %>% 
      {gsub("\\.", " ", .)} %>% 
      {gsub("Rheumatidisease", "Rheumatic disease", .)} %>% 
      {gsub("Diagnos", "Diagnosis:", .)}
  ) %>% 
  select(Predictor = variable, `90 days` = `90d`, `2 years` = `2y`) %>% 
knitr::kable(
  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model. Variables chosen at least 75 times out of 100 were kept in the reduced model as well. (CNS = central nervous system.)")
```

##### PAGE BREAK

```{r brlreduced}

load("../cache/tbl_coefs.RData")

tbl_coefs %>% 
  filter(outcome == "infection") %>% 
  select(-outcome, -Model, -math) %>% 
  mutate(
    time = replace(as.character(time), duplicated(time), ""),
    beta = round(beta, 2)
  ) %>% 
knitr::kable(
  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced model.'
)
```

### 2020-04-29 data/linkage.R

@@ -1,0 +1,80 @@
+lisa <- tbl(con, "lisa_1yr_before") %>%
+  select(LopNr, P_Side, civil_status, education)
+
+sql_shpr_orig <-
+  tbl(con, "primary") %>%
+  select(
+    LopNr,
+    P_Side,
+    DateOfDeath,
+    P_TypeOfHospital,
+    P_ProstType,
+    P_SurgDate,
+    P_Gender,
+    P_ASA,
+    P_BMI,
+    P_Age,
+    P_DiaGrp,
+    P_KVA1,
+    P_AcetCupCemMix,
+    P_FemStemCemMix
+  ) %>%
+  # NOTE! from 2006 to identify if second THA within two years (which should be exluded)
+  filter(between(P_SurgDate, "2006-01-01", "2015-12-31")) %>%
+  left_join(tbl(con, "operations_factors_opnr"),  c("LopNr", "P_Side")) %>%
+  left_join(tbl(con, "elix_1yr_before"),          c("LopNr", "P_Side")) %>%
+  left_join(tbl(con, "charlson_1yr_before"),      c("LopNr", "P_Side")) %>%
+  left_join(tbl(con, "rxriskv_pratt_1yr_before"), c("LopNr", "P_Side")) %>%
+  left_join(lisa, c("LopNr", "P_Side"))
+
+df_shpr_orig <-
+  sql_shpr_orig %>%
+  collect() %>%
+  mutate_if(is.character, na_if, "NA") %>%
+  mutate_at(vars(DateOfDeath, P_SurgDate), as.Date, format = "%Y-%m-%d") %>%
+  mutate_at(vars(LopNr, P_ASA, P_Age), as.integer) %>%
+  mutate_at(vars(P_BMI), as.numeric) %>%
+  mutate_at(vars(matches("elix|charlson|rxriskv"), -contains("index")), as.logical) %>%
+  mutate_if(is.character, as.factor)
+
+cache("df_shpr_orig")
+
+
+
+# Reoperation -------------------------------------------------------------
+
+# Needs reoperation if reason is dislocation or infection
+# Should also filter on time (< 2 years from primary)
+df_reops <-
+  sql_shpr_orig %>%
+  inner_join(tbl(con, "reoperations"), c("LopNr", "P_Side")) %>%
+  filter(
[…]

### 2020-04-29 lib/clean_names.R

@@ -1,0 +1,41 @@
+clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
+  x <-
+    gsub("(ECI|[cP])_|_?TRUE.?|_X|TypeOf", "", x) %>%
+    tolower() %>%
+    {gsub("_", " ", .)} %>%
+    {gsub("asa2", "ASA grade = 2", .)} %>%
+    {gsub("asa3", "ASA grade = 3", .)} %>%
+    {gsub("gender ?man", "male sex", .)} %>%
+    {gsub("diagrp", "Diagnos", .)}
+
+
+  if (!lvls) {
+    x <- unique(gsub("ASA grade = [23]", "ASA grade", x))
+    x <- gsub("male sex", "sex", x)
+  }
+  if (firstupper) {
+    x <- paste0(toupper(substr(x, 1, 1)), substring(x, 2))
+  }
+
+  x <-
+    ifelse(
+      grepl("bmi|asa|eci|cci", x, ignore.case = TRUE),
+      toupper(x),
+      x
+    )
+
+  x <- ifelse(grepl("aids", x, ignore.case = TRUE), "AIDS/HIV", x)
+  x <- ifelse(grepl("cns", x, ignore.case = TRUE), "CNS disease", x)
+  gsub("GRADE", "class", x, ignore.case = TRUE)
+}
+
+
+model_names <- function(x, age = FALSE) {
+  if (!age)
+    x <- gsub(" \\(age as main effect\\)", "", x)
+
+  x <- gsub("ASA", "ASA class", x)
+
+  gsub("BRL reduced", "Reduced model", x) %>%
+    {gsub("BRL", "Main model", .)}
+}

### 2020-04-29 lib/globals.R

@@ -1,0 +1,24 @@
+con <- shar_linkage_con()
+
+# General settings
+memory.limit(1e10)
+set.seed(132456798)
+future::plan("multiprocess")
+
+options(
+  digits    = 2,
+  na.action = "na.fail",
+  scipen    = 999
+)
+
+# Add any project specific configuration here.
+add.config(
+  apply.override = FALSE
+)
+
+# Add project specific configuration that can be overridden from load.project()
+add.config(
+  N_bots         = 100,
+  Bmax           = 100,     # No of external reruns for BRLasso
+  apply.override = TRUE
+)

### 2020-04-29 lib/lasso.R

@@ -1,0 +1,14 @@
+# Apply lasso regression with CV for best lambda and return coeficient estimates
+lasso <- function(x) {
+  fit <- glmnet::cv.glmnet(
+    select(x, -outcome) %>% as.matrix(),
+    x$outcome,
+    family = "binomial"
+  )
+  as_tibble(
+    as.matrix(coef(fit$glmnet.fit, s = fit$lambda.min)),
+    rownames = "variable"
+  ) %>%
+    rename(coef = `1`) %>%
+    filter(variable != "(Intercept)")
+}

### 2020-04-29 lib/misc.R

@@ -1,0 +1,7 @@
+zero <- function(x) {
+  gsub("0 ( 0.0)", "0", x, fixed = TRUE) %>%
+    {gsub("0.0", "<0.1", .)}
+}
+
+# Cached manually and not loaded by default
+model_data <- readRDS("cache/model_data.rds")

### 2020-04-29 lib/strip_glm.R

@@ -1,0 +1,31 @@
+# Borrowed function from
+# Author: Nina Zumel
+# http://www.win-vector.com/blog/2014/05/trimming-the-fat-from-glm-models-in-r/
+# Accessed: 2019-10-01
+#
+# Trim gml object to smallest possible size to still alow for predictions
+# I do not want to include any data that is not needed for the Shiny application.
+strip_glm = function(cm) {
+  cm$y = c()
+  cm$model = c()
+
+  cm$residuals = c()
+  cm$fitted.values = c()
+  cm$effects = c()
+  cm$qr$qr = c()
+  cm$linear.predictors = c()
+  cm$weights = c()
+  cm$prior.weights = c()
+  cm$data = c()
+
+
+  cm$family$variance = c()
+  cm$family$dev.resids = c()
+  cm$family$aic = c()
+  cm$family$validmu = c()
+  cm$family$simulate = c()
+  attr(cm$terms,".Environment") = c()
+  attr(cm$formula,".Environment") = c()
+
+  cm
+}

### 2020-04-29 munge/00-filter.R

@@ -1,0 +1,263 @@
+# Pre-filter: some pat ids have same hip primarly operated twice!
+dubbles <-
+  df_shpr_orig %>%
+  select(LopNr, P_Side, P_SurgDate) %>%
+  add_count(LopNr, P_Side, sort = TRUE) %>%
+  filter(n > 1)
+
+
+# Patients with second hip operated dwithin two years after the first one
+within2yrs <-
+  df_shpr_orig %>%
+  anti_join(dubbles, "LopNr") %>%
+  select(LopNr, opnr, P_SurgDate) %>%
+  pivot_wider(names_from = opnr, values_from = P_SurgDate) %>%
+  filter(!is.na(`1`), !is.na(`2`)) %>%
+  mutate(timediff = as.numeric(`2` - `1`)) %>%
+  filter(timediff < 2 * 365) %>%
+  mutate(within2yrs = TRUE)
+
+# Add column "withni2yrs" to identify patients with their 2nd op withni 2 years
+df_shpr_tmp <-
+  df_shpr_orig %>%
+  left_join(within2yrs, "LopNr")
+
+
+
+# Define filter steps -----------------------------------------------------
+
+# Make tibble with one row per filter step and columns:
+# step: Number identifying inclusion box.
+#       Does not need to be unique if several exlusions are combined.
+# incl: Description of included cases
+# excl: Description of cases to exclude
+# expr: quosure defining filter step
+# data: The data set to be filtered. Only required for initial step (first row).
+
+filters <-
+  bind_rows(
+    tibble(
+      step = 0,
+      incl = "all",
+      data = list(df_shpr_tmp)
+    ),
+
+    tibble(
+      step = 1,
+      excl = "Hemi- and surface prosthesis",
+      incl = "THA 2008-2015",
+      expr = list(quo(
+        # 2007-2008 previously included to identify THAs within two years
+        P_SurgDate >= as.Date("2008-01-01") &
+        P_ProstType == "Totalprotes" &
+        (is.na(P_KVA1) | P_KVA1 != "NFB62 - Primär total ytersättningspr")
+      ))
+    ),
+
+    tibble(
+      step = 2,
+      excl = "Diagnose: fracture/tumour/unspecified/unknown.",
+      incl = "Elective THA",
+      expr = list(quo(!P_DiaGrp %in% c(
+        "Akut trauma, höftfraktur",
+        "Akut trauma, övriga",
+        "Tumör",
+        "Övrigt")))
[…]

### 2020-04-29 munge/01-outcome.R

@@ -1,0 +1,94 @@
+
+# Outcome from NPR --------------------------------------------------------
+
+df_kva <-
+  df_kva_orig %>%
+  mutate(
+    id = as.character(LopNr),
+    code_date = as.Date(code_date, origin = "1970-01-01")
+    ) %>%
+  select(id, code, code_date) %>%
+  as.codedata()
+
+df_icd10 <-
+  df_icd10_orig %>%
+  mutate(code_date = as.Date(code_date, origin = "1970-01-01")) %>%
+  select(id = LopNr, code, code_date) %>%
+  as.codedata()
+
+ae_npr <- function(df_npr, regex, endday) {
+  df_shpr %>%
+    select(LopNr, P_SurgDate) %>%
+    mutate(LopNr = as.character(LopNr)) %>%
+    categorize(
+      df_kva,
+      cc = "hip_ae_hailer",
+      id = "LopNr",
+      codify_args = list(
+        date = "P_SurgDate",
+        days = c(0, endday)
+      ),
+      cc_args = list(
+        regex = regex,
+        tech_names = TRUE
+      )
+    )
+}
+
+outcome_kva_90d <- ae_npr(df_kva,   "regex_kva",   90)
+outcome_icd_90d <- ae_npr(df_icd10, "regex_icd10", 90)
+outcome_kva_2y  <- ae_npr(df_kva,   "regex_kva",   2 * 365)
+outcome_icd_2y  <- ae_npr(df_icd10, "regex_icd10", 2 * 365)
+
+# Outcome from SHAR -------------------------------------------------------
+outcome_reop <- function(endday) {
+  df_reops %>%
+    mutate_at(vars(P_SurgDate, R_SurgDate), as.Date) %>%
+    filter(R_SurgDate - P_SurgDate < endday) %>%
+    distinct(LopNr, P_Side, .keep_all = TRUE) %>%
+    transmute(
+      LopNr, P_SurgDate,
+      reop_luxation =
+        IND_ReSurgReason1 == "Luxation, instabilitet, subluxation" |
+        IND_ReSurgReason2 == "Luxation, instabilitet, subluxation",
+      roep_infektion =
+        IND_ReSurgReason1 == "Infektion" |
+        IND_ReSurgReason2 == "Infektion"
+    )
+}
+
+outcome_reop_90d <- outcome_reop(90)
+outcome_reop_2y  <- outcome_reop(2 * 356.241)
+
+
[…]

### 2020-04-29 munge/02-survdata.R

@@ -1,0 +1,86 @@
+# Add important variables to data
+
+df <-
+  df_shpr %>%
+  mutate_if(is.logical, coalesce, FALSE) %>%
+  mutate_at(vars(contains("index")), coalesce, 0) %>%
+  mutate(
+    P_SurgYear = as.numeric(substr(P_SurgDate, 1, 4)),
+    P_ASA = factor(P_ASA, 1:3, c("I", "II", "III")),
+
+    # Civilstånd
+    civil_status = as.character(civil_status),
+    civil_status =
+      ifelse(
+        civil_status %in% c("divorced", "unmarried"),
+        "single", civil_status) %>%
+      as.factor() %>% relevel("married"),
+
+    education =
+      factor(education, levels(education), c("low", "middle", "high")),
+    P_TypeOfHospital =
+      factor(P_TypeOfHospital, levels(P_TypeOfHospital),
+        c("University", "County", "Rural", "Private"),
+      ),
+    P_Sex = factor(P_Gender, c("Man", "Kvinna"), c("Male", "Female")),
+
+    # Survival
+    stime   =
+      as.numeric(coalesce(DateOfDeath, as.Date("2018-02-01")) - P_SurgDate),
+    status  = !is.na(DateOfDeath),
+
+    # truncate comorbidity indices
+    charlson_icd10_index_quan_original = pmin(charlson_icd10_index_quan_original, 4),
+    elix_icd10_index_sum_all           = pmin(elix_icd10_index_sum_all, 3),
+    rxriskv_index_pratt                = pmin(rxriskv_index_pratt, 10),
+
+    # fixation
+    cemented_stem = P_FemStemCemMix != "Cementfritt",
+    cemented_cup  = P_AcetCupCemMix != "Cementfritt",
+
+    # Diagnos
+    P_DiaGrp = factor(droplevels(P_DiaGrp),
+      c(
+         "Primär artros",
+         "Annan sekundär artros",
+         "Följdtillstånd efter barnsjukdom i höftleden",
+         "Idiopatisk nekros",
+         "Inflammatorisk ledsjukdom",
+         "Komplikation eller följdtillstånd efter fraktur el annat trauma"
+      ),
+      c(
+         "Primary osteoarthritis",
+         "Secondary osteoarthritis",
+         "Childhood disease",
+         "Idiopathic necrosis",
[…]

### 2020-04-29 munge/03-compositvariabler.R

@@ -1,0 +1,45 @@
+# Förbered kompisitvariabler som ev kan ersätta enskilda diagnoser
+
+# Fulfix!
+# Fel i coder-paketet döpte "malignancy" till "malingnancy"
+# Ska funka korrekt med senaste coder-version men vill inte köra om just nu!
+df <- rename(df, CCI_malignancy = CCI_malingnancy)
+
+
+# Make mutate statments from Excel-file -----------------------------------
+
+# Define it!
+categorization <-
+  categorization.Blad1 %>%
+  mutate_all(zoo::na.locf) %>%
+  mutate_at(vars(new, old), ~ gsub("/| ", "_", tolower(.))) %>%
+  mutate(
+    new = paste0("c_", new),
+  ) %>%
+  unite("old", from, old) %>%
+  group_by(new) %>%
+  summarise(old = paste(old, collapse = " | ")) %>%
+  ungroup() %>%
+  mutate(
+    new = syms(new),
+    old = parse_exprs(old)
+  )
+
+cache("categorization")
+
+# Do it!
+c_cols <- bind_cols(pmap(categorization, ~transmute(df, !!.x := !!.y)))
+
+
+# Combine some existing columns with new ones --------------------------------
+
+df <-
+  df %>%
+  select(
+    everything(),
+    -matches("^([CE]CI)|(Rx)_"),
+    matches("index")
+  ) %>%
+  bind_cols(c_cols)
+
+cache("df")

### 2020-04-29 munge/04-remove_empty_variables.R

@@ -1,0 +1,134 @@
+# We will not consider any variable without observations in any combination
+# of dead/alive-effected/not affected.
+# This does not exclude any cases. It only exclude potential predictor variables
+
+df0 <-
+  df %>%
+  select(
+    -stime, -status,
+    -contains("index"),
+    -starts_with("outcome")
+  )
+
+
+rare_conditions <- function(df, outcome, minimum = 10) {
+  df %>%
+  select(starts_with("c_")) %>%
+  add_column(outcome = outcome) %>%
+  mutate(outcome = factor(outcome, c(TRUE, FALSE), c("event", "no event"))) %>%
+  group_by(outcome) %>%
+  summarise_if(
+    is.logical,
+    list(
+      .cond    = ~sum(.),
+      .no_cond = ~ n() - sum(.))
+  ) %>%
+  pivot_longer(
+    starts_with("c_"),
+    names_to = c("variable", "condition"),
+    names_sep = "_\\."
+  ) %>%
+  pivot_wider(names_from = c(outcome, condition)) %>%
+  filter_at(vars(-variable), any_vars(. < minimum))
+}
+
+# Names of conditions to drop to present in text
+comb_lgl_text <- function(comb_lgl) {
+  comb_lgl$variable %>%
+  clean_names(FALSE) %>%
+  glue_collapse(", ", last = " and ")
+}
+
+
+comb_lgl <-
+  tibble(
+    outcome = c("dislocation", "infection"),
+    time    = c("90d", "2y")
+  ) %>%
+  expand(outcome, time) %>%
+  mutate(
+    outcome_var = paste0("outcome_", outcome, "_", time),
+    lgl = map(outcome_var, ~ rare_conditions(df0, df[[.]])),
+    lgl_vars = map(lgl, ~.$variable),
+    lgl_text = map_chr(lgl, comb_lgl_text)
+  )
+
+
+# Same for factor variables -----------------------------------------------
+
+comb_fct_tmp <-
+  df0 %>%
+  select_if(is.factor)
+
+n_lvls <-
+  comb_fct_tmp %>%
+  summarise_all(n_distinct) %>%
+  pivot_longer(everything(), values_to = "n_lvls")
+
+# Factor variables with less than 10 observations per outcome
+comb_fct <- function(y) {
+  comb_fct_tmp %>%
+  pivot_longer(everything()) %>%
+  nest(data = value) %>%
+  left_join(n_lvls, "name") %>%
[…]

### 2020-04-29 munge/05-BRLasso.R

@@ -1,0 +1,19 @@
+# Perform calculations (time consuming)
+
+best_coefs_fun <- function(df_model) {
+  tibble(B = seq_len(config$Bmax)) %>%
+  mutate(
+    coefs_all = future_map(B, ~ BR_lasso_coefs(df_model), .progress = TRUE),
+    breaks    = map_dbl(coefs_all, break_p),
+    coefs     = map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
+  )
+}
+
+
+model_data <-
+  model_data %>%
+  mutate(
+    best_coefs_tmp = map(df_model, best_coefs_fun)
+  )
+
+cache("model_data")

### 2020-04-29 munge/06-variables_differnet_BRLasso_models.R

@@ -1,0 +1,37 @@
+
+# Table to report with no of selections per variable
+brlasso_coefs_selected <- function(best_coefs_tmp) {
+
+  brlasso_tbl_selected <-
+    best_coefs_tmp %>%
+    unnest(coefs) %>%
+    count(variable, sort = TRUE)
+
+  # List of variables selected every time
+  best_coefs_reduced <-
+    brlasso_tbl_selected %>%
+    filter(n >= round(.75 * config$Bmax)) %>% # Urspr föreslogs intersect (variabler som tas varje gång)
+    select(variable) %>%
+    pluck(1)
+
+  # All variables ever selected
+  best_coefs <-
+    brlasso_tbl_selected %>%
+    select(variable) %>%
+    pluck(1)
+
+  # Return
+  list(
+    brlasso_tbl_selected = brlasso_tbl_selected,
+    best_coefs_reduced = best_coefs_reduced,
+    best_coefs = best_coefs
+  )
+}
+
+model_data <-
+  model_data %>%
+  mutate(
+    coefs_selected = map(best_coefs_tmp, brlasso_coefs_selected)
+  )
+
+cache("model_data")

### 2020-04-29 munge/07-compare_models.R

@@ -1,0 +1,139 @@
+df <-
+  df %>%
+  mutate(
+    P_SurgYear                = P_SurgYear - min(P_SurgYear),
+    outcome_dislocation_90d_f = as.factor(outcome_dislocation_90d),
+    outcome_infection_90d_f   = as.factor(outcome_infection_90d),
+    outcome_dislocation_2y_f  = as.factor(outcome_dislocation_2y),
+    outcome_infection_2y_f    = as.factor(outcome_infection_2y),
+  )
+
+# "unregularized least-square fit restricted to variables in J" /[@Bach2008]
+form <- function(nms) {
+  paste("outcome ~", paste(unique(gsub(
+    "_TRUE.|_X[23]|_(Man|Kvinna|Male|Female|married|single|widow.widower|Idiopathic.necrosis|Inflammatory.joint.disease|Sequelae)", "",
+    nms)), collapse = " + ")
+  )
+}
+
+# Splines for age - Har testat med både 2 och 3 knots!
+# Ingen skillnad så behåller med 3
+fns3 <- function(coefs) {
+  paste0(form(setdiff(coefs, "P_Age")), " + splines::ns(P_Age, 3)")
+}
+
+
+glmdf <- function(f, outcome){
+
+  df$outcome <- df[[outcome]]
+  glm <-  glm(f, data = df, family = binomial())
+
+  # I tried to use LASSO to avoid over-estimation but found that best penalty-terms were virtually 0 for all models.
+  # I describe this as a sensitivity analysis, but I will not use it any further.
+  lambda <- purrr::possibly(glmnet::cv.glmnet, list(lambda.min = 0))(
+    model.matrix(as.formula(f), df)[, -1, drop = FALSE],
+    df$outcome,
+    family = "binomial"
+  )$lambda.min
+  lrm <-  rms::lrm(as.formula(f), data = df, x = TRUE, y = TRUE, penalty = lambda)
+
+  list(glm = glm, lrm = lrm)
+}
+
+
+compare_models <- function(brlasso_coefs, outcome) {
+
+  best_coefs         <- brlasso_coefs$best_coefs
+  best_coefs_reduced <- brlasso_coefs$best_coefs_reduced
+
+  all_models_tmp <-
+    tribble(
+      ~Model,                                            ~tab, ~fig,   ~fit,
+      "BRL reduced (age as main effect)",                TRUE,  TRUE,  glmdf(form(best_coefs_reduced), outcome),
[…]

### 2020-04-29 src/brier_score.R

@@ -1,0 +1,31 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# See @Steyerberg2010 for scaled Brier score
+brier_scaled <- function(brier, p_mean) 1 - brier / (p_mean * (1 - p_mean))
+
+brier <- function(all_models) {
+  all_models %>%
+  mutate(
+
+    # Version corrected for optimism
+    brier.corrected = map_dbl(lrm_validate, ~.["B", "index.corrected"]),
+    p_mean = map_dbl(obspred, ~mean(.$pred)),
+
+    # Scaled version corrected for optimism
+    brierscaled = map2_dbl(brier.corrected, p_mean, brier_scaled)
+  ) %>%
+
+  # All scores are approximately equal
+  # The score does not work however for very rare events, see:
+  # https://journals.ametsoc.org/doi/full/10.1175/2009MWR2945.1
+  # We will therefore not report it, just keep it if required by reviewers.
+  select(brier.corrected, p_mean, brierscaled)
+}
+
+brier_scores <-
+  model_data %>%
+  mutate(brier = map(all_models, brier)) %>%
+  select(outcome, time, brier) %>%
+  unnest("brier")
+
+cache("brier_scores")

### 2020-04-29 src/fig_auc.R

@@ -1,0 +1,49 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+digs <- options("digits")
+options(digits = 2)
+
+
+# Prepare data ------------------------------------------------------------
+
+data_auc_ci <-
+  model_data %>%
+  select(outcome, time, all_models) %>%
+  unnest("all_models") %>%
+  #filter(fig) %>%
+  select(outcome, time, Model, starts_with("AUC")) %>%
+  pivot_longer(c(-outcome, -time, -Model)) %>%
+  filter(!is.na(value)) %>%
+  separate(name, c("AUC","level", "corr")) %>%
+  pivot_wider(names_from = level) %>%
+  mutate(
+    Model = fct_reorder(Model, est),
+    time = factor(time, c("90d", "2y"), c("90 days", "2 years"))
+  ) %>%
+  filter(!is.na(corr))
+
+annotates <-
+  data_auc_ci %>%
+  pivot_longer(c(lo, est, hi))
+
+# Figure ------------------------------------------------------------------
+
+ggplot(data_auc_ci, aes(Model, est, group = corr, color = corr)) +
+  #geom_hline(yintercept = .7, color = "darkgreen", linetype = "dashed", size = 1) +
+  geom_pointrange(aes(ymin = lo, ymax = hi), size = 1) + #, position = position_dodge(.5)) +
+  geom_text(aes(Model, value, label = round(value, 2)), data = annotates, nudge_x = .3, size = 2) +
+  coord_flip() +
+  #theme_light() +
+  #theme_minimal() +
+  theme(
+    axis.title.y = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "none"
+  ) +
+  ylab("AUC with 95% CI") +
+  scale_y_continuous(breaks = seq(0, 1, .05)) +
+  facet_grid(outcome ~ time)
+
+options(digits = digs$digits)
+
+ggsave("graphs/auc_ci.png", width = 30, height = 20, units = "cm")

### 2020-04-29 src/fig_calibration.R

@@ -1,0 +1,46 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# SHAR (internal calibration) ---------------------------------------------
+
+get_beltdata <- function(cb) {
+  tibble(
+    x = cb$seqP,
+    L = cb$cbBoundByConfLevel[[1]]$L,
+    U = cb$cbBoundByConfLevel[[1]]$U
+  )
+}
+
+shar_calibration <-
+  model_data %>%
+  select(outcome, time, all_models) %>%
+  unnest("all_models") %>%
+  filter(
+    "BRL (age as main effect)"
+  ) %>%
+  mutate(cal_belt = map(cal_belt, get_beltdata)) %>%
+  select(outcome,time,  Model, cal_belt) %>%
+  unnest(cal_belt) %>%
+  mutate(time = factor(time, c("90d", "2y"), c("90 days", "2 years")))
+
+# Make figure -------------------------------------------------------------
+calplot <- function(xlim = c(0, 0.17), ylim = c(0, 0.25)) {
+  ggplot(shar_calibration, aes(x, ymin = L, ymax = U)) +
+  geom_ribbon(alpha = .3) +
+  geom_abline(aes(intercept = 0, slope = 1)) +
+  # theme_minimal() +
+  scale_x_continuous(labels = scales::percent_format(1)) +
+  scale_y_continuous(labels = scales::percent_format(1)) +
+  coord_cartesian(xlim, ylim) +
+  xlab("Predicted probability") +
+  ylab("Observed proportion") +
+  theme(
+    legend.position = c(1, 0),
+    legend.justification = c(1, 0),
+    legend.title = element_blank()
+  ) +
+  facet_grid(outcome ~ time)
+}
+
+
+ggsave("graphs/calibration.png", calplot(), height = 10, width = 20 , units = "cm")
+ggsave("graphs/calibration_zoom.png", calplot(c(0, 0.05), c(0, 0.05)), height = 10, width = 20 , units = "cm")

### 2020-04-29 src/fig_roc.R

@@ -1,0 +1,36 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+# SHAR --------------------------------------------------------------------
+
+shar_roc <-
+  model_data %>%
+  select(outcome, time, all_models) %>%
+  unnest("all_models") %>%
+  #filter(fig) %>%
+  unnest(roc_curve) %>%
+  select(outcome, time, Model, specificity, sensitivity) %>%
+  mutate(
+    Model = model_names(Model),
+    time = factor(time, c("90d", "2y"), c("90 days", "2 years"))
+  )
+
+# Figure ------------------------------------------------------------------
+
+fig_roc <-
+  shar_roc %>%
+  ggplot(aes(1 - specificity, sensitivity, col = Model)) +
+  geom_path(size = 1) +
+  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
+  # theme_minimal() +
+  theme(
+    legend.position = "bottom", # c(1, 0),
+    #legend.justification = c(1, 0),
+    legend.title = element_blank()
+  ) +
+  facet_grid(outcome ~ time) +
+  scale_x_continuous(labels = scales::percent) +
+  scale_y_continuous(labels = scales::percent)
+
+
+ggsave("graphs/roc.png", fig_roc, height = 30, width = 30, units = "cm")

### 2020-04-29 src/fig_separation.R

@@ -1,0 +1,87 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+obspred <-
+  model_data %>%
+  select(outcome, time, all_models) %>%
+  unnest("all_models") %>%
+  select(outcome, time, Model, fig, obspred, AUC_lo) %>%
+  unnest(obspred) %>%
+  filter(
+    Model == "BRL (age as main effect)"
+  ) %>%
+  mutate(time = factor(time, c("90d", "2y"), c("90 days", "2 years")))
+
+
+# Histogram ---------------------------------------------------------------
+
+fig_separation_hist <-
+  obspred %>%
+  ggplot(aes(pred)) +
+  geom_histogram(
+    aes(fill = obs),
+    alpha = .5, position = "identity"
+  ) +
+  theme_minimal() +
+  theme(
+    legend.position = "none",
+    panel.grid = element_blank(),
+    axis.text.x = element_text(angle = 30),
+    axis.ticks = element_line(size = 1)
+  ) +
+  xlab("") +
+  ylab(expression(paste(sqrt(n)))) +
+  scale_x_log10(
+    #breaks = c(.01, .1, 1, 3, 6, 10) / 100,
+    limits = c(.001, .1),
+    labels = scales::percent
+  ) +
+  scale_y_sqrt(
+    #breaks = c(0, 100, 500, 1000, 2000, 3000, 4000, 5000), #seq(0, 6000, 1000),
+    labels = function(x) format(x, big.mark = ",")
+  ) +
+  expand_limits(x = 0) +
+  facet_grid(outcome ~ time, scales = "free_y")
+
+
+# Density -----------------------------------------------------------------
+
+fig_separation_density <-
+  obspred %>%
+  ggplot(aes(pred)) +
+  geom_histogram(
+    aes(
+      y = ..density..,
+      fill = obs
+    ),
+    alpha = .5, position = "identity"
+  ) +
+  geom_density(aes(col = obs), bw = .1) +
+  theme_minimal() +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    panel.grid = element_blank(),
+    axis.text.x = element_text(angle = 30),
+    axis.ticks = element_line(size = 1),
+    axis.ticks.y = element_line(color = "white"),
+    axis.text.y = element_text(color = "white", size = 17)     # To align with upper panel
+  ) +
[…]

### 2020-04-29 src/overall_surv.R

@@ -1,0 +1,28 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+# This is just to calculate over-all survival as background info
+
+fit <- survfit(Surv(stime, status) ~ 1, df)
+
+surv90d <-
+  with(
+    summary(fit, 90),
+    sprintf("%.2f %% (95 %% CI: %.2f - %.2f)",
+    (1 - surv) * 100, (1 - upper) * 100, (1 - lower) * 100)
+  )
+cache("surv90d")
+
+
+p <-
+  survminer::ggsurvplot(
+  fit,
+  risk.table = TRUE,
+  censor = FALSE,
+  ylim = c(.7, 1),
+  surv.scale = "percent",
+  xscale = "d_y",
+  break.time.by = 365.241
+)
+
+ggpubr::ggexport(p, filename = "graphs/km.png")

### 2020-04-29 src/tab_coefs.R

@@ -1,0 +1,85 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# Table to present --------------------------------------------------------
+
+tbl_coefs <-
+  model_data %>%
+  select(outcome, time, all_models) %>%
+  unnest("all_models") %>%
+  filter(
+    Model == "BRL (age as main effect)"
+  ) %>%
+  select(outcome, time, Model, tidy) %>%
+  unnest(tidy) %>%
+  transmute(
+    outcome,
+    time = factor(time, c("90d", "2y"), c("90 days", "2 years")),
+    Model,
+    term = clean_names(term),
+    term = gsub("_", " ", term),
+    term = gsub("Diagnos", "Diagnos: ", term),
+    math = sprintf("\\text{%s} \\\\ ", term) %>% {gsub("\\text{(intercept)} \\\\ ", "", ., fixed = TRUE)},
+    beta = log(estimate),
+    `OR 95 % CI` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
+    p    = pvalue_format(.01)(p.value)
+  ) %>%
+  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(term == "(intercept)", "", .))
+
+cache("tbl_coefs")
+
+
+# List of coefficients for text -------------------------------------------
+
+
+set_first <- function(coefs, x) {
+  if (x %in% coefs) c(x, setdiff(coefs, x))
+  else coefs
+}
+
+coefs_print_string <- function(name) {
+  all_vars <-
+    filter(all_models, Model == !!name) %>%
+    {.$tidy[[1]]$term[-1]}
+
+  c_vars     <- all_vars[startsWith(all_vars, "c_")]
+
+  basic_vars <-
+    setdiff(all_vars, c_vars) %>%
+    clean_names(firstupper = FALSE, lvls = FALSE) %>%
+    set_first("ASA class") %>%
+    set_first("sex") %>%
+    set_first("age")
+
+  c_vars <- clean_names(c_vars, FALSE)
+
+  paste(
+    glue::glue_collapse(basic_vars, ", "),
+    "and the precense of",
+    glue::glue_collapse(c_vars, ", ", last = " and ")
+  )
+}
+
+# coefs_print_reduced <- coefs_print_string("BRL reduced (age as main effect)")
+# coefs_print         <- coefs_print_string("BRL (age as main effect)")
[…]

### 2020-04-29 src/tab_define_groups.R

@@ -1,0 +1,15 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+tab_categorization <-
+  categorization.Blad1 %>%
+  mutate_all(zoo::na.locf) %>%
+  group_by(new, from) %>%
+  summarise(old = paste(str_to_sentence(old), collapse = ", ")) %>%
+  pivot_wider(names_from = from, values_from = old) %>%
+  rename(
+    `Comorbidities by groups` = new,
+    Charlson = CCI,
+    Elixhauser = ECI
+  )
+
+cache("tab_categorization")

### 2020-04-29 src/tab_table1.R

@@ -1,0 +1,88 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+# Prepare data ------------------------------------------------------------
+tab1 <- function(df) {
+  dft1 <-
+    df %>%
+    mutate(
+      Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original == 4, "4+"),
+      Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all == 3, "3+"),
+      RxRiskV    = Rx_index_pratt
+    ) %>%
+    select(
+      outcome,
+      P_Age,
+      P_Sex,
+      P_BMI,
+      P_ASA,
+      P_DiaGrp,
+      cemented_stem,
+      cemented_cup,
+      P_TypeOfHospital,
+      education,
+      civil_status,
+      Charlson,
+      Elixhauser,
+      RxRiskV,
+      starts_with("c_"),
+    ) %>%
+    setNames(clean_names(names(.))) %>%
+    rename(`ASA class` = ASA)
+
+  fct <- c("Sex", "ASA class", "Elixhauser",
+           "Charlson", "Diagnos", "Cemented stem", "Cemented cup")
+
+    t1 <-
+    tableone::CreateTableOne(
+      strata = "Outcome",
+      vars = setdiff(names(dft1), "Outcome"),
+      factorVars = fct,
+      data = dft1,
+      test = FALSE
+    )
+
+  t1_all <-
+    tableone::CreateTableOne(
+      vars = setdiff(names(dft1), "Outcome"),
+      factorVars = fct,
+      data = dft1,
+      test = FALSE
+    ) %>%
+    print(
+      printToggle = FALSE
+    ) %>%
+    as_tibble()
+
+  table1 <-
+    t1 %>%
+    print(
+      printToggle = FALSE
+    ) %>%
+    as_tibble(rownames = "what") %>%
+    add_column(Total = t1_all$Overall) %>%
+    mutate(
+      level = trimws(ifelse(startsWith(what, " "), what, "")),
+      level = paste0(toupper(substr(level, 1, 1)), substring(level, 2)),
+      what = trimws(ifelse(level == "", gsub(" = TRUE", "", what), ""))
+    ) %>%
+    mutate_at(vars(`TRUE`, `FALSE`), zero) %>%
+    select(what, level, `TRUE`, `FALSE`, Total)
+
+  list(t1 = t1_all, table1 = table1)
+}
+
[…]

### 2020-05-09 munge/05-BRLasso.R
Anpassat till endast infektion (ej luxation).

@@ -3,7 +3,7 @@
 
 infection_data <-
   model_data %>%
-  filter(outcome == "infection")
+  dplyr::filter(outcome == "infection")
 
 
 

### 2020-05-09 reports/article.Rmd
Inkoorporera NHs textförslag. Fix #10.

@@
 ---
-title: "Title"
-author: X^1,2,3^, Y^2,4^, ... Z^5^
+title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and ..."
+author: Erik BülowX^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
 date: '`r Sys.Date()`'
@@
 
-## Affiliations
+1. The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
+2. Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
+3. Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
 
-1. Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
-2. The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
-3. Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-4. ...
-5. ...
+**Correspondance:** erik.bulow@vgregion.se. +46 70 08 234 28.
+Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
 
@@
 
-**SUBTITLE 1** Paragraph 1 ...
+**OJECTIVE: ** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and healthcare resources.
+Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population.
+We thus developed a parsimonious model to predict 90-day mortality after elective THA.
 
-**SUBTITLE 2** Paragraph 2 ...
+**PARTICIPANTS AND METHODS: ** We analysed XXX patients registered in the Swedish hip arthroplasty register 2008 - 2015 and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery.
+Models were internally validated by bootstrapping and externally validated on a cohort of XXX patients from XXX.
+We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration plots.
 
-**SUBTITLE 3** Paragraph 3 ...
+**RESULTS: ** The unadjusted cumulative incidence of PJI was XXX within 90 days and XXX within two years in the Swedish derivation cohort.
+A prediction model for PJI within 90 days combining the underlying diagnosis for THA, the presence of psoriatic or CNS disease, cancer, substance abuse, body mass index, and gender had good predictive performance internally (AUC=XXX, 95% CI: XXX to XXX) and externally (AUC=XXX, 95% CI: XXX to XXX), and a similar model to predict PJI within two years had an AUC of XXX (95% CI: XXX to XXX) internally and XXX (95% CI: XXX to XXX) externally.
+These models were superior to traditional models based on American Society for Anesthesiologists (ASA) classification (AUC=XXX, 95% CI: XXX to XXX), the Charlson (AUC=XXX, 95% CI: XXX to XXX) or Elixhauser (AUC=XXX, 95% CI: XXX to XXX) co-morbidity indices.
+
+**CONCLUSION: ** The proposed prediction models for PJI after THA are relatively parsimonious since they are based on easily accessible clinical data, and they are superior to predictions based on ASA class or more complex comorbidity indices.
+Our web-based calculator could simplify the individualized risk assessment prior to THA surgery.
 
@@
 
+Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.1-3 Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.4-12
+Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context.
+Although high ASA class is a risk factor associated with development of PJI the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.13, 14 The Charlson and Elixhauser co-morbidity indices are based on 17 or 31 International Classification of Diseases (ICD)-codes, respectively, but they cannot be used in a clinical setting, 15, 16{walraven} and they were also not designed for prediction of specific adverse events after THA.
+Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.17, 18 Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.18-20 This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
+Thus, there is an urgent need for improved prediction models of PJI.
+We therefore developed a prediction model of PJI within 90 days or two years after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.
+
 # Patients and Methods
 
-We used data from two national quality registers 2008-2015: The Swedish hip arthroplasty register (SHAR), with a completeness of 96-98%,[@Karrholm2018] for model development and internal validation, and XXX for external validation.
+We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
+The best models were then validated externally on a cohort from the XXX.
 
-SHAR was linked using Swedish personal identity numbers[@Ludvigsson2009] to enhance the variable set with education and civil status from Statistics Sweden,[@Ludvigsson2019] comorbidity and adverse events from the national patient register,[@Ludvigsson2011] and prescribed medications from the medical prescription register.[@Cnudde2016]
-We excluded patients with hip arthroplasty due to fractures, tumours, unspecified, or unknown reasons.
-We included patients with either unilateral THA, or with their second staged bilateral hip.[@Bulow2020] Patients younger than 18 or older than 100 years were excluded, as were patients with a body mass index (BMI) above 50, or with missing data on BMI, ASA class, education, type of hospital, and cementation (Figure \@ref(fig:flowchart)).
+## Derivation cohort (Sweden)
 
-Comorbidity, during one year before surgery, was based on the Swedish version of the 10th revsin of the international classification of diseases (ICD-10-SE), as recorded at any hospital physician appointment prior to surgery.
-Individual codes were then combined using Charlson and Elixhauser comorbidity groups.[@Quan2005] A similar procedure was performed for codes of the anatomical therapeutic chemical (ATC) classification, grouped by Rx Risk V.[@Pratt2018] Similar conditions were then combined to a broader classification scheme based on clinical relevance, and those categories were used as possible predictors (Table \@ref(tab:tabcategorization)).
+The derivation cohort was obtained from patients registered in the SHAR from 2008 to 2015.
+The starting point was chosen because registration of ASA class and body mass index (BMI) in the SHAR was initiated in 2008, and we had no access to data from the national patient register (NPR) after 2015.
+Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.
+The diagnosis underlying the index THA procedure (categorized into primary osteoarthritis, secondary osteoarthritis, pediatric hip disorder, avascular necrosis of the femoral head [AVN], and inflammatory joint disease), type of THA (cemented, uncemented hybrid), age at index surgery, gender, BMI, ASA class, type of hospital, and year of surgery were obtained from the SHAR.
+Data on co-morbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
+Data on marital status and the educational level  of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
+The SHAR has completeness of 96-98 % and 100% coverage,21 the LISA has coverage of 98% with 85% accuracy,{ludvigsson 2019 LISA} and completeness for the NPR is above 99% and the positive predictive value is 85-95%.22 Death dates were linked from the national population register.
+Patients with missing information on or an ASA class of IV and above were excluded, as were patients with missing information on BMI or a BMI above 50.
+Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.
 
-Infections within 90 days and 2 years respectively, were identified by relevant ICD-10-SE codes, or by codes from the Classification of Surgical Procedures (NCSP) from the Nordic Medico-Statistical Committee (NOMESCO) recorded in the national patient register.
-An infection was also identified if causing a reoperation recorded to SHAR.
+## External validation cohort (XXX)
 
-## Statistical analysis for model development
+The Charlson and Elixhauser co-morbidity indices are based on the presence of ICD-10 codes grouped into main categories, 17 for the Charlson 15{deyo}16 and 31 for the Elixhauser index.{elixhauser} Diagnostic categories from both indices were combined to define 21 categories of co-morbidity (Tab.
+1) that were chosen in order to be easily identifiable in a clinical setting.
+Some co-morbidities defined by either the Charlson or the Elixhauser co-morbidity index were not present in any patient affected by PJI in the derivation cohort, and these diagnoses were therefore excluded as potential predictors.
+PJI was defined by the occurrence of any ICD- or NOMESCO-code indicative of this adverse event, and the time point at which any of these codes was registered allowed for the distinction between PJI occurring within 90 days or within two years .
 
-Infections within 90 days or 2 years were modelled with logistic regression, ignoring possi
+## Model development
 
-We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA.
-Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales.
-We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.
-Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables.
-Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process.
-1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model.
-We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample.
-Those $\lambda$:s were used to estimate model coefficients for each potential predictor.
-The magnitude (absolute values) of those estimates were used as a measure of variable importance.
-Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed.
-Potential predictors with variable importance above this breaking point were considered important and kept as model candidates.
-The whole process was repeated 100 times.
-Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model).
-A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison.
-This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies.
-Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex.
-Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.
-Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]
+The cumulative incidence of PJI was estimated using the Kaplan-Meier method.
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).{guo 2015; Baranowski 2020} Age and BMI were normalized prior to modelling, and the estimated coefficients thus indicate variable importance on the same scale as categorical variables.
+1,000 bootstrap samples were drawn from the observed data set.{Austin 2004} Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values (??:s) in a logistic LASSO model, keeping only ??:s that minimized the mean cross-validated deviances in each sample and using these ??:s to estimate model coefficients for each potential predictor.
+Potential predictors were identified using piece-wise linear regression and kept as model candidates, regressions were iterated 100 times, and potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
+Even more parsimonious models with potential predictors chosen at least XXX out of 100 times were evaluated for comparison.
+Univariable prediction models based on either ASA class, the Charlson or the Elixhauser co-morbidity index were fitted for comparison.
+Age and BMI were either treated as continuous variables in main effect models, or modelled using restricted cubic splines with two or three knots.
+In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI this variable was also treated as a categorical variable divided into the six BMI intervals suggested by the World Health Organization (WHO; <18.5: underweight, 18.5-24.9: normal, 25-29.9: overweight, 30-34.9: class I obesity, 35.0-39.9: class II obesity, and ???40: class III obesity).
+Odds ratios for the final models were estimated with 95% confidence intervals.
 
-## Statistical analysis for model validation
+## Model validation
 
-Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation).
-Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good.
-Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples.
-We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally.
-An AUC with 95 % confidence interval was calculated for the model as-is.
-Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales.
-An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.
+Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation).
+Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95% confidence limits that were based on percentiles from 2,000 non-parametric bootstrap samples.
+The bias-corrected Somers' D_xy rank correlation based on 100 resamples was used to adjust for optimism.
+Predicted probabilities and observed proportions were plotted with 95% confidence intervals to graphically assess model calibration.{nattino 2016} The derived models were externally validated on a XXX cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to XXX.
+Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.
 
@@
 
-We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC.
+We built an online web calculator available at https XXX to be used in clinical practice.
+R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny was used.
 
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13). Informed consent was not mandatory according to the Swedish patient data law from 2009.
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and XXX.
 
@@
 
-`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel).
-`r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that.
-The unadjusted risk of 90-day mortality was therefore `r surv90d`.
-Further characteristics of the study population are presented in Table \@ref(tab:tab1).
-`r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser.
-The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.
-
-In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel).
-Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.
+We included XXX patients (XXX% females) with mean age of XXX (SD XXX) in the derivation cohort from the SHAR (Fig.
+1, left panel).
+The cumulative incidence of PJI within 90 days was XXX% and XXX% within two years (Table 2).
+The proportion of patients with ASA class III was XXX%, XXX% had at least one comorbidity according to the Charlson co-morbidity index, and XXX% had at least one comorbidity defined by the Elixhauser index.
+In addition, XXX patients from XXX were included as an external validation cohort (Fig.
+1, right panel).
 
@@
 
-```{r}
-spell <- function(x)
-  c("one", "two", "three", "four", "five", "six", 
-    "seven", "eight", "nine", "ten", "eleven", "twelve")[x]
-```
-
-There were `r spell(nrow(comb_lgl))` co-morbidities that were not recorded for any patient who died in the Swedish derivation cohort: `r comb_lgl_text`.
-Those variables were therefore excluded as potential predictors prior to any statistical model derivation.
-The derived main model included `r coefs_print`.
-The reduced model, with covariates included at least 33 out of 100 times, was restricted to `r coefs_print_reduced` (Table \@ref(tab:brlprop)).
-
-There were no differences between models including age as a main effect, compared to those were age was modeled by restricted cubic splines with either two or three knots.
-We therefore focused on the more parsimonious models with age as a main effect.
-Similarly, the correction for optimism only affected the third decimals of the AUC confidence intervals and was therefore omitted.
-
-```{r}
-half_paran <- function(x) 
-  gsub(
-    "(\\(AUC = )(0.\\d*)(, )(95 % CI: 0.\\d{2} to 0.\\d{2}\\))", 
-    "\\2 \\(\\4", 
-    get_auc(x), 
-    perl = TRUE
-  )
-```
-
-The main and reduced models were no different regarding discriminative power `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`.
-We therefore considered the reduced model as superior due to its simplicity.
-Traditional models performed poorly with 95 % confidence intervals not above 0.7: The Charlson co-morbitiy index had an AUC of `r half_paran("Charlson")` and the Elixhauser co-morbidity an AUC of `r gsub("\\)", ";", half_paran("Elixhauser"))` Figure \@ref(fig:rocs) left panel and Figure \@ref(fig:aucci)).
-
-The ability of the reduced model to estimate probabilities of death within 90 days is further illustrated in Figure \@ref(fig:sep).
-Model calibration was good for estimated probabilities up to 3 % and acceptable up to 5 %, although with predicted probabilities usually higher than observed proportions (Figure \@ref(fig:calibration)).
-
-Estimated model coefficients and corresponding odds ratios for the reduced model are presented in Table \@ref(tab:brlreduced).
-
-Omitting cancer from the reduced model did not affect the AUC or calibration for estimated probabilities below 3 %, but calibration outside this range deteriorated, and we thus retained cancer as an important predictor.
+The main model for the prediction of PJI within 90 days in the Swedish cohort included seven predictor variables, one of which was the underlying diagnosis for THA surgery.
+Other predictors remaining in the main model were the presence of psoriatic and CNS diseases, cancer, substance abuse, BMI, and gender (Tab.
+XXX a).
+Predictors associated with a statistically significant risk increase were AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender.
+The main model for the precition of PJI within 90 days had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) had less discriminative power (Fig.
+XXX, XXX).
+Calibration was good for the developed main model (Fig.
+XXX).
+The main model for PJI within two years included the same predictors, with the addition of the presence of rheumatic disease (Tab.
+XXX b).
+Predictors conferring a statistically significant risk increase were those mentioned as predictive for PJI within 90 days, and additionally included the presence of psoriatic, rheumatic, or CNS disease, and cancer.
+The main model for the prediction of PJI within two years had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) again had less discriminative power (Fig.
+XXX, XXX).
+Calibration was good for the developed main model (Fig.
+XXX).
 
@@
 
+XXX
+
 # Discussion
+
+## Principal findings
+
+We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson or Elixhauser co-morbidity indices.
+A similarly developed model to predict PJI within two years included the same predictors as the model developed to predict PJI within 90 days, but added the presence of rheumatic disease to the list of predictor variables.
+This model also performed better than the established co-morbidity measures in terms of AUC and calibration.
+
+## Clinical usage
+
+Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into in a clinical setting, either by use of the formula given in Tab.
+XXX, or by accessing a simple web calculator online.
+For example, ...
+
+## Model predictors
+
+The predictors in our models were chosen based on a strict analytical approach, not on proposed relevance of patient or procedure characteristics.
+The majority of the identified predictor variables are however associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.
+
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred an about two-fold increased risk of PJI in our prediction models.23, 24 Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,9, 20 but this predictor variable only reached borderline statistical significance in our model of 90-day PJI.
+However, the presence of psoriatic disease, and, for the outcome PJI within two years, the presence of rheumatic disease, were both statistically significant predictors of outcome in our models.
+This seeming discrepancy may be explained by the fact that other physicians may have identified patients as suffering from any of these conditions, thereby assigning the indicative ICD codes, whereas the operating surgeon may have classified the underlying reason for THA as something other than inflammatory joint disease.
+Obesity is associated with an increase in the risk of reoperations or adverse events after THA25 and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.6, 10, 12, 26 Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.27 In accordance with observational studies9 we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
+Cancer seems to confer an increased risk of PJI in several observational studies,9, 26, 28 and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.18 Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,24 and gender is included as a predictor variable in several previously developed risk calculators.18, 27 Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+High ASA class is frequently associated with an increased risk of PJI,14, 29, 30 as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.7, 31 In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
+Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,11 our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
+THA fixation using cement without antibiotics confers an increased risk of PJI,24 but such cement brands were not in clinical use in Sweden during the studied period.
+We found no indication that the type of fixation was associated with the risk of PJI, and this variable was also not selected as a potential predictor variable in any of the iterations during model development.
+
+## Strengths and limitations
+
+Our study is hampered by the limitations inherent to observational research, such as misclassification, missing information, and .
+The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
+This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,8, 10, 26 and diabetes is also found among the risk factors in previously developed risk calculators.18, 27, 32 On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.30 Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.33, 34 In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,11 but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings.
+Use of TNF-a blockers may be associated with the risk of developing PJI,35, 36 but we had no information on medication.
+Similarly, intra-articular steroid application may enhance the risk of developing PJI,38 but such manipulations would not be identifiable within a Swedish registry setting.
+Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,12, 37 and procedure time is one of the risk factors in the Mayo PJI risk score,14 but we had no information on this potentially important parameter.
+Taken together, the inclusion of some or all of the variables mentioned above-unfortunately unavailable to us-might improve the predictive power of PJI risk assessment models.
+The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
+The question of whether there is a genetic predisposition towards developing orthopaedic infections has only been ??
 
@@
 
-##### PAGE BREAK
+Our results indicate that the risk of PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
+We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
+
+#### Page break ####
 
@@
 
+NPH initiated the study, OR managed the ethical review board application in Sweden, EB developed the statistical model, XXX performed external validation with data from XXX.
+NPH and EB drafted, and all authors edited and finalized the manuscript.
+
+# Competing interest statement
+
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare ...
+
+# Transparency Statement
+
+NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
+
+# Public and patient involvement statement
+
+There is a patient representative at SHAR to ensure the patients' interests are respected in planned research.
+
+# Data sharing statement
+
+Data is available upon reasonable request to the principal investigator.
+
 # Acknowledgement
+
+We would like to thank .
 

### 2020-05-09 reports/article.Rmd
Included Zotero references. Fix #15.

@@
 title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and ..."
-author: Erik BülowX^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
+author: Erik Bülow^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
 date: '`r Sys.Date()`'
-bibliography: H:/zotero_lib.bib
+# bibliography: H:/zotero_lib.bib
+bibliography: ../../../library.bib
 csl: the-lancet.csl
@@
   knitr.kable.NA = '',
-  repos          = list(CRAN = "https://cran.rstudio.com/")
+  repos          = list(CRAN = "https://cran.rstudio.com/"),
+  citr.bibliography_path = "C:/Users/erik_/Documents/library.bib"
 )
@@
 **Correspondance:** erik.bulow@vgregion.se. +46 70 08 234 28.
-Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden
+Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.1-3 Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.4-12
+Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
 Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context.
-Although high ASA class is a risk factor associated with development of PJI the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.13, 14 The Charlson and Elixhauser co-morbidity indices are based on 17 or 31 International Classification of Diseases (ICD)-codes, respectively, but they cannot be used in a clinical setting, 15, 16{walraven} and they were also not designed for prediction of specific adverse events after THA.
-Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.17, 18 Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.18-20 This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
+Although high ASA class is a risk factor associated with development of PJI the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on 17 or 31 International Classification of Diseases (ICD)-codes, respectively, but they are cumbersome be use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
 Thus, there is an urgent need for improved prediction models of PJI.
@@
 Data on marital status and the educational level  of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
-The SHAR has completeness of 96-98 % and 100% coverage,21 the LISA has coverage of 98% with 85% accuracy,{ludvigsson 2019 LISA} and completeness for the NPR is above 99% and the positive predictive value is 85-95%.22 Death dates were linked from the national population register.
+The SHAR has completeness of 96-98 % and 100% coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99% and the positive predictive value is 85-95%.[@Ludvigsson2011] Death dates were linked from the national population register.
 Patients with missing information on or an ASA class of IV and above were excluded, as were patients with missing information on BMI or a BMI above 50.
@@
 
-The Charlson and Elixhauser co-morbidity indices are based on the presence of ICD-10 codes grouped into main categories, 17 for the Charlson 15{deyo}16 and 31 for the Elixhauser index.{elixhauser} Diagnostic categories from both indices were combined to define 21 categories of co-morbidity (Tab.
+The Charlson and Elixhauser co-morbidity indices are based on the presence of ICD-10 codes grouped into main categories, 17 for the Charlson 15 and 31 for the Elixhauser index.[@Quan2005] Diagnostic categories from both indices were combined to define 21 categories of co-morbidity (Tab.
 1) that were chosen in order to be easily identifiable in a clinical setting.
@@
 The cumulative incidence of PJI was estimated using the Kaplan-Meier method.
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).{guo 2015; Baranowski 2020} Age and BMI were normalized prior to modelling, and the estimated coefficients thus indicate variable importance on the same scale as categorical variables.
-1,000 bootstrap samples were drawn from the observed data set.{Austin 2004} Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values (??:s) in a logistic LASSO model, keeping only ??:s that minimized the mean cross-validated deviances in each sample and using these ??:s to estimate model coefficients for each potential predictor.
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age and BMI were normalized prior to modeling, and the estimated coefficients thus indicate variable importance on the same scale as categorical variables.
+1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor.
 Potential predictors were identified using piece-wise linear regression and kept as model candidates, regressions were iterated 100 times, and potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
@@
 Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95% confidence limits that were based on percentiles from 2,000 non-parametric bootstrap samples.
-The bias-corrected Somers' D_xy rank correlation based on 100 resamples was used to adjust for optimism.
-Predicted probabilities and observed proportions were plotted with 95% confidence intervals to graphically assess model calibration.{nattino 2016} The derived models were externally validated on a XXX cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to XXX.
+The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism.
+Predicted probabilities and observed proportions were plotted with 95% confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a XXX cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to XXX.
 Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.
@@
 Predictors associated with a statistically significant risk increase were AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender.
-The main model for the precition of PJI within 90 days had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) had less discriminative power (Fig.
+The main model for the predicition of PJI within 90 days had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) had less discriminative power (Fig.
 XXX, XXX).
@@
 
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred an about two-fold increased risk of PJI in our prediction models.23, 24 Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,9, 20 but this predictor variable only reached borderline statistical significance in our model of 90-day PJI.
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred an about two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] but this predictor variable only reached borderline statistical significance in our model of 90-day PJI.
 However, the presence of psoriatic disease, and, for the outcome PJI within two years, the presence of rheumatic disease, were both statistically significant predictors of outcome in our models.
 This seeming discrepancy may be explained by the fact that other physicians may have identified patients as suffering from any of these conditions, thereby assigning the indicative ICD codes, whereas the operating surgeon may have classified the underlying reason for THA as something other than inflammatory joint disease.
-Obesity is associated with an increase in the risk of reoperations or adverse events after THA25 and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.6, 10, 12, 26 Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.27 In accordance with observational studies9 we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
-Cancer seems to confer an increased risk of PJI in several observational studies,9, 26, 28 and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.18 Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,24 and gender is included as a predictor variable in several previously developed risk calculators.18, 27 Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
-High ASA class is frequently associated with an increased risk of PJI,14, 29, 30 as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.7, 31 In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
+Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.[@Tan2018] In accordance with observational studies9 we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.[@Wingert2016a] Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.7, 31 In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
 Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,11 our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
@@
 The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
-This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,8, 10, 26 and diabetes is also found among the risk factors in previously developed risk calculators.18, 27, 32 On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.30 Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.33, 34 In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,11 but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings.
-Use of TNF-a blockers may be associated with the risk of developing PJI,35, 36 but we had no information on medication.
-Similarly, intra-articular steroid application may enhance the risk of developing PJI,38 but such manipulations would not be identifiable within a Swedish registry setting.
-Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,12, 37 and procedure time is one of the risk factors in the Mayo PJI risk score,14 but we had no information on this potentially important parameter.
+This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings.
+Use of TNF-a blockers may be associated with the risk of developing PJI,[@Gilson2010; @Momohara2011] but we had no information on medication.
+Similarly, intra-articular steroid application may enhance the risk of developing PJI,[@Cancienne2015] but such manipulations would not be identifiable within a Swedish registry setting.
+Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter.
 Taken together, the inclusion of some or all of the variables mentioned above-unfortunately unavailable to us-might improve the predictive power of PJI risk assessment models.
 The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
-The question of whether there is a genetic predisposition towards developing orthopaedic infections has only been ??
+The question of whether there is a genetic predisposition towards developing orthopaedic infections has only been ...
 
@@
 
-#### Page break ####
+##### Page break
 
@@
 
-```{r, child = "figs.Rmd"}
+```{r, child = "figs.Rmd", eval=FALSE}
 ```
 
-```{r, child = "tabs.Rmd"}
+```{r, child = "tabs.Rmd", eval=FALSE}
 ```

### 2020-05-09 reports/article.Rmd
m

@@
 Age and BMI were either treated as continuous variables in main effect models, or modelled using restricted cubic splines with two or three knots.
-In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI this variable was also treated as a categorical variable divided into the six BMI intervals suggested by the World Health Organization (WHO; <18.5: underweight, 18.5-24.9: normal, 25-29.9: overweight, 30-34.9: class I obesity, 35.0-39.9: class II obesity, and ???40: class III obesity).
+In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI this variable was also treated as a categorical variable divided into the six BMI intervals suggested by the World Health Organization (WHO; <18.5: underweight, 18.5-24.9: normal, 25-29.9: overweight, 30-34.9: class I obesity, 35.0-39.9: class II obesity, and $\geq 40$: class III obesity).
 Odds ratios for the final models were estimated with 95% confidence intervals.

### 2020-05-11 reports/article.Rmd
Uppdaterad version skickad till NH.

--- 2020-05-09 reports/article.Rmd
@@
 title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and ..."
-author: Erik Bülow^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
+author: Erik Bulow^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
 date: '`r Sys.Date()`'
-# bibliography: H:/zotero_lib.bib
-bibliography: ../../../library.bib
+bibliography: H:/zotero_lib.bib
 csl: the-lancet.csl
@@
   knitr.kable.NA = '',
-  repos          = list(CRAN = "https://cran.rstudio.com/"),
-  citr.bibliography_path = "C:/Users/erik_/Documents/library.bib"
+  repos          = list(CRAN = "https://cran.rstudio.com/")
 )
@@
 library(tidyverse)
+
+nmbr <- function(x) format(as.numeric(trimws(as.character(x))), big.mark = ",")
+
+# Pre-load all cached files < 10 MB
+cache_files <- dir("../cache", ".RData", full.names = TRUE)
+filesizes <- do.call(file.size, list(cache_files)) / 2^20
+Vectorize(load, "file")(cache_files[filesizes < 10], .GlobalEnv)
 ```
@@
 
-**Correspondance:** erik.bulow@vgregion.se. +46 70 08 234 28.
-Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
+**Correspondance:** erik.bulow@vgregion.se; +46 70 08 234 28; Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
 
 ##### PAGE BREAK
+
+```{r}
+# Participants and methods --------------------------------------------
+N_shar <- 
+  table1 %>% 
+  filter(what == "n") %>% 
+  select(Total) %>% 
+  pluck(1)
+
+period <- paste(range(df$P_SurgYear), collapse = "-")
+
+# Results -------------------------------------------------------------
+crude_rate_90d <- sum(df$outcome_infection_90d) / nrow(df)
+crude_rate_2y  <- sum(df$outcome_infection_2y) / nrow(df)
+
+# Extract predictors for specified period
+# Needs additional formatting
+predictors <- function(period) {
+  tbl_coefs %>% 
+  filter(
+    time == period,
+    term != "(Intercept)"
+  ) %>% 
+  transmute(
+    term = tolower(term),
+    term = replace(
+      term, 
+      startsWith(term, "diagnosis"), 
+      "the underlying diagnosis for THA"
+    ),
+    term = replace(term, startsWith(term, "bmi"), "body mass index"),
+    term = gsub("male ", "", term)
+  ) %>% 
+  pluck(1) %>% 
+  unique() %>%
+  glue::glue_collapse(", ", last = " and ")
+}
+
+get_auc <- function(Time, model, version = "1") {
+  data_auc_ci %>% 
+  filter(time == Time, Model == model) %>% 
+  select(ends_with(version)) %>% 
+  pluck(1)
+}
+```
 
@@
 Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population.
-We thus developed a parsimonious model to predict 90-day mortality after elective THA.
+We thus developed a parsimonious model to predict PJI within 90 days and two years after elective THA.
 
-**PARTICIPANTS AND METHODS: ** We analysed XXX patients registered in the Swedish hip arthroplasty register 2008 - 2015 and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+**PARTICIPANTS AND METHODS: ** We analysed `r nmbr(N_shar)` patients recorded in the Swedish hip arthroplasty register `r period` and obtained data on co-morbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery.
-Models were internally validated by bootstrapping and externally validated on a cohort of XXX patients from XXX.
-We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration plots.
+Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**.
+We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.
 
-**RESULTS: ** The unadjusted cumulative incidence of PJI was XXX within 90 days and XXX within two years in the Swedish derivation cohort.
-A prediction model for PJI within 90 days combining the underlying diagnosis for THA, the presence of psoriatic or CNS disease, cancer, substance abuse, body mass index, and gender had good predictive performance internally (AUC=XXX, 95% CI: XXX to XXX) and externally (AUC=XXX, 95% CI: XXX to XXX), and a similar model to predict PJI within two years had an AUC of XXX (95% CI: XXX to XXX) internally and XXX (95% CI: XXX to XXX) externally.
-These models were superior to traditional models based on American Society for Anesthesiologists (ASA) classification (AUC=XXX, 95% CI: XXX to XXX), the Charlson (AUC=XXX, 95% CI: XXX to XXX) or Elixhauser (AUC=XXX, 95% CI: XXX to XXX) co-morbidity indices.
+**RESULTS: ** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort.
+A prediction model for PJI within 90 days combined `r predictors("90 days")` and had `r get_auc("90 days", "Reduced model", "2")` internally  and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
+A similar model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**.
+These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` co-morbidity indices.
 
@@
 Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
-Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context.
-Although high ASA class is a risk factor associated with development of PJI the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on 17 or 31 International Classification of Diseases (ICD)-codes, respectively, but they are cumbersome be use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of co-morbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context.
+Although high ASA class is a risk factor associated with development of PJI, the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012]
+The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
 Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
@@
 We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
-The best models were then validated externally on a cohort from the XXX.
+The best models were then validated externally on a cohort from the **XXX**.
 
@@
 
-The derivation cohort was obtained from patients registered in the SHAR from 2008 to 2015.
+The derivation cohort was obtained from patients recorded in the SHAR `r period`.
 The starting point was chosen because registration of ASA class and body mass index (BMI) in the SHAR was initiated in 2008, and we had no access to data from the national patient register (NPR) after 2015.
-Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.
-The diagnosis underlying the index THA procedure (categorized into primary osteoarthritis, secondary osteoarthritis, pediatric hip disorder, avascular necrosis of the femoral head [AVN], and inflammatory joint disease), type of THA (cemented, uncemented hybrid), age at index surgery, gender, BMI, ASA class, type of hospital, and year of surgery were obtained from the SHAR.
-Data on co-morbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
+Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
+Data on co-morbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level  of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
-The SHAR has completeness of 96-98 % and 100% coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99% and the positive predictive value is 85-95%.[@Ludvigsson2011] Death dates were linked from the national population register.
-Patients with missing information on or an ASA class of IV and above were excluded, as were patients with missing information on BMI or a BMI above 50.
+The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
+Only the last operated hip was considered for patients with bilateral THA,[@Bulow2020] and only patients who survivied at least two years aftert surgery.
+Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a BMI above 50.
 Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.
@@
 
-The Charlson and Elixhauser co-morbidity indices are based on the presence of ICD-10 codes grouped into main categories, 17 for the Charlson 15 and 31 for the Elixhauser index.[@Quan2005] Diagnostic categories from both indices were combined to define 21 categories of co-morbidity (Tab.
-1) that were chosen in order to be easily identifiable in a clinical setting.
-Some co-morbidities defined by either the Charlson or the Elixhauser co-morbidity index were not present in any patient affected by PJI in the derivation cohort, and these diagnoses were therefore excluded as potential predictors.
-PJI was defined by the occurrence of any ICD- or NOMESCO-code indicative of this adverse event, and the time point at which any of these codes was registered allowed for the distinction between PJI occurring within 90 days or within two years .
+## Definitions of co-morbidity and outcomes
+
+The Charlson, Elixhauser and Rx Risk V co-morbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of co-morbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+\@ref(tab:tabcategorization)).
+Some of the combined co-morbidities were observed for less than 10 patients either with or without PJI.
+These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
+PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in NPR within 90 days or 2 years (Tab.
+\@ref(tab:infcodes)), or if the patient was reported to SHAR, within the same time frame, as reoperated due to infection.
 
@@
 
-The cumulative incidence of PJI was estimated using the Kaplan-Meier method.
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age and BMI were normalized prior to modeling, and the estimated coefficients thus indicate variable importance on the same scale as categorical variables.
-1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor.
-Potential predictors were identified using piece-wise linear regression and kept as model candidates, regressions were iterated 100 times, and potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
-Even more parsimonious models with potential predictors chosen at least XXX out of 100 times were evaluated for comparison.
-Univariable prediction models based on either ASA class, the Charlson or the Elixhauser co-morbidity index were fitted for comparison.
-Age and BMI were either treated as continuous variables in main effect models, or modelled using restricted cubic splines with two or three knots.
-In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI this variable was also treated as a categorical variable divided into the six BMI intervals suggested by the World Health Organization (WHO; <18.5: underweight, 18.5-24.9: normal, 25-29.9: overweight, 30-34.9: class I obesity, 35.0-39.9: class II obesity, and ???40: class III obesity).
-Odds ratios for the final models were estimated with 95% confidence intervals.
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age was normalized prior to modeling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
+In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI, this variable was treated as a categorical variable divided into four BMI intervals based on the six intervals propsed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
+
+For each outcome (PJI within 90 days or 2 years), 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor.
+Those coefficients were ranked based on their absolute values as a measure of  variable importance.
+Piece-wise linear regression was applied to find a break-point for which influential predictors had their estimated values above.
+Such predictors scored a point and the whole process was repeated 100 times.
+Potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
+Even more parsimonious models with potential predictors chosen at least 33 out of 100 times were evaluated for comparison.
+Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V co-morbidity indices were fitted for comparison.
+Odds ratios for the final models were estimated with 95 % confidence intervals.
 
@@
 Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation).
-Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95% confidence limits that were based on percentiles from 2,000 non-parametric bootstrap samples.
+Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples.
 The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism.
-Predicted probabilities and observed proportions were plotted with 95% confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a XXX cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to XXX.
+Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**.
 Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.
@@
 We built an online web calculator available at https XXX to be used in clinical practice.
-R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny was used.
+R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.
 
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and XXX.
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and **XXX**.
 
@@
 
+```{r}
+
+males <- 
+  sprintf(
+    "%.0f %%, n = %s", 
+    mean(df$P_Sex == "Male") * 100,
+    format(sum(df$P_Sex == "Male"), big.mark = ",")
+)
+
+age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
+
+comorb <- function(x) {
+  sprintf("%.0f %% (n = %s)", mean(x) * 100, nmbr(sum(x)))
+}
+```
+
 ## Study participants
 
-We included XXX patients (XXX% females) with mean age of XXX (SD XXX) in the derivation cohort from the SHAR (Fig.
-1, left panel).
-The cumulative incidence of PJI within 90 days was XXX% and XXX% within two years (Table 2).
-The proportion of patients with ASA class III was XXX%, XXX% had at least one comorbidity according to the Charlson co-morbidity index, and XXX% had at least one comorbidity defined by the Elixhauser index.
-In addition, XXX patients from XXX were included as an external validation cohort (Fig.
-1, right panel).
+We included `r nmbr(N_shar)` patients (`r males` males) with mean age `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
+`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients had PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab.
+\@ref(tab:tab1)).
+The proportion of patients with ASA class III was
+`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson co-morbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V.
+In addition, **XXX** patients from **XXX **were included as an external validation cohort (Fig.
+\@ref(fig:flowchart), right panel).
 
@@
 
-The main model for the prediction of PJI within 90 days in the Swedish cohort included seven predictor variables, one of which was the underlying diagnosis for THA surgery.
-Other predictors remaining in the main model were the presence of psoriatic and CNS diseases, cancer, substance abuse, BMI, and gender (Tab.
-XXX a).
-Predictors associated with a statistically significant risk increase were AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender.
-The main model for the predicition of PJI within 90 days had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) had less discriminative power (Fig.
-XXX, XXX).
-Calibration was good for the developed main model (Fig.
-XXX).
-The main model for PJI within two years included the same predictors, with the addition of the presence of rheumatic disease (Tab.
-XXX b).
-Predictors conferring a statistically significant risk increase were those mentioned as predictive for PJI within 90 days, and additionally included the presence of psoriatic, rheumatic, or CNS disease, and cancer.
-The main model for the prediction of PJI within two years had an AUC of XXX (95%CI: XXX-XXX), whereas ASA class (AUC: XXX, 95%CI: XXX-XXX), the Charlson co-morbidity index (AUC: XXX, 95%CI: XXX-XXX), and the Elixhauser co-morbidity index (AUC: XXX, 95%CI: XXX-XXX) again had less discriminative power (Fig.
-XXX, XXX).
-Calibration was good for the developed main model (Fig.
-XXX).
+Patietns with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling.
+ROC-curves (Fig.
+\@ref(fig:rocs)) and AUC-values (Fig.
+\@ref(fig:aucci)) for the main and reduced models were no different.
+We therefore focus on the reduced models due to simplicity.
+The correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
+Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (\@ref(fig:separation)).
+The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r predictors("90 days")` (Tab.
+\@ref(tab:coefs)).
+Predictors associated with a statistically significant risk increase were **AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender**.
+The reduced model for the predicition of PJI within 90 days had `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative power (Fig.
+\@ref(fig:rocs) and \@ref(fig:aucci)).
+Calibration was good for the reduced model (Fig.
+\@ref(fig:calibration), left panel).
+The reduced model for PJI within two years included the same predictors, with the addition of the **presence of rheumatic disease** (Tab.
+\@ref(tab:coefs)).
+Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, and additionally included the presence of psoriatic, rheumatic, or CNS disease, and cancer**.
+The reduced model for the prediction of PJI within two years had `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative power (Fig.
+\@ref(fig:rocs) and \@ref(fig:aucci)).
+Calibration was good for the reduced model (Fig.
+\@ref(fig:calibration), right panel).
 
@@
 
-We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson or Elixhauser co-morbidity indices.
-A similarly developed model to predict PJI within two years included the same predictors as the model developed to predict PJI within 90 days, but added the presence of rheumatic disease to the list of predictor variables.
+We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V co-morbidity indices.
+A similarly developed model to predict PJI within two years included the same predictors as the model developed to predict PJI within 90 days, but added the presence of **rheumatic disease** to the list of predictor variables.
 This model also performed better than the established co-morbidity measures in terms of AUC and calibration.
@@
 
-Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into in a clinical setting, either by use of the formula given in Tab.
-XXX, or by accessing a simple web calculator online.
-For example, ...
+Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simpole formula $\hat p = (e^{\hat \alpha + \mathbf{\hat \beta X}})/(1 + e^{\hat \alpha + \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ the estimated intercept and coefficients (Tab.
+\@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ a vector of corresponding patient characteristics.
+For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.
 
@@
 This seeming discrepancy may be explained by the fact that other physicians may have identified patients as suffering from any of these conditions, thereby assigning the indicative ICD codes, whereas the operating surgeon may have classified the underlying reason for THA as something other than inflammatory joint disease.
-Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.[@Tan2018] In accordance with observational studies9 we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
+Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
 Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.[@Wingert2016a] Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
-High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.7, 31 In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
-Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,11 our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
-THA fixation using cement without antibiotics confers an increased risk of PJI,24 but such cement brands were not in clinical use in Sweden during the studied period.
+High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
+Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,[@Rasouli2014a] our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
+THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
 We found no indication that the type of fixation was associated with the risk of PJI, and this variable was also not selected as a potential predictor variable in any of the iterations during model development.
@@
 
-Our study is hampered by the limitations inherent to observational research, such as misclassification, missing information, and .
+Our study is hampered by the limitations inherent to observational research, such as misclassification, missing information, and XXX.
 The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
@@
 
-NPH initiated the study, OR managed the ethical review board application in Sweden, EB developed the statistical model, XXX performed external validation with data from XXX.
+NPH initiated the study, OR managed the ethical review board application in Sweden, EB developed the statistical model, **XXX** performed external validation with data from **XXX**.
 NPH and EB drafted, and all authors edited and finalized the manuscript.
@@
 
-```{r, child = "figs.Rmd", eval=FALSE}
+<!-- FIGURES -->
+
+```{r flowchart, fig.cap = "Flowchart with inclusion criteria and number of patients.
+Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation (left).
+Data from XXX were used for external validation (right)."}
+knitr::include_graphics("../graphs/flowchart.png")
 ```
 
-```{r, child = "tabs.Rmd", eval=FALSE}
+##### PAGE BREAK
+
+```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity tp illustrate discriminative abilities of the different models.
+The reduced model perform almost as good as the main model for prediction of PJI within 90 days after surgery, and as good for PJI within 2 years."}
+knitr::include_graphics("../graphs/roc.png")
+```
+
+##### PAGE BREAK
+
+```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
+The reduced model performs no different than the main model, but both of these perform better than traditional co-morbidity measures alone."}
+
+knitr::include_graphics("../graphs/auc_ci.png")
+```
+
+##### PAGE BREAK
+
+```{r sep, fig.cap = "The vast majority of patients did not have PJI.
+Hence, the blue bars dominates the histograms (upper panels; note the scales).
+Normalized density plots reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
+knitr::include_graphics("../graphs/separation.png")
+```
+
+##### PAGE BREAK
+
+```{r calibration, fig.cap = "This figure illustrates calibration between observed proportions and predicted probabilities with 95 % confidence intervals.
+Calibration is good for predictions up to 5 %, although higher probabilities tend to over-estimate the observed outcome.
+"}
+knitr::include_graphics("../graphs/calibration.png")
+```
+
+<!-- Tables -->
+
+##### PAGE BREAK
+
+```{r tabcategorization}
+load("../cache/tab_categorization.RData")
+
+tab_categorization %>%
+  ungroup() %>% 
+  mutate(
+    `Comorbidities by groups` = 
+      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
+  ) %>% 
+  knitr::kable(
+    caption = "Categorization of co-morbidities from the Charlson, Elixhauser and Rx RIsk V co-morbidities. (CNS = central nervous system.)")
+
+```
+
+##### PAGE BREAK
+
+```{r infcodes}
+library(coder)
+
+icd10_codes <- 
+  summary(
+      hip_ae_hailer, 
+      "icd10se", 
+      cc_args = list(regex = "icd10")
+    )$codes_vct$Infection %>% 
+    paste(collapse = ", ")
+
+nomesco_codes <- 
+  summary(
+      hip_ae_hailer, 
+      "kva", 
+      cc_args = list(regex = "kva")
+    )$codes_vct$Infection %>% 
+    paste(collapse = ", ")
+
+tibble(
+  classification = c("ICD-10", "NOMESCO"),
+  codes = c(icd10_codes, nomesco_codes)
+) %>% 
+  knitr::kable(
+    caption = "Codes identifying infection if recorded in the national patient register within 90 days or 2 years after THA respectively."
+  )
+
+```
+
+##### PAGE BREAK
+
+```{r tab1}
+source("../lib/clean_names.R")
+
+knitr::kable(
+  mutate(table1, what = gsub("Heart infarct", "Myocardial infarction", what)), 
+  caption = "Patient characteristics in the model derivation cohort.
+Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years).
+BMI = Body mass index.
+ASA class = American Society for Anesthesiologists classification.
+CNS = central nervous system.")
+```
+
+##### PAGE BREAK
+
+```{r brlprop}
+load("../cache/infection_data.RData")
+
+infection_data %>% 
+  select(outcome, time, coefs_selected) %>% 
+  mutate(coefs_selected = map(coefs_selected, "brlasso_tbl_selected")) %>% 
+  unnest("coefs_selected") %>%
+  filter(outcome == "infection") %>% 
+  select(-outcome) %>% 
+  pivot_wider(names_from = "time", values_from = "n") %>% 
+  mutate(
+    variable = clean_names(variable) %>% 
+      {gsub("\\.", " ", .)} %>% 
+      {gsub("Rheumatidisease", "Rheumatic disease", .)}
+  ) %>% 
+  select(Predictor = variable, `90 days` = `90d`, `2 years` = `2y`) %>% 
+knitr::kable(
+  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model.
+Variables chosen at least 33 times out of 100 were kept in the reduced model as well.
+(BMI = body mass index.
+ASA class = American Society for Anesthesiologists classification.
+CNS = central nervous system.)")
+```
+
+##### PAGE BREAK
+
+```{r coefs}
+tbl_coefs %>% 
+  mutate(
+    time = replace(as.character(time), duplicated(time), ""),
+    beta = round(beta, 2)
+  ) %>% 
+knitr::kable(
+  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models.
+BMI = body mass index ("under/normal weight as baseline).
+CNS = central nervous system.
+Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
+)
 ```

### 2020-05-11 munge/02-survdata.R

@@ -30,7 +30,6 @@ df <-
         "class II-III obesity"),
       right = FALSE
     ),
-    P_BMI = relevel(P_BMI, "overweight"),
 
     # Survival
     stime   =

### 2020-05-11 munge/07-compare_models.R

@@ -3,18 +3,15 @@ df <-
   mutate(
     P_SurgYear                = P_SurgYear - min(P_SurgYear),
     outcome_infection_90d_f   = as.factor(outcome_infection_90d),
-    outcome_infection_2y_f    = as.factor(outcome_infection_2y),
-    P_PMI = factor(P_BMI, c("under/normal weight", "overweight",
-                                "class I obesity", "class II-III obesity"
-    ))
+    outcome_infection_2y_f    = as.factor(outcome_infection_2y)
   )
 
 # "unregularized least-square fit restricted to variables in J" /[@Bach2008]
 form <- function(nms) {
   paste("outcome ~", paste(unique(gsub(paste0(
     "_TRUE.|_X[23]|_(I|II|III|Man|Kvinna|Male|Female|married|single|widow.widower|",
-    "under.normal.weight|class.II.III.obesity|class.I.obesity|",
-    "Idiopathic.necrosis|overweight|Inflammatory.joint.disease|Secondary.osteoarthritis)"), "",
+    "under.normal.weight|overweight|class.I.obesity|class.II.III.obesity|",
+    "Idiopathic.necrosis|Inflammatory.joint.disease|Secondary.osteoarthritis)"), "",
     nms)), collapse = " + ")
   )
 }

### 2020-05-11 src/examples.R

@@ -1,0 +1,27 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+examples <-
+  tibble(
+    c_psoriasis      = c(FALSE,                     TRUE),
+    P_BMI            = c("under/normal weight",     "overweight"),
+    P_DiaGrp         = c("Primary osteoarthritis",  "Secondary osteoarthritis"),
+    P_Sex            = c("Female",                  "Male"),
+    c_cns_disease    = c(FALSE,                     FALSE),
+    c_liver_disease  = c(FALSE,                     FALSE),
+  )
+
+p <- predict(fit_brl$fit[[1]], newdata = examples, type = "response")
+
+examples$p <- predict(fit_brl$fit[[1]], newdata = examples, type = "response")
+
+examples <-
+  examples %>%
+  mutate(
+    p = sprintf("%.1f %%", p * 100),
+    desc = c(
+      "a female of normal weigth with primary osteoarthritis and no co-morbidities",
+      "a male with overweight, secondary osteoarthritis and proriasis"),
+    desc = sprintf("%s would have a probability of %s", desc, p)
+  )
+
+cache("examples")

### 2020-05-11 src/fig_auc.R

@@ -13,8 +13,13 @@ data_auc_ci <-
   select(time, Model, starts_with("AUC")) %>%
   mutate(
     Model = reorder(Model, AUC_est),
-    time = factor(time, c("90d", "2y"), c("90 days", "2 years"))
+    time  = factor(time, c("90d", "2y"), c("90 days", "2 years")),
+    ci    = sprintf("95 %% CI: %.2f to %.2f", AUC_lo, AUC_hi),
+    text1  = sprintf("(AUC = %.2f, %s)", AUC_est, ci),
+    text2 = sprintf("AUC = %.2f (%s)", AUC_est, ci),
   )
+
+cache("data_auc_ci")
 
 annotates <-
   data_auc_ci %>%

### 2020-05-11 src/fig_calibration.R

@@ -13,7 +13,7 @@ df_calibration <-
   infection_data %>%
   select(time, all_models) %>%
   unnest("all_models") %>%
-  filter(Model == "Main model") %>%
+  filter(Model == "Reduced model") %>%
   mutate(
     cal_belt = furrr::future_map(
       obspred, ~ givitiR::givitiCalibrationBelt(

### 2020-05-11 src/fig_separation.R

@@ -7,7 +7,7 @@ obspred <-
   select(outcome, time, Model, obspred, AUC_lo) %>%
   unnest(obspred) %>%
   filter(
-    Model == "Main model"
+    Model == "Reduced model"
   ) %>%
   mutate(time = factor(time, c("90d", "2y"), c("90 days", "2 years")))
 

### 2020-05-11 src/tab_coefs.R

@@ -6,7 +6,7 @@ tbl_coefs <-
   infection_data %>%
   unnest("all_models") %>%
   filter(
-    Model == "Main model"
+    Model == "Reduced model"
   ) %>%
   select(time, tidy) %>%
   unnest(tidy) %>%
@@ -17,41 +17,7 @@ tbl_coefs <-
     `OR 95 % CI` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
     p    = pvalue_format(.01)(p.value)
   ) %>%
-  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(term == "(intercept)", "", .))
+  arrange(time) %>%
+  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(term == "(Intercept)", "", .))
 
 cache("tbl_coefs")
-
-
-# List of coefficients for text -------------------------------------------
-
-
-set_first <- function(coefs, x) {
-  if (x %in% coefs) c(x, setdiff(coefs, x))
-  else coefs
-}
-
-coefs_print_string <- function(name) {
-  all_vars <-
-    filter(all_models, Model == !!name) %>%
-    {.$tidy[[1]]$term[-1]}
-
-  c_vars     <- all_vars[startsWith(all_vars, "c_")]
-
-  basic_vars <-
-    setdiff(all_vars, c_vars) %>%
-    clean_names(firstupper = FALSE, lvls = FALSE) %>%
-    set_first("ASA class") %>%
-    set_first("sex") %>%
-    set_first("age")
-
-  c_vars <- clean_names(c_vars, FALSE)
-
-  paste(
-    glue::glue_collapse(basic_vars, ", "),
-    "and the precense of",
-    glue::glue_collapse(c_vars, ", ", last = " and ")
-  )
-}
-
-# coefs_print <- coefs_print_string("BRL (age as main effect)")
-# cache("coefs_print")

### 2020-05-11 src/tab_table1.R

@@ -65,7 +65,7 @@ table1 <-
     what = trimws(ifelse(level == "", gsub(" = TRUE", "", what), ""))
   ) %>%
   mutate_at(vars(`TRUE`, `FALSE`), zero) %>%
-  select(what, level, `PJI within 2 years` = `TRUE`, `No PJI` = `FALSE`)
+  select(what, level, `PJI within 2 years` = `TRUE`, `No PJI` = `FALSE`, Total)
 
 
 cache("table1")

### 2020-05-13 reports/article.Rmd
manuscript updates due to NH:s corrections. Fix #25.

@@
 
-# Extract predictors for specified period
-# Needs additional formatting
-predictors <- function(period) {
-  tbl_coefs %>% 
-  filter(
-    time == period,
-    term != "(Intercept)"
-  ) %>% 
-  transmute(
-    term = tolower(term),
-    term = replace(
-      term, 
-      startsWith(term, "diagnosis"), 
-      "the underlying diagnosis for THA"
-    ),
-    term = replace(term, startsWith(term, "bmi"), "body mass index"),
-    term = gsub("male ", "", term)
-  ) %>% 
-  pluck(1) %>% 
-  unique() %>%
-  glue::glue_collapse(", ", last = " and ")
-}
-
 get_auc <- function(Time, model, version = "1") {
@@
 **RESULTS: ** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort.
-A prediction model for PJI within 90 days combined `r predictors("90 days")` and had `r get_auc("90 days", "Reduced model", "2")` internally  and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
-A similar model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**.
+A prediction model for PJI within 90 days combined `r coefs_text["90 days"]`, resulting in an `r get_auc("90 days", "Reduced model", "2")` internally  and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
+A model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**.
 These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` co-morbidity indices.
@@
 The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
-Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been proposed based on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low predictive and discriminative ability of the prediction models, or the absence of external model validation.
+Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low  discriminative ability or low accuracy of the prediction models, or the absence of external model validation.
 Thus, there is an urgent need for improved prediction models of PJI.
@@
 \@ref(tab:tabcategorization)).
-Some of the combined co-morbidities were observed for less than 10 patients either with or without PJI.
+Some of the combined co-morbidities were observed for less than 10 patients with or without PJI.
 These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
@@
 Those coefficients were ranked based on their absolute values as a measure of variable importance.
-Piece-wise linear regression was applied to find a break-point for which influential predictors had their estimated values above.
-Such predictors scored a point and the whole process was repeated 100 times.
+Piece-wise linear regression was applied to find a break-point, and influential predictors with absolute coefficients value above this pointm, scored a point.
+The whole process was repeated 100 times.
 Potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
@@
 Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples.
-The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism.
+The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism,
+but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
 Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**.
@@
 
-We built an online web calculator available at https XXX to be used in clinical practice.
+We built an online web calculator available at https **XXX** to be used in clinical practice.
 R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.
@@
 ```{r}
-
-males <- 
-  sprintf(
-    "%.0f %%, n = %s", 
-    mean(df$P_Sex == "Male") * 100,
-    format(sum(df$P_Sex == "Male"), big.mark = ",")
-)
-
 age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
@@
 
-We included `r nmbr(N_shar)` patients (`r males` males) with mean age `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
-`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients had PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab.
+We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
+`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients developed a PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab.
 \@ref(tab:tab1)).
@@
 `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson co-morbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V.
-In addition, **XXX** patients from **XXX **were included as an external validation cohort (Fig.
+In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig.
 \@ref(fig:flowchart), right panel).
@@
 
-Patietns with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling.
+Patients with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling.
 ROC-curves (Fig.
@@
 \@ref(fig:aucci)) for the main and reduced models were no different.
-We therefore focus on the reduced models due to simplicity.
-The correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (\@ref(fig:separation)).
-The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r predictors("90 days")` (Tab.
+For simplicity, we therefore focus on the reduced models.
+Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (Fig.
+\@ref(fig:sep)).
+
+The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text["90 days"]` (Tab.
 \@ref(tab:coefs)).
-Predictors associated with a statistically significant risk increase were **AVN or secondary osteoarthritis as the underlying reason for THA surgery, the presence of CNS disease, obesity, and male gender**.
-The reduced model for the predicition of PJI within 90 days had `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative power (Fig.
+Predictors associated with a statistically significant risk increase were **secondary osteoarthritis as the underlying reason for THA surgery, male gender, obesity, and the presence of psoriasis, osteoporosis, Paget's or liver disease**.
+This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
-Calibration was good for the reduced model (Fig.
+The reduced model had good accuracy as seen by the calibration plot (Fig.
 \@ref(fig:calibration), left panel).
-The reduced model for PJI within two years included the same predictors, with the addition of the **presence of rheumatic disease** (Tab.
+
+The reduced model for PJI within two years included the same predictors, with the addition of `r coefs_text["2 years"]` (Tab.
 \@ref(tab:coefs)).
-Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, and additionally included the presence of psoriatic, rheumatic, or CNS disease, and cancer**.
-The reduced model for the prediction of PJI within two years had `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative power (Fig.
+Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, excluding osteoporosis or Paget's disease, and additionally included an underlying diagnosis of AVN or inflammatory joint disease, and the presence of  CNS disease**.
+The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
@@
 We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V co-morbidity indices.
-A similarly developed model to predict PJI within two years included the same predictors as the model developed to predict PJI within 90 days, but added the presence of **rheumatic disease** to the list of predictor variables.
+A similarly developed model to predict PJI within two years included similar predictors as the model developed to predict PJI within 90 days, but added `r coefs_text["2 years"]` to the list of predictor variables.
 This model also performed better than the established co-morbidity measures in terms of AUC and calibration.
@@
 
-Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simpole formula $\hat p = (e^{\hat \alpha + \mathbf{\hat \beta X}})/(1 + e^{\hat \alpha + \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ the estimated intercept and coefficients (Tab.
-\@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ a vector of corresponding patient characteristics.
+Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab.
+\@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics.
 For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.
@@
 
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred an about two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] but this predictor variable only reached borderline statistical significance in our model of 90-day PJI.
-However, the presence of psoriatic disease, and, for the outcome PJI within two years, the presence of rheumatic disease, were both statistically significant predictors of outcome in our models.
-This seeming discrepancy may be explained by the fact that other physicians may have identified patients as suffering from any of these conditions, thereby assigning the indicative ICD codes, whereas the operating surgeon may have classified the underlying reason for THA as something other than inflammatory joint disease.
-Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both main models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator.[@Wingert2016a] Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred a nearly two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] and this predictor variable was statistically significance in our model of PJI within two years.
+
+Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI within two years, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
+
+Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018]
+Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015]Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+
+Some previously mentioned risk factors for developing PJI were not selected during our model development.
 High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
-Similarly, although mentioned as a risk factor for surgical site infection after joint arthroplasty,[@Rasouli2014a] our prediction models based on the Charlson co-morbidity index alone resulted in poor predictive power.
+
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a]
+20 but this comorbidity did not reach statistical significance in our models.
+
 THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
@@
 
+Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the co-morbidity indices developed by Charlson or Elixhauser and the RxRisk V index alone, resulted in poor predictive power.
+This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019]
+
 ## Strengths and limitations
 
-Our study is hampered by the limitations inherent to observational research, such as misclassification, missing information, and XXX.
+Our study is hampered by the limitations inherent to observational research, such as misclassification, and residual confounding due to known confounders that we had no information on, or due to unknown confounders.
+Importantly, some previously identified risk factors for PJI were not included in our models, mostly because we lacked detailed information on laboratory findings and non-prescribed medications.
+
 The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
 This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings.
-Use of TNF-a blockers may be associated with the risk of developing PJI,[@Gilson2010; @Momohara2011] but we had no information on medication.
-Similarly, intra-articular steroid application may enhance the risk of developing PJI,[@Cancienne2015] but such manipulations would not be identifiable within a Swedish registry setting.
+Use of TNF-a blockers may be associated with the risk of developing PJI.[@Gilson2010; @Momohara2011] Similarly, intra-articular steroid application may enhance this risk,[@Cancienne2015] but injections given by physicians would not necessarily be identifiable within a Swedish registry setting.
 Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter.
-Taken together, the inclusion of some or all of the variables mentioned above-unfortunately unavailable to us-might improve the predictive power of PJI risk assessment models.
 The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
-The question of whether there is a genetic predisposition towards developing orthopaedic infections has only been ...
+
+Taken together, the inclusion of some or all of the variables mentioned above -- unfortunately unavailable to us -- might improve the predictive power of PJI risk assessment models.
 
@@
 ```{r coefs}
-tbl_coefs %>% 
+tbl_coefs_present %>% 
   mutate(

### 2020-05-13 lib/clean_names.R

@@ -6,7 +6,6 @@ clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
   {gsub("ASA",     "ASA class: ", .)} %>%
   {gsub("BMI",     "BMI: ",       .)} %>%
   {gsub("aids",    "AIDS/HIV",    .)} %>%
-  {gsub("cns",     "CNS disease", .)} %>%
   {paste0(toupper(substr(., 1, 1)), substring(., 2))}
 }
 

### 2020-05-13 munge/02-survdata.R

@@ -58,7 +58,7 @@ df <-
       c(
          "Primary osteoarthritis",
          "Secondary osteoarthritis",
-         "Childhood disease",
+         "Sequelae after childhood hip disease",
          "Avascular necrosis of the femoral head (AVN)",
          "Inflammatory joint disease",
          "Secondary osteoarthritis"                   # merge of categories

### 2020-05-13 src/tab_coefs.R

@@ -1,7 +1,7 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
-# Table to present --------------------------------------------------------
 
+# Baseline levels for cqategorical variables -----------------------------------
 baseline <-
   crossing(
     time = factor(c("90 days", "2 years")),
@@ -14,6 +14,7 @@ baseline <-
   mutate(beta = 0, `OR 95 % CI` = "(baseline)", p = "")
 
 
+# Prepare temp table ------------------------------------------------------
 
 tbl_coefs <-
   infection_data %>%
@@ -25,8 +26,8 @@ tbl_coefs <-
   unnest(tidy) %>%
   transmute(
     time = factor(time, c("90d", "2y"), c("90 days", "2 years")),
-    term = gsub("Male sex", "Sex: Male", term),
     term = clean_names(term),
+    term = gsub("Male sex", "Sex: Male", term),
     beta = log(estimate),
     `OR 95 % CI` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
     p    = pvalue_format(.01)(p.value)
@@ -34,7 +35,16 @@ tbl_coefs <-
   bind_rows(baseline) %>%
   separate(term, c("variable", "level"), sep = ": ") %>%
   arrange(time, variable, `OR 95 % CI`) %>%
-  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(variable == "(Intercept)", "", .)) %>%
+  mutate_at(vars(`OR 95 % CI`, p), ~ if_else(variable == "(Intercept)", "", .))
+
+
+cache("tbl_coefs")
+
+
+# Table to present in a nice way -----------------------------------------------
+
+tbl_coefs_present <-
+  tbl_coefs %>%
   group_by(time) %>%
   mutate(variable = replace(variable, duplicated(variable), "")) %>%
   ungroup() %>%
@@ -43,4 +53,42 @@ tbl_coefs <-
     time = replace(time, duplicated(time), "")
   )
 
-cache("tbl_coefs")
+cache("tbl_coefs_present")
+
+
+# List covariates as text -------------------------------------------------
+
+coefs_text <-
+  tbl_coefs %>%
+  filter(
+    variable != "(Intercept)",
+    !duplicated(tibble(variable, level)) # ONly unique factors for 2 years
+  ) %>%
+  mutate(fct = !is.na(level)) %>%
+  select(time, fct, variable) %>%
+  mutate(
[…]

### 2020-05-28 reports/article.Rmd
M

@@
       "icd10se", 
-      cc_args = list(regex = "icd10")
+      cc_args = list(regex = "icd10", stop = TRUE)
     )$codes_vct$Infection %>% 
@@
       "kva", 
-      cc_args = list(regex = "kva")
+      cc_args = list(regex = "kva", stop = TRUE)
     )$codes_vct$Infection %>% 

### 2020-05-28 munge/01-outcome.R

@@ -21,7 +21,7 @@ ae_npr <- function(df_npr, regex, endday) {
     select(LopNr, P_SurgDate) %>%
     mutate(LopNr = as.character(LopNr)) %>%
     categorize(
-      df_kva,
+      df_npr,
       cc = "hip_ae_hailer",
       id = "LopNr",
       codify_args = list(
@@ -30,7 +30,8 @@ ae_npr <- function(df_npr, regex, endday) {
       ),
       cc_args = list(
         regex = regex,
-        tech_names = TRUE
+        tech_names = TRUE,
+        stop = TRUE
       )
     )
 }
@@ -45,15 +46,15 @@ outcome_reop <- function(endday) {
   df_reops %>%
     mutate_at(vars(P_SurgDate, R_SurgDate), as.Date) %>%
     filter(R_SurgDate - P_SurgDate < endday) %>%
-    distinct(LopNr, P_Side, .keep_all = TRUE) %>%
-    transmute(
-      LopNr, P_SurgDate,
+    # Räcker med outcome vid en av ev flera reop under perioden.
+    group_by(LopNr, P_SurgDate) %>%
+    summarise(
       reop_luxation =
-        IND_ReSurgReason1 == "Luxation, instabilitet, subluxation" |
-        IND_ReSurgReason2 == "Luxation, instabilitet, subluxation",
+        any(IND_ReSurgReason1 == "Luxation, instabilitet, subluxation") |
+        any(IND_ReSurgReason2 == "Luxation, instabilitet, subluxation"),
       roep_infektion =
-        IND_ReSurgReason1 == "Infektion" |
-        IND_ReSurgReason2 == "Infektion"
+        any(IND_ReSurgReason1 == "Infektion") |
+        any(IND_ReSurgReason2 == "Infektion")
     )
 }
 
@@ -68,7 +69,12 @@ df_outcome <- function(outcome_icd10, outcome_kva, outcome_reop) {
   as_tibble(outcome_icd10) %>%
     left_join(as_tibble(outcome_kva), by = c("LopNr", "P_SurgDate")) %>%
     left_join(outcome_reop,           by = c("LopNr", "P_SurgDate")) %>%
-    transmute(
+    mutate(across(
+      c(roep_infektion, reop_luxation, starts_with("hip_ae_hailer_regex")),
+      coalesce, FALSE
+      )
+    ) %>%
+    mutate(
       LopNr = as.numeric(LopNr),
       P_SurgDate,
       outcome_infection =

### 2020-05-28 munge/04-remove_empty_variables.R

@@ -130,6 +130,15 @@ model_data <-
 cache("model_data")
 
 
+
+# Tidiagre steg utgick från både infektion och luxation kombinerat.
+# JAg väljer nu att fortsätta med enbart infektion och hanteras luxation separat sernare
+
+infection_data <-
+  model_data %>%
+  dplyr::filter(outcome == "infection")
+
+
 # Save list of excluded co-morbidities for presentation
 excl_factors <-
   infection_data$lgl_text[[1]] %>%

### 2020-05-28 munge/05-BRLasso.R

@@ -1,12 +1,3 @@
-# Tidiagre steg utgick från både infektion och luxation kombinerat.
-# JAg väljer nu att fortsätta med enbart infektion och hanteras luxation separat sernare
-
-infection_data <-
-  model_data %>%
-  dplyr::filter(outcome == "infection")
-
-
-
 # Perform calculations (time consuming)
 
 best_coefs_fun <- function(df_model) {

### 2020-05-28 munge/07-compare_models.R

@@ -11,7 +11,7 @@ form <- function(nms) {
   paste("outcome ~", paste(unique(gsub(paste0(
     "_TRUE.|_X[23]|_(I|II|III|Man|Kvinna|Male|Female|married|single|widow.widower|",
     "under.normal.weight|overweight|class.I.obesity|class.II.III.obesity|",
-    "Idiopathic.necrosis|Inflammatory.joint.disease|Secondary.osteoarthritis)"), "",
+    "Avascular.necrosis.of.the.femoral.head..AVN.|Inflammatory.joint.disease|Secondary.osteoarthritis)"), "",
     nms)), collapse = " + ")
   )
 }

### 2020-06-03 reports/article.Rmd
have re-run the analysis and adjusted graphs accordingly

@@
 date: '`r Sys.Date()`'
-bibliography: H:/zotero_lib.bib
+bibliography: 
+#  H:/zotero_lib.bib
+  C:\\Users\\erik_\\Documents\\library.bib
 csl: the-lancet.csl
@@
       {gsub("\\.", " ", .)} %>% 
-      {gsub("Rheumatidisease", "Rheumatic disease", .)}
+      {gsub("Rheumatidisease", "Rheumatic disease", .)} %>% 
+      {gsub("Sex Male", "Male gender", .)}
   ) %>% 

### 2020-06-03 lib/BR_lasso_coefs.R

@@ -4,19 +4,24 @@
 #' variable importance to make any sense!
 #' @param df data.frame
 #' @return tibble with important variables and a measure thereof
-BR_lasso_coefs <- function(df) {
+I <- 0
+BR_lasso_coefs <- function(df, lambda) {
+  I <<- I + 1
+  i <<- 0
+  cat(paste0("\n", I, ": "))
+
   df %>%
     rsample::bootstraps(config$N_bots) %>%
     mutate(
       data  = map(splits, as_tibble),
-      lasso = map(data, lasso)
+      lasso = map(data, lasso, lambda)
     ) %>%
     unnest(lasso) %>%
     group_by(variable) %>%
     summarise(
-      impor = abs(mean(coef))
+      impor = abs(mean(coef)),
+      .groups = "drop"
     ) %>%
-    ungroup() %>%
     filter(impor > 0) %>%
     arrange(desc(impor)) %>%
     add_rowindex()

### 2020-06-03 lib/globals.R

@@ -1,14 +1,16 @@
-con <- shar_linkage_con()
+# con <- shar_linkage_con()
 
 # General settings
 memory.limit(1e10)
 set.seed(132456798)
-future::plan("multiprocess")
+future::plan(sequential)
+# future::plan("multiprocess", workers = 2) # future::availableCores()
 
 options(
   digits    = 2,
   na.action = "na.fail",
-  scipen    = 999
+  scipen    = 999,
+  repos     = "https://mran.microsoft.com/snapshot/2020-06-02"
 )
 
 # Add any project specific configuration here.

### 2020-06-03 lib/lasso.R

@@ -1,14 +1,18 @@
 # Apply lasso regression with CV for best lambda and return coeficient estimates
-lasso <- function(x) {
-  fit <- glmnet::cv.glmnet(
+lasso <- function(x, l) {
+  i <<- i + 1
+  if (i %% 10 == 0) cat(".")
+
+  fit <- glmnet::glmnet(
     select(x, -outcome) %>% as.matrix(),
     x$outcome,
-    family = "binomial"
+    family = "binomial",
+    lambda = l
   )
   as_tibble(
-    as.matrix(coef(fit$glmnet.fit, s = fit$lambda.min)),
+    as.matrix(coef(fit)),#, s = fit$lambda.min)),
     rownames = "variable"
   ) %>%
-    rename(coef = `1`) %>%
+    rename(coef = s0) %>%
     filter(variable != "(Intercept)")
 }

### 2020-06-03 munge/01-outcome.R

@@ -74,7 +74,7 @@ df_outcome <- function(outcome_icd10, outcome_kva, outcome_reop) {
       coalesce, FALSE
       )
     ) %>%
-    mutate(
+    transmute(
       LopNr = as.numeric(LopNr),
       P_SurgDate,
       outcome_infection =

### 2020-06-03 munge/05-BRLasso.R

@@ -1,19 +1,32 @@
 # Perform calculations (time consuming)
 
+if (!exists("infection_data"))
+  load("cache/infection_data.RData")
+
 best_coefs_fun <- function(df_model) {
+
+  lambda <- glmnet::cv.glmnet(
+    select(df_model, -outcome) %>% as.matrix(),
+    df_model$outcome,
+    family = "binomial"
+  )$lambda.min
+
   tibble(B = seq_len(config$Bmax)) %>%
   mutate(
-    coefs_all = future_map(B, ~ BR_lasso_coefs(df_model), .progress = TRUE),
+    coefs_all = map(B, ~ BR_lasso_coefs(df_model, lambda), .progress = TRUE),
     breaks    = map_dbl(coefs_all, break_p),
     coefs     = map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
   )
 }
 
+Sys.time()
+tmp1 <- best_coefs_fun(infection_data$df_model[[1]]); cache("tmp1"); Sys.time()
+tmp2 <- best_coefs_fun(infection_data$df_model[[2]]); cache("tmp2"); Sys.time()
 
 infection_data <-
   infection_data %>%
   mutate(
-    best_coefs_tmp = map(df_model, best_coefs_fun)
+    best_coefs_tmp = list(tmp1, tmp2)
   )
 
 cache("infection_data")

### 2020-06-03 munge/07-compare_models.R

@@ -63,6 +63,8 @@ compare_models <- function(brlasso_coefs, outcome) {
 
 # Add models --------------------------------------------------------------
 
+future::plan("multiprocess", workers = 2)
+
 infection_data <-
   infection_data %>%
   mutate(

### 2020-06-03 src/fig_auc.R

@@ -12,11 +12,11 @@ data_auc_ci <-
   unnest("all_models") %>%
   select(time, Model, starts_with("AUC")) %>%
   mutate(
-    Model = reorder(Model, AUC_est),
-    time  = factor(time, c("90d", "2y"), c("90 days", "2 years")),
-    ci    = sprintf("95 %% CI: %.2f to %.2f", AUC_lo, AUC_hi),
+    Model  = reorder(Model, AUC_est),
+    Period = factor(time, c("90d", "2y"), c("90 days", "2 years")),
+    ci     = sprintf("95 %% CI: %.2f to %.2f", AUC_lo, AUC_hi),
     text1  = sprintf("(AUC = %.2f, %s)", AUC_est, ci),
-    text2 = sprintf("AUC = %.2f (%s)", AUC_est, ci),
+    text2  = sprintf("AUC = %.2f (%s)", AUC_est, ci),
   )
 
 cache("data_auc_ci")
@@ -27,26 +27,33 @@ annotates <-
 
 # Figure ------------------------------------------------------------------
 
-ggplot(data_auc_ci, aes(Model, AUC_est)) +
-  geom_hline(yintercept = 0.7, color = "darkgreen", size = 1, linetype = "dashed") +
-  geom_pointrange(aes(ymin = AUC_lo, ymax = AUC_hi), size = 1) +
-  geom_text(aes(Model, value, label = round(value, 2)), data = annotates, nudge_x = .3, size = 3) +
+ggplot(data_auc_ci, aes(Model, AUC_est, group = Period, color = Period)) +
+  #geom_hline(yintercept = 0.7, color = "darkgreen", size = 1, linetype = "dashed") +
+  geom_pointrange(
+    aes(ymin = AUC_lo, ymax = AUC_hi),
+    size = .5,
+    position = position_dodge(width = .25)) +
+  geom_text(
+    aes(Model, value, label = round(value, 2)),
+    data = annotates,
+    #nudge_x = .3,
+    size = 3,
+    position = position_dodge(width = .85)
+  ) +
   coord_flip() +
   theme_minimal(15) +
   theme(
     axis.title.y = element_blank(),
-    legend.title = element_blank(),
-    legend.position = "none",
+    # legend.title = element_blank(),
+    legend.position = "bottom",
     panel.grid.major.y = element_blank(),
     panel.grid.minor.y = element_blank(),
-    panel.grid.major.x = element_blank(),
[…]

### 2020-06-03 src/fig_calibration.R

@@ -46,4 +46,4 @@ calplot <- function(xlim = c(0, 0.1), ylim = c(0, 0.1)) {
 }
 
 
-ggsave("graphs/calibration.png", calplot(), height = 10, width = 20 , units = "cm")
+ggsave("graphs/calibration.png", calplot(), height = 12, width = 20 , units = "cm")

### 2020-06-03 src/fig_separation.R

@@ -9,7 +9,10 @@ obspred <-
   filter(
     Model == "Reduced model"
   ) %>%
-  mutate(time = factor(time, c("90d", "2y"), c("90 days", "2 years")))
+  mutate(
+    time = factor(time, c("90d", "2y"), c("90 days", "2 years")),
+    # pred = factor(pred, c(TRUE, FALSE), c("PJI", "no PJI"))
+  )
 
 
 # Histogram ---------------------------------------------------------------
@@ -20,7 +23,7 @@ fig_separation_hist <-
   geom_histogram(
     aes(fill = obs),
     alpha = .5, position = "identity",
-    bins = 15
+    bins = 10
   ) +
   theme_minimal(15) +
   theme(
@@ -32,7 +35,7 @@ fig_separation_hist <-
   xlab("") +
   ylab(expression(paste(sqrt(n)))) +
   scale_x_log10(
-    limits = c(.0025, .1),
+    limits = c(.01, .1),
     labels = scales::percent
   ) +
   scale_y_sqrt(
@@ -68,11 +71,14 @@ fig_separation_density <-
   ) +
   ylab("Density") +
   scale_x_log10(
-    limits = c(.0025, .1),
+    name = "Predicted probability",
+    limits = c(.01, .1),
     labels = scales::percent
   ) +
   expand_limits(x = 0) +
-  facet_wrap(~ time, scales = "free_y")
+  facet_wrap(~ time, scales = "free_y") +
+  scale_fill_discrete(labels = c("no PJI", "PJI")) +
+  scale_color_discrete(labels = c("no PJI", "PJI"))
 
 
 # Combine plots -----------------------------------------------------------

### 2020-06-03 src/tab_coefs.R

@@ -6,7 +6,7 @@ baseline <-
   crossing(
     time = factor(c("90 days", "2 years")),
     term = c(
-      paste("Sex:", levels(df$P_Sex)[1]),
+      #paste("Sex:", levels(df$P_Sex)[1]),
       paste("BMI:", levels(df$P_BMI)[1]),
       paste("Diagnosis:", levels(df$P_DiaGrp)[1])
     )

### 2020-06-04 reports/article.Rmd
version sent to NH and ABP

@@
   data_auc_ci %>% 
-  filter(time == Time, Model == model) %>% 
+  filter(Period == Time, Model == model) %>% 
   select(ends_with(version)) %>% 
@@
 Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age was normalized prior to modeling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
-In order to account for the potentially U-shaped risk of developing PJI associated with either low or high BMI, this variable was treated as a categorical variable divided into four BMI intervals based on the six intervals propsed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
+BMI divided into four intervals based on the six intervals proposed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
 
-For each outcome (PJI within 90 days or 2 years), 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] Ten-fold cross validation was performed for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model, keeping only $\lambda$:s that minimized the mean cross-validated deviances in each sample and using these $\lambda$:s to estimate model coefficients for each potential predictor.
-Those coefficients were ranked based on their absolute values as a measure of variable importance.
-Piece-wise linear regression was applied to find a break-point, and influential predictors with absolute coefficients value above this pointm, scored a point.
-The whole process was repeated 100 times.
-Potential predictors that were selected at least once were used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
-Even more parsimonious models with potential predictors chosen at least 33 out of 100 times were evaluated for comparison.
+For each outcome (PJI within 90 days or 2 years), ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model.
+We kept the $\lambda_i$ that minimized the mean cross-validated deviance.
+100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
+
+Logistic LASSO regression was performed for each sample using the chosen $\lambda_i$ as penalty term.
+Regression coefficients were ranked based on their absolute values as a measure of variable importance.
+We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model.
+The second change point was then used as a break-point.
+Influential predictors with absolute coefficients value above this point, scored a point.
+The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total).
+Potential predictors that were selected at least once out of 100 were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
+Even more parsimonious models with potential predictors chosen at least 50 out of 100 times were evaluated for comparison.
 Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V co-morbidity indices were fitted for comparison.
@@
 
-Patients with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potentisl predictors prioir to further modeling.
+Patients with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potential predictors prior to further modeling.
 ROC-curves (Fig.
@@
 For simplicity, we therefore focus on the reduced models.
-Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilites for such event (Fig.
+Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for such event (Fig.
 \@ref(fig:sep)).
@@
 
-```{r tab1}
+```{r tab1, warning=FALSE}
 source("../lib/clean_names.R")

### 2020-06-04 munge/05-BRLasso.R

@@ -30,3 +30,49 @@ infection_data <-
   )
 
 cache("infection_data")
+
+
+
+
+
+
+
+
+
+
+# Find the second change point for piecewise linear regression with three segments
+# instead of
+cp2 <- function(df) {
+  df$row <- df$.row
+  model  <- list(impor ~ 1 + row, ~ 0 + row, ~ 0 + row)
+  fit    <- suppressMessages(mcp(model, df))
+  f      <- suppressWarnings(fixef(fit))
+  f[f$name == "cp_2", "mean"]
+}
+
+future::plan(multiprocess, workers = 7)
+
+# Replace old break points with new --------------------------------------------
+infection_data$best_coefs_tmp[[1]]$breaks <-
+  infection_data$best_coefs_tmp[[1]]$coefs_all %>%
+  furrr::future_map_dbl(cp2, .progress = TRUE)
+
+infection_data$best_coefs_tmp[[2]]$breaks <-
+  infection_data$best_coefs_tmp[[2]]$coefs_all %>%
+  furrr::future_map_dbl(cp2, .progress = TRUE)
+
+
+# Include variables based on new break points
+infection_data$best_coefs_tmp[[1]]$coefs <-
+  with(
+    infection_data$best_coefs_tmp[[1]],
+    map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
+  )
+
+infection_data$best_coefs_tmp[[2]]$coefs <-
+  with(
+    infection_data$best_coefs_tmp[[2]],
+    map2(coefs_all, breaks, ~slice(.x, seq_len(.y)))
+  )
+
+cache("infection_data")

### 2020-06-04 munge/06-variables_differnet_BRLasso_models.R

@@ -10,7 +10,7 @@ brlasso_coefs_selected <- function(best_coefs_tmp) {
   # List of variables selected every time
   best_coefs_reduced <-
     brlasso_tbl_selected %>%
-    filter(n >= round(1 / 3 * config$Bmax)) %>% # Urspr föreslogs intersect (variabler som tas varje gång)
+    filter(n >= round(1 / 2 * config$Bmax)) %>% # Urspr föreslogs intersect (variabler som tas varje gång)
     select(variable) %>%
     pluck(1)
 

### 2020-06-04 munge/07-compare_models.R

@@ -11,7 +11,8 @@ form <- function(nms) {
   paste("outcome ~", paste(unique(gsub(paste0(
     "_TRUE.|_X[23]|_(I|II|III|Man|Kvinna|Male|Female|married|single|widow.widower|",
     "under.normal.weight|overweight|class.I.obesity|class.II.III.obesity|",
-    "Avascular.necrosis.of.the.femoral.head..AVN.|Inflammatory.joint.disease|Secondary.osteoarthritis)"), "",
+    "Avascular.necrosis.of.the.femoral.head..AVN.|Inflammatory.joint.disease|",
+    "Secondary.osteoarthritis|Sequelae.after.childhood.hip.disease)"), "",
     nms)), collapse = " + ")
   )
 }

### 2020-06-04 src/examples.R

@@ -1,13 +1,22 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+# Find coefs for 90-day model
+# writeLines(paste(names(coef(fit_brl$fit[[1]])), collapse = " = c(, ),\n"))
+
 examples <-
   tibble(
     c_psoriasis      = c(FALSE,                     TRUE),
     P_BMI            = c("under/normal weight",     "overweight"),
     P_DiaGrp         = c("Primary osteoarthritis",  "Secondary osteoarthritis"),
     P_Sex            = c("Female",                  "Male"),
-    c_cns_disease    = c(FALSE,                     FALSE),
+    c_cns_disease    = c(FALSE,                     TRUE),
     c_liver_disease  = c(FALSE,                     FALSE),
+
+    P_ASA                      = c("I",     "III"),
+    c_pancreatic_insufficiency = c(FALSE, FALSE),
+    c_drug_alcohol_abuse       = c(FALSE, FALSE),
+    c_rheumatic_disease        = c(FALSE, FALSE),
+    c_cancer                   = c(FALSE, FALSE)
   )
 
 p <- predict(fit_brl$fit[[1]], newdata = examples, type = "response")
@@ -19,8 +28,8 @@ examples <-
   mutate(
     p = sprintf("%.1f %%", p * 100),
     desc = c(
-      "a female of normal weigth with primary osteoarthritis and no co-morbidities",
-      "a male with overweight, secondary osteoarthritis and proriasis"),
+      "a female with normal BMI, primary osteoarthritis and no co-morbidities",
+      "a male with overweight, secondary osteoarthritis, proriasis, dementia (CNS disease) and ASA class III"),
     desc = sprintf("%s would have a probability of %s", desc, p)
   )
 

### 2020-06-04 src/tab_coefs.R

@@ -6,7 +6,8 @@ baseline <-
   crossing(
     time = factor(c("90 days", "2 years")),
     term = c(
-      #paste("Sex:", levels(df$P_Sex)[1]),
+      paste("Sex:", levels(df$P_Sex)[1]),
+      paste("ASA class:", levels(df$P_ASA)[1]),
       paste("BMI:", levels(df$P_BMI)[1]),
       paste("Diagnosis:", levels(df$P_DiaGrp)[1])
     )
@@ -72,6 +73,7 @@ coefs_text <-
       variable == "bmi" ~ "body mass index",
       variable == "diagnosis" ~ "the underlaying diagnosis for THA",
       variable == "sex" ~ "gender",
+      variable == "asa class" ~ "ASA class",
       variable == "cns disease" ~ "CNS disease",
       TRUE ~ variable
     )

### 2020-09-03 src/tab_coefs.R
include validation template

@@ -75,6 +75,8 @@ coefs_text <-
       variable == "sex" ~ "gender",
       variable == "asa class" ~ "ASA class",
       variable == "cns disease" ~ "CNS disease",
+      variable == "pancreatiinsufficiency" ~ "pancreatic insufficiency",
+      variable == "rheumatidisease" ~ "rheumatic disease",
       TRUE ~ variable
     )
   ) %>%

### 2020-09-03 src/export_for_validation.R
include validation template

@@ -1,0 +1,24 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+# Hitta alla koefficientnamn som krävs för både 90 dagar och 2 år
+model_vars <-
+  fit_brl %>%
+  mutate(
+    vars = map(fit, ~names(coef(.)))
+  ) %>%
+  select(vars) %>%
+  pluck(1) %>%
+  c(recursive = TRUE) %>%
+  unique() %>%
+  {.[-1]} %>%
+  {gsub("TRUE|I{2,3}|Male", "", .)} %>%
+  {gsub("(P_BMI)(.*)", "\\1", ., perl = TRUE)} %>%
+  {gsub("(P_DiaGrp)(.*)", "\\1", ., perl = TRUE)} %>%
+  unique()
+
+# Export data frame of data needed for external validation
+ext_val_required_data <-
+  df %>%
+  select(one_of(model_vars))
+
+cache("ext_val_required_data")

### 2020-09-03 diagnostics/external_validation_template.Rmd
include validation template

---
title: "External validation"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
bibliography: 
  C:\\Users\\erik_\\Documents\\library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)

load("../cache/ext_val_required_data.RData")
load("../cache/tab_categorization.RData")
load("../cache/df.RData")
```

# Start

Install and attatch some useful packages

```{r}
# First, set random seed for reproducability:
set.seed(123)

# USeful packages
pkgs <- c("tidyverse", "doParallel", "pROC", "rms", "givitiR")

# Install if not already installed
pkgsinst <- setdiff(pkgs, rownames(installed.packages()))
if (length(pkgsinst)) install.packages(pkgsinst)

# Attatch
purrr::walk(pkgs, ~suppressPackageStartupMessages(library(., character.only = TRUE)))
```

Load the exported models to validate (the R object previously sent, probably with another path). This is a nested data frame with one row for each prediction period (90 days and 2 years). Column `fit` contains `glm` object stripped from any personal patient data.

```{r}
# Load the exported model object
load("../cache/fit_brl_reduced_lean.RData")
fit_brl_reduced_lean
```

Let's inspect the model for 90 days just to get a sence of it:

```{r}
fit_brl_reduced_lean$fit[[1]]
```

We should now use this model with the `predict` function combined with new data from Denmark. So, how should this data look like?

# Prepare data

## Inclusion/exlusion

Those were the inclusions/exclusions from Sweden. Similar criteria should apply also for the external validation data set in to get comparable cohorts.

Ignore filtering of missing educational level, hospital type and fixation, however, since those variables are not needed in the model to validate.

```{r}
knitr::include_graphics("../graphs/flowchart.png")
```

# Variables

## Outcome

The outcome variables are simply called `outcome` in each model (hence, adjusted for each model). Those are logical/boolean (or 0/1-numeric) indicating whether the patient got PJI within 90 days (2 years) after THA (`TRUE`/1) or not (`FALSE`/0). Note that we had previously excluded all patients who died within two years. A competing risk model for survival analysis might be better but we were pragmatic in this case.

## Predictors

The data to evaluate (in addition to the respective `outcome` variable) should look like this:

```{r}
# Object I use on my end, which includes patient data and can therefore not be shared
head(ext_val_required_data) 
```

thus with columns:

```{r}
names(ext_val_required_data)
```
Some of those are factor variables:
```{r}
ext_val_required_data %>% 
  select(is.factor) %>% 
  pivot_longer(everything(), values_ptypes = list(value = character())) %>% 
  distinct() %>% 
  arrange(name, value) %>% 
  group_by(name) %>% 
  mutate(name = replace(name, duplicated(name), ""))
```
- `P_Sex` and `P_ASA` should be self-explanatory
- `P_BMI` is a broader categorization based on BMI and the [WHO classification] (https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi) where overweight = "pre-obesity"  
- `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories as found in a separate CSV-file:

```{r}
readr::read_csv2("../data/P_DiaGrp.csv", trim_ws = TRUE)
```

## Comorbidities

Variables prefixed by `c_`:

```{r}
nms     <- names(ext_val_required_data)
comorbs <- nms[startsWith(nms, "c_")]
comorbs
```
are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register. INdividual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018. Those conditions were then further combined according to table 1 in the drafted manuscript:

```{r}
comorbs_nms <- 
  comorbs %>% 
  tolower() %>% 
  {gsub("^c_", "", .)} %>%
  {gsub("_", " ", .)}

tab_categorization %>% 
  filter(
    tolower(`Comorbidities by groups`) %in% comorbs_nms
  )
```

## Example validation for model of 90 days

We have the 90 day model from the shared R object;
```{r}
model <- fit_brl_reduced_lean$fit[[1]]
```

Let's assume we now have the `outcome` variable and a data frame `X` with the  predictors (I will use the Swedish data just as an example but this should of course be changed for the external validation).
```{r}
outcome <- df$outcome_infection_90d
X       <- ext_val_required_data
```

# Validation of model as is

```{r}
# Tibble with observed and predicted outcome
obspred <- 
  tibble(
    obs  = outcome, 
    pred = predict(model, X, type = "response")
  )

# ROC curve
ROC <- pROC::roc(obspred, "obs", "pred", direction = "<")

# Estimate CI for AUC based on bootstrapping
# Use parallel processing to speed up the process
doParallel::registerDoParallel()
AUCci <- 
  pROC::ci.auc(
    ROC, 
    method          = "bootstrap", 
    boot.n          = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel        = TRUE
  )

# Check calibration. Note that devel should actually be "internal" for this example but I use
# "external", since that's what you will use for the external validation. 
calibration <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred$obs), 
    obspred$pred, 
    devel = "external"
  )
```

## Results

For this example we had AUC:

```{r}
AUCci
plot(ROC)
```

A calibration belt plot might be illustrated as (this example might look strange due to the intentional internal/external miss-specification above):

```{r}
plot(calibration, xlim = c(0, 0.3), ylim = c(0, 0.3))
```

# Re-calibrated intercept

Method 2 from table 1 in @Steyerberg2004.

```{r}
Z <- predict(model, X, type = "response")

# Refit the intercept using Z = a + Xb from above as offset
fit2 <- glm(outcome ~ 1, offset = Z)

# Same calibration and validation as above
obspred2     <- tibble(obs  = outcome, pred = predict(fit2, type = "response"))
ROC2         <- pROC::roc(obspred2, "obs", "pred", direction = "<")
AUCci2 <- 
  pROC::ci.auc(
    ROC2, 
    method = "bootstrap", 
    boot.n = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel = TRUE
  )

calibration2 <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred2$obs), 
    obspred2$pred, 
    devel = "external"
  )
```

## Results

```{r}
AUCci2
plot(ROC2)
plot(calibration2, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

# Re-calibration of intercenpt and calibration slope

Method 3 from table 1 in @Steyerberg2004.

```{r}
fit3         <- glm(outcome ~ 1 + Z)
obspred3     <- tibble(obs  = outcome, pred = predict(fit3, type = "response"))
ROC3         <- pROC::roc(obspred3, "obs", "pred", direction = "<")

AUCci3 <- 
  pROC::ci.auc(
    ROC3, 
    method = "bootstrap", 
    boot.n = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel = TRUE
  )

calibration3 <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred3$obs), 
    obspred3$pred, 
    devel = "external"
  )
```

## Results

```{r}
AUCci3
plot(ROC3)
plot(calibration3, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

# Export data to Sweden

## ROC

If it would be OK to export coordinates for ROC-plots I would recommend this code to extract only the minimal data needed:

```{r}

roc_plot_coords <- 
  tibble(
    model = c("ias is", "re-calibrated intercept", "recalibrated"),
    data = 
      list( 
        tibble(
          specificities = ROC$specificities, 
          sensitivities = ROC$sensitivities
        ),
        tibble(
          specificities = ROC2$specificities, 
          sensitivities = ROC2$sensitivities
        ),
        tibble(
          specificities = ROC3$specificities, 
          sensitivities = ROC3$sensitivities
        )
      )
  )

```

## AUC with CI

The text output from `AUCci`, `AUCci2` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
```{r}
AUCci_print  <- capture.output(AUCci)
AUCci2_print <- capture.output(AUCci2)
AUCci3_print <- capture.output(AUCci3)
```

## Export objects

Save objects above to file `export_90d.RData` (in the current working directory).

```{r}
save(
  roc_plot_coords,
  AUCci_print,
  AUCci2_print,
  AUCci3_print,
  calibration,
  calibration2,
  calibration3,
  file = "export_90d.RData"
)
```

# 2 years

Repeat for the 2-year model.

# Bibliography

### 2020-09-03 diagnostics/README.Rmd
försöker med README.md

--- 2020-09-03 diagnostics/external_validation_template.Rmd
@@
   html_document: 
-    toc: true
-    toc_float: true
-    df_print: paged
+    toc: yes
+    toc_float: yes
     code_folding: show
+    keep_md: yes
+  md_document:
+    variant: markdown_github
 bibliography: 
@@
 
-Install and attatch some useful packages
+Install and attach some useful packages
 
@@
 
-Load the exported models to validate (the R object previously sent, probably with another path).
+Load the exported models to validate (also found "manually" at [Github](https://github.com/eribul/NH_luxation_infektion/blob/master/cache/ext_val_required_data.RData)).
 This is a nested data frame with one row for each prediction period (90 days and 2 years).
-Column `fit` contains `glm` object stripped from any personal patient data.
+Column `fit` contains `glm` objects stripped from any personal patient data.
 
@@
 ext_val_required_data %>% 
-  select(is.factor) %>% 
+  select(where(is.factor)) %>% 
   pivot_longer(everything(), values_ptypes = list(value = character())) %>% 

### 2020-09-03 diagnostics/README.Rmd
include figs for readme

@@
 output: 
-  html_document: 
+  github_document:
     toc: yes
-    toc_float: yes
-    code_folding: show
-    keep_md: yes
-  md_document:
-    variant: markdown_github
+    df_print: "kable"
 bibliography: 
@@
 load("../cache/fit_brl_reduced_lean.RData")
-fit_brl_reduced_lean
 ```
@@
 
-Those were the inclusions/exclusions from Sweden. Similar criteria should apply also for the external validation data set in to get comparable cohorts.
+Those were the inclusions/exclusions from Sweden. Similar criteria should apply also for the external validation data set to get comparable cohorts.
 
@@
 ```{r}
-# Object I use on my end, which includes patient data and can therefore not be shared
 head(ext_val_required_data) 

### 2020-09-03 diagnostics/README.Rmd
m

@@
 - `P_BMI` is a broader categorization based on BMI and the [WHO classification] (https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi) where overweight = "pre-obesity"
-- `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories as found in a separate CSV-file:
+- `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories.
 
@@
 ```
+...
 are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register.
-INdividual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018.
+Individual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018.
 Those conditions were then further combined according to table 1 in the drafted manuscript:
@@
 ```
-
-A calibration belt plot might be illustrated as (this example might look strange due to the intentional internal/external miss-specification above):
 

### 2020-09-29 diagnostics/README.Rmd
instructions for external validation

@@
 A competing risk model for survival analysis might be better but we were pragmatic in this case.
+
+We identified PJI within 90 days/2 years as either the primary or secondary reson for reoperation performed within this time frame, as recorded to SHAR, or if a relevant ICD-10/NOMESCO code was recorded during a hospitl visit/admission during this period.
+
+We used regular expression to identify such codes:
+
+```{r}
+coder::hip_ae_hailer %>% 
+  filter(group == "Infection")
+```
 

### 2020-09-30 reports/article.Rmd
include validation template

@@
 
+# Note!
+
+MOst text in the article is automatically generated! Bold text is not!
+It is thus possible that this text does not correspond to the rest!
+It must be updated manually!
+
 ##### PAGE BREAK
@@
 
-**PARTICIPANTS AND METHODS: ** We analysed `r nmbr(N_shar)` patients recorded in the Swedish hip arthroplasty register `r period` and obtained data on co-morbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+**PARTICIPANTS AND METHODS: ** We analysed `r nmbr(N_shar)` patients recorded in the Swedish hip arthroplasty register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery.
@@
 A model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**.
-These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` co-morbidity indices.
+These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` comorbidity indices.
 
@@
 Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
-Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of co-morbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex co-morbidity indices cannot be used in this context.
+Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
 Although high ASA class is a risk factor associated with development of PJI, the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012]
-The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] co-morbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
 The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
@@
 Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on co-morbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
+Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level  of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
 The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
-Only the last operated hip was considered for patients with bilateral THA,[@Bulow2020] and only patients who survivied at least two years aftert surgery.
+Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020].
 Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a BMI above 50.
@@
 
-## Definitions of co-morbidity and outcomes
+## Definitions of comorbidity and outcomes
 
-The Charlson, Elixhauser and Rx Risk V co-morbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of co-morbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+The Charlson, Elixhauser and Rx Risk V comorbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
 \@ref(tab:tabcategorization)).
-Some of the combined co-morbidities were observed for less than 10 patients with or without PJI.
+Some of the combined comorbidities were observed for less than 10 patients with or without PJI.
 These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
@@
 Even more parsimonious models with potential predictors chosen at least 50 out of 100 times were evaluated for comparison.
-Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V co-morbidity indices were fitted for comparison.
+Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.
 Odds ratios for the final models were estimated with 95 % confidence intervals.
@@
 The proportion of patients with ASA class III was
-`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson co-morbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V.
+`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V.
 In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig.
@@
 
-Patients with `r excl_factors` were uncommon, wherefore those co-morbidities were excluded as potential predictors prior to further modeling.
+Patients with `r excl_factors` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
 ROC-curves (Fig.
@@
 Predictors associated with a statistically significant risk increase were **secondary osteoarthritis as the underlying reason for THA surgery, male gender, obesity, and the presence of psoriasis, osteoporosis, Paget's or liver disease**.
-This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson co-morbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig.
+This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson comorbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
@@
 Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, excluding osteoporosis or Paget's disease, and additionally included an underlying diagnosis of AVN or inflammatory joint disease, and the presence of  CNS disease**.
-The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson co-morbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser co-morbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig.
+The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson comorbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
@@
 
-We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V co-morbidity indices.
+We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
 A similarly developed model to predict PJI within two years included similar predictors as the model developed to predict PJI within 90 days, but added `r coefs_text["2 years"]` to the list of predictor variables.
-This model also performed better than the established co-morbidity measures in terms of AUC and calibration.
+This model also performed better than the established comorbidity measures in terms of AUC and calibration.
 
@@
 Some previously mentioned risk factors for developing PJI were not selected during our model development.
-High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are co-morbidity in a broader sense, and, more specifically, the presence of cardiovascular co-morbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
+High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are comorbidity in a broader sense, and, more specifically, the presence of cardiovascular comorbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
 
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined co-morbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a]
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a]
 20 but this comorbidity did not reach statistical significance in our models.
@@
 
-Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the co-morbidity indices developed by Charlson or Elixhauser and the RxRisk V index alone, resulted in poor predictive power.
+Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index alone, resulted in poor predictive power.
 This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019]
@@
 ```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
-The reduced model performs no different than the main model, but both of these perform better than traditional co-morbidity measures alone."}
+The reduced model performs no different than the main model, but both of these perform better than traditional comorbidity measures alone."}
 
@@
   knitr::kable(
-    caption = "Categorization of co-morbidities from the Charlson, Elixhauser and Rx RIsk V co-morbidities. (CNS = central nervous system.)")
+    caption = "Categorization of comorbidities from the Charlson, Elixhauser and Rx RIsk V comorbidities. (CNS = central nervous system.)")
 

### 2020-10-01 reports/article.Rmd
include NHs changes
Fix  #31

@@
 ---
-title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and ..."
-author: Erik Bulow^1,2,3^, Ola Rolfson^1,2,3^, ...,  Nils P. Hailer^1,3^
+title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
+author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^
 date: '`r Sys.Date()`'
@@
 2.  Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-3. Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
+3.  Department of Clinical Medicine - Department of Clinical Epidemiology, Aarhus university, Aarhus, Denmark
+4.  XXX, Denmark
+5.  Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
 
-**Correspondance:** erik.bulow@vgregion.se; +46 70 08 234 28; Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
+**Correspondance:** [erik.bulow\@vgregion.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
 
@@
 
-MOst text in the article is automatically generated! Bold text is not!
-It is thus possible that this text does not correspond to the rest!
-It must be updated manually!
+Most text in the article is automatically generated! Bold text is not! It is thus possible that this text does not correspond to the rest! It must be updated manually!
 
@@
 
-**OJECTIVE: ** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and healthcare resources.
+**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and healthcare resources.
 Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population.
@@
 
-**PARTICIPANTS AND METHODS: ** We analysed `r nmbr(N_shar)` patients recorded in the Swedish hip arthroplasty register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+**PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery.
@@
 
-**RESULTS: ** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort.
+**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort.
 A prediction model for PJI within 90 days combined `r coefs_text["90 days"]`, resulting in an `r get_auc("90 days", "Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
@@
 
-**CONCLUSION: ** The proposed prediction models for PJI after THA are relatively parsimonious since they are based on easily accessible clinical data, and they are superior to predictions based on ASA class or more complex comorbidity indices.
-Our web-based calculator could simplify the individualized risk assessment prior to THA surgery.
+**CONCLUSION:** The proposed prediction models for PJI after THA are relatively parsimonious since they are based on easily accessible clinical data, and they are superior to predictions based on ASA class or more complex comorbidity indices.
+Our web-based calculator could simplify the individualized assessment of the risk of developing PJI prior to THA surgery and enable risk-stratified patient management.
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of all early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]
-Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
-Although high ASA class is a risk factor associated with development of PJI, the classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012]
-The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions coded by the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
-The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified by the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018]
-Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low  discriminative ability or low accuracy of the prediction models, or the absence of external model validation.
+Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
+Although high ASA class is a risk factor associated with development of PJI, this classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminative ability or low accuracy of the prediction models, or the absence of external model validation.
 Thus, there is an urgent need for improved prediction models of PJI.
@@
 Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
+Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
@@
 Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020].
-Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a BMI above 50.
+Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
 Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.
@@
 
-The Charlson, Elixhauser and Rx Risk V comorbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)` and `r nrow(coder::rxriskv)` respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+The Charlson, Elixhauser and Rx Risk V comorbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)`, and `r nrow(coder::rxriskv)` categories, respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
 \@ref(tab:tabcategorization)).
-Some of the combined comorbidities were observed for less than 10 patients with or without PJI.
-These diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
-PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in NPR within 90 days or 2 years (Tab.
-\@ref(tab:infcodes)), or if the patient was reported to SHAR, within the same time frame, as reoperated due to infection.
+Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
+PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in the NPR within 90 days or 2 years (Tab.
+\@ref(tab:infcodes)), or if the patient was reported to SHAR as re-operated due to infection within the same time frame.
 
@@
 Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age was normalized prior to modeling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
-BMI divided into four intervals based on the six intervals proposed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
+BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
 
@@
 Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples.
-The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism,
-but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
+The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
 Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**.
@@
 
-We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). 
+We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig.
+\@ref(fig:flowchart), left panel).
 `r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients developed a PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab.
 \@ref(tab:tab1)).
-The proportion of patients with ASA class III was
-`r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser, and `r comorb(df$Rx_index_pratt > 0)` according to Rx Risk V.
+The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
 In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig.
@@
 
-Patients with `r excl_factors` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
+Patients with `r tolower(excl_factors)` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
 ROC-curves (Fig.
 \@ref(fig:rocs)) and AUC-values (Fig.
-\@ref(fig:aucci)) for the main and reduced models were no different.
-For simplicity, we therefore focus on the reduced models.
-Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for such event (Fig.
+\@ref(fig:aucci)) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced models.
+Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for this outcome (Fig.
 \@ref(fig:sep)).
@@
 \@ref(tab:coefs)).
-Predictors associated with a statistically significant risk increase were **secondary osteoarthritis as the underlying reason for THA surgery, male gender, obesity, and the presence of psoriasis, osteoporosis, Paget's or liver disease**.
 This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson comorbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
-The reduced model had good accuracy as seen by the calibration plot (Fig.
+The reduced model had good accuracy as visualized by the calibration plot (Fig.
 \@ref(fig:calibration), left panel).
@@
 
-Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018]
-Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015]Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015]Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
 
@@
 
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a]
-20 but this comorbidity did not reach statistical significance in our models.
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a] 20 but this comorbidity did not reach statistical significance in our models.
 
@@
 
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare ...
+All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi\_disclosure.pdf and declare ...
 

### 2020-10-01 src/tab_coefs.R

@@ -70,10 +70,10 @@ coefs_text <-
   mutate(
     variable = tolower(variable),
     variable = case_when(
-      variable == "bmi" ~ "body mass index",
+      variable == "bmi" ~ "body mass index (BMI)",
       variable == "diagnosis" ~ "the underlaying diagnosis for THA",
       variable == "sex" ~ "gender",
-      variable == "asa class" ~ "ASA class",
+      variable == "asa class" ~ "American Society for Anesthesiologists (ASA) class",
       variable == "cns disease" ~ "CNS disease",
       variable == "pancreatiinsufficiency" ~ "pancreatic insufficiency",
       variable == "rheumatidisease" ~ "rheumatic disease",

### 2020-10-23 diagnostics/README.Rmd
Exclude RxRiskV

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
+knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
 
@@
 ```
+
 Some of those are factor variables:
+
 ```{r}
@@
 ```
+
 -   `P_Sex` and `P_ASA` should be self-explanatory
-- `P_BMI` is a broader categorization based on BMI and the [WHO classification] (https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi) where overweight = "pre-obesity"
+-   `P_BMI` is a broader categorization based on BMI and the [WHO classification] (<https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi>) where overweight = "pre-obesity"
 -   `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories.
@@
 ```
+
 ...
 are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register.
-Individual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018.
+Individual codes were grouped according to Charlson and Elixhauser as codified by table 2 in \@Quan2005, and as RxRisk V according to table 1 in \@Pratt2018.
 Those conditions were then further combined according to table 1 in the drafted manuscript:
@@
 We have the 90 day model from the shared R object;
+
 ```{r}
@@
 Let's assume we now have the `outcome` variable and a data frame `X` with the predictors (I will use the Swedish data just as an example but this should of course be changed for the external validation).
+
 ```{r}
@@
 
-Method 2 from table 1 in @Steyerberg2004.
+Method 2 from table 1 in \@Steyerbeg2004.
 
@@
 
-Method 3 from table 1 in @Steyerberg2004.
+Method 3 from table 1 in \@Steyerberg2004.
 
@@
 The text output from `AUCci`, `AUCci2` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
+
 ```{r}

### 2020-10-23 munge/01-outcome.R

@@ -18,12 +18,14 @@ df_icd10 <-
   select(LopNr, code, hospital) %>%
   as.data.table(key = c("LopNr", "code", "hospital"))
 
-df_shpr <- as.data.table(df_shpr, key = c("LopNr", "P_Side", "P_SurgDate"))
+df_shpr <-
+  df_shpr %>%
+  mutate(LopNr = as.character(LopNr)) %>%
+  as.data.table(key = c("LopNr", "P_Side", "P_SurgDate"))
 
 ae_npr <- function(df_npr, regex, endday) {
   df_shpr %>%
     select(LopNr, P_SurgDate) %>%
-    mutate(LopNr = as.character(LopNr)) %>%
     categorize(
       codedata = df_npr,
       cc = "hip_ae_hailer",
@@ -42,10 +44,10 @@ ae_npr <- function(df_npr, regex, endday) {
     )
 }
 
-outcome_kva_90d <- ae_npr(df_kva,   "regex_kva",   90)
-outcome_icd_90d <- ae_npr(df_icd10, "regex_icd10", 90)
-outcome_kva_2y  <- ae_npr(df_kva,   "regex_kva",   2 * 365)
-outcome_icd_2y  <- ae_npr(df_icd10, "regex_icd10", 2 * 365)
+outcome_kva_90d <- ae_npr(df_kva,   "kva",   90)
+outcome_icd_90d <- ae_npr(df_icd10, "icd10", 90)
+outcome_kva_2y  <- ae_npr(df_kva,   "kva",   2 * 365)
+outcome_icd_2y  <- ae_npr(df_icd10, "icd10", 2 * 365)
 
 # Outcome from SHAR -------------------------------------------------------
 outcome_reop <- function(endday) {
@@ -76,21 +78,21 @@ df_outcome <- function(outcome_icd10, outcome_kva, outcome_reop) {
     left_join(as_tibble(outcome_kva), by = c("LopNr", "P_SurgDate")) %>%
     left_join(outcome_reop,           by = c("LopNr", "P_SurgDate")) %>%
     mutate(across(
-      c(roep_infektion, reop_luxation, starts_with("hip_ae_hailer_regex")),
+      c(roep_infektion, reop_luxation, starts_with("hip_ae_hailer")),
       coalesce, FALSE
       )
     ) %>%
     transmute(
-      LopNr = as.numeric(LopNr),
+      LopNr,
       P_SurgDate,
       outcome_infection =
[…]

### 2020-10-23 munge/03-compositvariabler.R

@@ -12,6 +12,7 @@ df <- rename(df, CCI_malignancy = CCI_malingnancy)
 categorization <-
   categorization.Blad1 %>%
   mutate_all(zoo::na.locf) %>%
+  filter(from != "Rx") %>%                        # NOTE!!!!!!!!!!!! TO TRY WITHOUT
   mutate_at(vars(new, old), ~ gsub("/| ", "_", tolower(.))) %>%
   mutate(
     new = paste0("c_", new),

### 2020-11-30 diagnostics/README.Rmd

@@
 output: 
-  github_document:
+#  github_document:
+#    toc: yes
+#    df_print: "kable"
+  html_document:
     toc: yes
     df_print: "kable"
+    toc_float: yes
+    number_sections: yes
+    code_folding: "hide"
+    keep_md: yes
 bibliography: 

### 2020-12-01 diagnostics/README.Rmd
m

--- 2020-09-29 diagnostics/README.Rmd
@@
 bibliography: 
-  C:\\Users\\erik_\\Documents\\library.bib
+  # C:\\Users\\erik_\\Documents\\library.bib
+  H:\\zotero_lib.bib
 ---

### 2020-12-01 diagnostics/external_validation_template.Rmd
m

@@
     code_folding: show
-bibliography: 
-  C:\\Users\\erik_\\Documents\\library.bib
 ---
@@
 knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
-
-load("../cache/ext_val_required_data.RData")
-load("../cache/tab_categorization.RData")
-load("../cache/df.RData")
 ```
@@
 ```{r}
-# First, set random seed for reproducability:
-set.seed(123)
-
-# USeful packages
 pkgs <- c("tidyverse", "doParallel", "pROC", "rms", "givitiR")
 
-# Install if not already installed
+# Install
 pkgsinst <- setdiff(pkgs, rownames(installed.packages()))
@@
 
-Load the exported models to validate (the R object previously sent, probably with another path).
-This is a nested data frame with one row for each prediction period (90 days and 2 years).
-Column `fit` contains `glm` object stripped from any personal patient data.
+Load the exported model to validate (the R object previsously sent, probably with another path).
 
@@
 # Load the exported model object
-load("../cache/fit_brl_reduced_lean.RData")
-fit_brl_reduced_lean
+load("../cache/fit_export.RData")
 ```
 
-Let's inspect the model for 90 days just to get a sence of it:
+Set random seed for reproducability:
 
 ```{r}
-fit_brl_reduced_lean$fit[[1]]
+set.seed(123)
 ```
-
-We should now use this model with the `predict` function combined with new data from Denmark. So, how should this data look like?
 
@@
 
-Those were the inclusions/exclusions from Sweden. Similar criteria should apply also for the external validation data set in to get comparable cohorts.
+Those were the inclusions/exklusions from Sweden. It might not be necessary to filter out on BMI, hospital and education however. Those variables are not used in the model.
 
-Ignore filtering of missing educational level, hospital type and fixation, however, since those variables are not needed in the model to validate.
+Additional filter to ages 35-99 years to match the Swedish cohort.
 
@@
 
-The outcome variables are simply called `outcome` in each model (hence, adjusted for each model).
-Those are logical/boolean (or 0/1-numeric) indicating whether the patient got PJI within 90 days (2 years) after THA (`TRUE`/1) or not (`FALSE`/0).
-Note that we had previously excluded all patients who died within two years.
-A competing risk model for survival analysis might be better but we were pragmatic in this case.
+The outcome variable is boolean (or numeric) indicating wether the patient died of any cause within 90 days after THA (`TRUE`/1) or not (`FALSE`/0).
+We did not have any censoring in Sweden.
+I guess we can simply drop cases where status is unknown?
 
@@
 
-The data to evaluate (in addition to the respective `outcome` variable) should look like this:
+The data to evaluate should look like this:
 
 ```{r}
-# Object I use on my end, which includes patient data and can therefore not be shared
-head(ext_val_required_data) 
+head(fit_export$data)
 ```
 
-thus with columns:
+## Baseline variables
 
-```{r}
-names(ext_val_required_data)
-```
-Some of those are factor variables:
-```{r}
-ext_val_required_data %>% 
-  select(is.factor) %>% 
-  pivot_longer(everything(), values_ptypes = list(value = character())) %>% 
-  distinct() %>% 
-  arrange(name, value) %>% 
-  group_by(name) %>% 
-  mutate(name = replace(name, duplicated(name), ""))
-```
-- `P_Sex` and `P_ASA` should be self-explanatory
-- `P_BMI` is a broader categorization based on BMI and the [WHO classification] (https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi) where overweight = "pre-obesity"
-- `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories as found in a separate CSV-file:
-
-```{r}
-readr::read_csv2("../data/P_DiaGrp.csv", trim_ws = TRUE)
-```
+- **P_Gender**: Kvinna/Man = Female/Male
+- **P_ASA**: level 1-3
+- **P_Age**: 35 - 99
 
@@
 
-Variables prefixed by `c_`:
+Comorbidities are identified by Elixhauser and Charlson comorbidity during one year before surgery.
+This is done by ICD-10 codes in Sweden but other versions exist.
+If ICD-10-codes are used those might be identified by regular expressions in the table below.
+Those are based on codes like "C123".
+Hence no puntctuation (not "C12.3").
+The regular expressions might be modified to allow possible other patterns.
+Regular expressoins below are just combined from differnet groups (separated by "|").
+They could be rewritten for clarity.
+If you don't see any regular expressions, click the small arrow in the upper right corner of the table.
+The `grepl` function (base R) can be used to identify ICD-10-codes based on the regular expressions.
 
-```{r}
-nms     <- names(ext_val_required_data)
-comorbs <- nms[startsWith(nms, "c_")]
-comorbs
-```
-are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register.
-INdividual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018.
-Those conditions were then further combined according to table 1 in the drafted manuscript:
+```{r, echo = FALSE}
+load("../cache/categorization.RData")
+comorbidities <- 
+  categorization %>% 
+  map_df(as.character) %>% 
+  filter(new %in% names(fit_export$data)[grepl("c_", names(fit_export$data))]) %>% 
+  separate(old, letters, sep = "\\|", fill = "right") %>% 
+  pivot_longer(letters, values_drop_na = TRUE) %>% 
+  mutate(value = trimws(value)) %>% 
+  separate(value, c("index", "name"), extra = "merge") %>% 
+  mutate(name = gsub("_", " ", name))
 
-```{r}
-comorbs_nms <- 
-  comorbs %>% 
-  tolower() %>% 
-  {gsub("^c_", "", .)} %>%
-  {gsub("_", " ", .)}
+CCI <- 
+  comorbidities %>% 
+  filter(index == "CCI") %>% 
+  left_join(coder::charlson_icd10, c(name = "group"))
 
-tab_categorization %>% 
-  filter(
-    tolower(`Comorbidities by groups`) %in% comorbs_nms
-  )
+ECI <- 
+  comorbidities %>% 
+  filter(index == "ECI") %>% 
+  left_join(coder::elix_icd10, c(name = "group"))
+
+comorb_defs <- 
+  bind_rows(CCI, ECI) %>% 
+  select(new, regex) %>% 
+  mutate(regex = sprintf("(%s)", gsub("^", "", regex, fixed = TRUE))) %>% 
+  group_by(new) %>% 
+  summarise(regex = paste(regex, collapse = "|"))
+
+comorb_defs
 ```
 
-## Example validation for model of 90 days
+## Example data
 
-We have the 90 day model from the shared R object;
+Let's assume we now have some data (I will use the Swedish data just as an example).
 ```{r}
-model <- fit_brl_reduced_lean$fit[[1]]
-```
-
-Let's assume we now have the `outcome` variable and a data frame `X` with the  predictors (I will use the Swedish data just as an example but this should of course be changed for the external validation).
-```{r}
-outcome <- df$outcome_infection_90d
-X       <- ext_val_required_data
+y <- fit_export$y
+X <- fit_export$data
 ```
@@
   tibble(
-    obs  = outcome, 
-    pred = predict(model, X, type = "response")
+    obs  = y, 
+    pred = predict(fit_export, X, type = "response")
   )
@@
     method          = "bootstrap", 
-    boot.n          = 100,     # Remove this argument and use the default!
     boot.stratified = FALSE, 
@@
 # Check calibration. Note that devel should actually be "internal" for this example but I use
-# "external", since that's what you will use for the external validation. 
+# "external", since that's what you will use for the UK data. 
 calibration <- 
   givitiR::givitiCalibrationBelt(
-    as.numeric(obspred$obs), 
+    obspred$obs, 
     obspred$pred, 
@@
 
-A calibration belt plot might be illustrated as (this example might look strange due to the intentional internal/external miss-specification above):
+A calibration belt plot might be illustrated as:
 
 ```{r}
-plot(calibration, xlim = c(0, 0.3), ylim = c(0, 0.3))
+plot(calibration, xlim = c(0, 0.03), ylim = c(0, 0.06))
 ```
@@
 
-Method 2 from table 1 in @Steyerberg2004.
+Method 2 from table 1 in Steyerberg 2004.
 
 ```{r}
-Z <- predict(model, X, type = "response")
+Z <- predict(fit_export, X, type = "response")
 
 # Refit the intercept using Z = a + Xb from above as offset
-fit2 <- glm(outcome ~ 1, offset = Z)
+fit2 <- glm(y ~ 1, offset = Z)
 
 # Same calibration and validation as above
-obspred2     <- tibble(obs  = outcome, pred = predict(fit2, type = "response"))
+obspred2     <- tibble(obs  = y, pred = predict(fit2, type = "response"))
 ROC2         <- pROC::roc(obspred2, "obs", "pred", direction = "<")
-AUCci2 <- 
-  pROC::ci.auc(
-    ROC2, 
-    method = "bootstrap", 
-    boot.n = 100,     # Remove this argument and use the default!
-    boot.stratified = FALSE, 
-    parallel = TRUE
-  )
-
-calibration2 <- 
-  givitiR::givitiCalibrationBelt(
-    as.numeric(obspred2$obs), 
-    obspred2$pred, 
-    devel = "external"
-  )
+AUCci2       <- pROC::ci.auc(ROC2, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
+calibration2 <- givitiR::givitiCalibrationBelt(obspred2$obs, obspred2$pred, devel = "external")
 ```
@@
 
-Method 3 from table 1 in @Steyerberg2004.
+Method 3 from table 1 in Steyerberg 2004.
 
 ```{r}
-fit3         <- glm(outcome ~ 1 + Z)
-obspred3     <- tibble(obs  = outcome, pred = predict(fit3, type = "response"))
+fit3         <- glm(y ~ 1 + Z)
+obspred3     <- tibble(obs  = y, pred = predict(fit3, type = "response"))
 ROC3         <- pROC::roc(obspred3, "obs", "pred", direction = "<")
-
-AUCci3 <- 
-  pROC::ci.auc(
-    ROC3, 
-    method = "bootstrap", 
-    boot.n = 100,     # Remove this argument and use the default!
-    boot.stratified = FALSE, 
-    parallel = TRUE
-  )
-
-calibration3 <- 
-  givitiR::givitiCalibrationBelt(
-    as.numeric(obspred3$obs), 
-    obspred3$pred, 
-    devel = "external"
-  )
+AUCci3       <- pROC::ci.auc(ROC3, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
+calibration3 <- givitiR::givitiCalibrationBelt(obspred3$obs, obspred3$pred, devel = "external")
 ```
@@
 roc_plot_coords <- 
-  tibble(
-    model = c("ias is", "re-calibrated intercept", "recalibrated"),
-    data = 
-      list( 
-        tibble(
+  data.frame(
     specificities = ROC$specificities, 
     sensitivities = ROC$sensitivities
-        ),
-        tibble(
-          specificities = ROC2$specificities, 
-          sensitivities = ROC2$sensitivities
-        ),
-        tibble(
+  )
+
+roc3_plot_coords <- 
+  data.frame(
     specificities = ROC3$specificities, 
     sensitivities = ROC3$sensitivities
-        )
-      )
   )
@@
 
-The text output from `AUCci`, `AUCci2` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
+The text output from `AUCci` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
 ```{r}
 AUCci_print  <- capture.output(AUCci)
-AUCci2_print <- capture.output(AUCci2)
 AUCci3_print <- capture.output(AUCci3)
@@
 
-Save objects above to file `export_90d.RData` (in the current working directory).
+Save objects above to file `export.RData` (in the current working directory).
 
@@
   roc_plot_coords,
+  roc3_plot_coords,
   AUCci_print,
-  AUCci2_print,
   AUCci3_print,
-  calibration,
-  calibration2,
-  calibration3,
-  file = "export_90d.RData"
+  file = "export.RData"
 )
@@
 
-# 2 years
+## Calibration plots
 
-Repeat for the 2-year model.
+### Help function
 
-# Bibliography
+This is a simple help function to make a clean calibration belt plot and save it as TIFF (in the curent working directory):
+
+```{r}
+makeplot <- function(x, file_name = deparse(substitute(x))) {
+  tiff(
+    paste0(file_name, ".tiff"), 
+    1024, 1024, pointsize = 36, 
+    compression = "lzw"
+  )
+
+  tcks <- seq(.0, .1, .01)
+
+  plot(
+    x,
+    xlim             = c(0, .06),
+    ylim             = c(0, .08),
+    xlab             = "Predicted probabilities [%]",
+    ylab             = "Observed  probabilities [%]",
+    main             = NULL,
+    table            = FALSE,
+    polynomialString = FALSE,
+    pvalueString     = FALSE,
+    nString          = FALSE,
+    mar              = c(5, 4, 0, 0) + 0.1,
+    xaxt             = "n",
+    yaxt             = "n"
+  )
+  abline(v = .03, lty = "dashed", col = "darkgreen", lwd = 3)
+  axis(1, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)
+  axis(2, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)
+
+  dev.off()
+}
+```
+### Make and save figures
+
+```{r}
+makeplot(calibration)
+makeplot(calibration3)
+```
+
+## ROC plots
+
+If it not possible to export data for ROC-plot, here is some code to make a figure and save it as TIFF (in the current working directory) instead:
+
+```{r}
+
+roc_plot_coords %>% 
+  ggplot(aes(1 - specificities, sensitivities)) +
+  geom_path(size = 2) +
+  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
+  theme_minimal() +
+  theme(
+    legend.position = c(1, 0),
+    legend.justification = c(1, 0),
+    legend.title = element_blank()
+  )
+
+ggsave(
+  "roc.tiff", 
+  height = 10, 
+  width = 10, 
+  unit = "cm", 
+  dpi = 900, 
+  compression = "lzw"
+)
+```

### 2020-12-01 src/fit_export.R

@@ -1,0 +1,17 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+
+fit_export <- fit_brl_reduced$glm
+
+fit_export$data <-
+  select(
+    fit_export$data,
+    one_of(gsub("TRUE|[23]|Man", "",names(coef(fit_export))[-1]))
+  ) # %>%
+# head()
+
+cache("fit_export")
+
+
+options(digits = 22, pillar.sigfig = 22)
+enframe(coef(fit_export))

### 2020-12-14 diagnostics/README.Rmd

--- 2020-10-23 diagnostics/README.Rmd
@@
 output: 
-  github_document:
+#  github_document:
+#    toc: yes
+#    df_print: "kable"
+  html_document:
     toc: yes
     df_print: "kable"
+    toc_float: yes
+    number_sections: yes
+    code_folding: "hide"
+    keep_md: yes
 bibliography: 
-  C:\\Users\\erik_\\Documents\\library.bib
+  # C:\\Users\\erik_\\Documents\\library.bib
+  H:\\zotero_lib.bib
 ---

### 2020-12-14 reports/article.Rmd

@@
 The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total).
-Potential predictors that were selected at least once out of 100 were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
-Even more parsimonious models with potential predictors chosen at least 50 out of 100 times were evaluated for comparison.
+Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
+Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
 Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.

### 2021-01-19 diagnostics/README.Rmd
m

--- 2020-11-30 diagnostics/README.Rmd
@@
 -   `P_Sex` and `P_ASA` should be self-explanatory
--   `P_BMI` is a broader categorization based on BMI and the [WHO classification] (<https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi>) where overweight = "pre-obesity"
+-   `P_BMI` is a broader categorization based on BMI and the [WHO classification] (<https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi>) where overweight = "pre-obesity"\
 -   `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories.

### 2021-01-19 reports/article.Rmd
Include T846F in inclusion list
Fix #37 OBS! Denna output hör ihop med cache_T846F som inte versionshanteras!

@@
   bookdown::word_document2:
+    number_sections: false
     reference_docx: article.dotx
@@
 Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
-Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
+Even more parsimonious models with potential predictors chosen more than `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
 Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.

### 2021-01-19 munge/03-compositvariabler.R

@@ -1,5 +1,8 @@
 # Förbered kompisitvariabler som ev kan ersätta enskilda diagnoser
 
+if (!exists("categorization.Blad1"))
+  categorization.Blad1 <- readxl::read_excel("data/categorization.xlsx")
+
 # Fulfix!
 # Tidigare Fel i coder-paketet döpte "malignancy" till "malingnancy".
 # Nu åtgärdat i paketet men fortf fel i samkörningsdatabsen

### 2021-01-19 munge/06-variables_differnet_BRLasso_models.R

@@ -14,7 +14,7 @@ brlasso_coefs_selected <- function(best_coefs_tmp) {
   # List of variables selected every time
   best_coefs_reduced <-
     brlasso_tbl_selected %>%
-    filter(n >= brlasso_selection_n["reduced"]) %>% # Urspr föreslogs intersect (variabler som tas varje gång) OBS!!!!!!!!!
+    filter(n > brlasso_selection_n["reduced"]) %>% # Urspr föreslogs intersect (variabler som tas varje gång) OBS!!!!!!!!!
     select(variable) %>%
     pluck(1)
 

### 2021-01-19 munge/07-compare_models.R

@@ -12,7 +12,8 @@ form <- function(nms) {
     "_TRUE.|_X[23]|_(I|II|III|Man|Kvinna|Male|Female|married|single|widow.widower|",
     "under.normal.weight|overweight|class.I.obesity|class.II.III.obesity|",
     "Avascular.necrosis.of.the.femoral.head..AVN.|Inflammatory.joint.disease|",
-    "Secondary.osteoarthritis|Sequelae.after.childhood.hip.disease)"), "",
+    "Secondary.osteoarthritis|Sequelae.after.childhood.hip.disease|",
+    "County)"), "",
     nms)), collapse = " + ")
   )
 }
@@ -56,7 +57,7 @@ compare_models <- function(brlasso_coefs, outcome) {
                       method = "bootstrap",
                       boot.stratified = FALSE,
                       .progress = TRUE,
-                      .options = future_options(seed = TRUE)
+                      .options = furrr_options(seed = TRUE)
                     ),
     AUC_lo        = map_dbl(AUCci, 1),
     AUC_est       = map_dbl(AUCci, 2),

### 2021-01-19 src/examples.R

@@ -3,6 +3,8 @@ suppressMessages({library(ProjectTemplate); load.project()})
 # Find coefs for 90-day model
 # writeLines(paste(names(coef(fit_brl$fit[[1]])), collapse = " = c(, ),\n"))
 
+load("cache/fit_brl.RData")
+
 examples <-
   tibble(
     c_psoriasis      = c(FALSE,                     TRUE),
@@ -18,7 +20,8 @@ examples <-
     c_rheumatic_disease           = c(FALSE, FALSE),
     c_cancer                      = c(FALSE, FALSE),
     c_peptic_ulcer                = c(FALSE, FALSE),
-    c_fluid_electrolyte_disorders = c(FALSE, FALSE)
+    c_fluid_electrolyte_disorders = c(FALSE, FALSE),
+    c_lung_airways_disease        = c(FALSE, FALSE)
   )
 
 p <- predict(fit_brl$fit[[1]], newdata = examples, type = "response")

### 2021-01-19 src/export_for_validation.R

@@ -1,5 +1,8 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/fit_brl.RData")
+load("cache/df.RData")
+
 # Hitta alla koefficientnamn som krävs för både 90 dagar och 2 år
 model_vars <-
   fit_brl %>%

### 2021-01-19 src/fig_auc.R

@@ -3,6 +3,7 @@ suppressMessages({library(ProjectTemplate); load.project()})
 digs <- options("digits")
 options(digits = 2)
 
+load("cache/infection_data.RData")
 
 # Prepare data ------------------------------------------------------------
 

### 2021-01-19 src/fig_calibration.R

@@ -1,5 +1,6 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/infection_data.RData")
 
 get_beltdata <- function(cb) {
   tibble(

### 2021-01-19 src/fig_roc.R

@@ -1,5 +1,6 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/infection_data.RData")
 
 # SHAR --------------------------------------------------------------------
 

### 2021-01-19 src/fig_separation.R

@@ -1,5 +1,7 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/infection_data.RData")
+
 obspred <-
   infection_data %>%
   select(outcome, time, all_models) %>%

### 2021-01-19 src/tab_coefs.R

@@ -1,5 +1,6 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/infection_data.RData")
 
 # Baseline levels for cqategorical variables -----------------------------------
 baseline <-

### 2021-01-19 src/tab_define_groups.R

@@ -1,5 +1,7 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+categorization.Blad1 <- readxl::read_excel("data/categorization.xlsx")
+
 tab_categorization <-
   categorization.Blad1 %>%
   mutate_all(zoo::na.locf) %>%

### 2021-01-19 src/tab_table1.R

@@ -1,5 +1,7 @@
 suppressMessages({library(ProjectTemplate); load.project()})
 
+load("cache/df.RData")
+
 dft1 <-
   df %>%
   mutate(

### 2021-02-03 reports/article.Rmd
updated article with only 90 days

--- 2020-12-14 reports/article.Rmd
@@
 # Results -------------------------------------------------------------
-crude_rate_90d <- sum(df$outcome_infection_90d) / nrow(df)
-crude_rate_2y  <- sum(df$outcome_infection_2y) / nrow(df)
+crude_rate <- sum(df$outcome) / nrow(df)
 
-get_auc <- function(Time, model, version = "1") {
+get_auc <- function(model, version = "1") {
   data_auc_ci %>% 
-  filter(Period == Time, Model == model) %>% 
+  filter(Model == model) %>% 
   select(ends_with(version)) %>% 
@@
 **PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
-Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery.
+Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery.
 Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**.
@@
 
-**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort.
-A prediction model for PJI within 90 days combined `r coefs_text["90 days"]`, resulting in an `r get_auc("90 days", "Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
-A model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**.
-These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` comorbidity indices.
+**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort.
+A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
+These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("ASA class")`, Charlson `r get_auc("Charlson")`, Elixhauser `r get_auc("Elixhauser")`, and the Rx Risk V `r get_auc("Rx Risk V")` comorbidity indices.
 
@@
 We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
-The best models were then validated externally on a cohort from the **XXX**.
+The best models were then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).
 
@@
 
-## External validation cohort (XXX)
+## External validation cohort (Denmark)
 
@@
 
-The Charlson, Elixhauser and Rx Risk V comorbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)`, and `r nrow(coder::rxriskv)` categories, respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
 \@ref(tab:tabcategorization)).
 Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
-PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in the NPR within 90 days or 2 years (Tab.
+PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-codes recorded in the NPR within 90 days (Tab.
 \@ref(tab:infcodes)), or if the patient was reported to SHAR as re-operated due to infection within the same time frame.
@@
 
-For each outcome (PJI within 90 days or 2 years), ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model.
-We kept the $\lambda_i$ that minimized the mean cross-validated deviance.
+Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model.
+We kept the $\lambda$ that minimized the mean cross-validated deviance.
 100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
 
-Logistic LASSO regression was performed for each sample using the chosen $\lambda_i$ as penalty term.
+Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term.
 Regression coefficients were ranked based on their absolute values as a measure of variable importance.
@@
 Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.
-Odds ratios for the final models were estimated with 95 % confidence intervals.
+The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
+This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals.
 
@@
 
-Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation).
+Each developed model was used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
 Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples.
 The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**.
-Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.
+Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a Danish cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to Denmark.
 
@@
 
-We built an online web calculator available at https **XXX** to be used in clinical practice.
-R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.
+We built an online web calculator available at **XXX** to be used in clinical practice.
+`r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.
 
@@
 \@ref(fig:flowchart), left panel).
-`r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients developed a PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab.
+`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of the patients developed a PJI within 90 days (Tab.
 \@ref(tab:tab1)).
@@
 
-Patients with `r tolower(excl_factors)` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
+Patients with `r excl_factors` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
 ROC-curves (Fig.
@@
 
-The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text["90 days"]` (Tab.
+The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text` (Tab.
 \@ref(tab:coefs)).
-This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson comorbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig.
+This reduced model for the predicition of PJI within 90 days had an `r get_auc("Reduced model", "2")`, whereas ASA class `r get_auc("ASA class")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk V")` had less discriminative ability (Fig.
 \@ref(fig:rocs) and \@ref(fig:aucci)).
@@
 \@ref(fig:calibration), left panel).
-
-The reduced model for PJI within two years included the same predictors, with the addition of `r coefs_text["2 years"]` (Tab.
-\@ref(tab:coefs)).
-Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, excluding osteoporosis or Paget's disease, and additionally included an underlying diagnosis of AVN or inflammatory joint disease, and the presence of CNS disease**.
-The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson comorbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig.
-\@ref(fig:rocs) and \@ref(fig:aucci)).
-Calibration was good for the reduced model (Fig.
-\@ref(fig:calibration), right panel).
 
@@
 We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
-A similarly developed model to predict PJI within two years included similar predictors as the model developed to predict PJI within 90 days, but added `r coefs_text["2 years"]` to the list of predictor variables.
 This model also performed better than the established comorbidity measures in terms of AUC and calibration.
@@
 
-Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab.
+Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab.
 \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics.
-For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.
+For example, `r examples$desc` for PJI within 90 days.
 
@@
 The predictors in our models were chosen based on a strict analytical approach, not on proposed relevance of patient or procedure characteristics.
-The majority of the identified predictor variables are however associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.
+The majority of the identified predictor variables are, however, associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.
 
@@
 
-NPH initiated the study, OR managed the ethical review board application in Sweden, EB developed the statistical model, **XXX** performed external validation with data from **XXX**.
+NPH initiated the study.
+OR managed the ethical review board application in Sweden and ABP in Denmark.
+EB developed the statistical model.
+ITA and UH performed the external validation.
 NPH and EB drafted, and all authors edited and finalized the manuscript.
-
-# Competing interest statement
-
-All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi\_disclosure.pdf and declare ...
-
-# Transparency Statement
-
-NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.
-
-# Public and patient involvement statement
-
-There is a patient representative at SHAR to ensure the patients' interests are respected in planned research.
-
-# Data sharing statement
-
-Data is available upon reasonable request to the principal investigator.
 
@@
 ```{r brlprop}
-load("../cache/infection_data.RData")
-
-infection_data %>% 
-  select(outcome, time, coefs_selected) %>% 
-  mutate(coefs_selected = map(coefs_selected, "brlasso_tbl_selected")) %>% 
-  unnest("coefs_selected") %>%
-  filter(outcome == "infection") %>% 
-  select(-outcome) %>% 
-  pivot_wider(names_from = "time", values_from = "n") %>% 
-  mutate(
-    variable = clean_names(variable) %>% 
-      {gsub("\\.", " ", .)} %>% 
-      {gsub("Rheumatidisease", "Rheumatic disease", .)} %>% 
-      {gsub("Sex Male", "Male gender", .)}
-  ) %>% 
-  select(Predictor = variable, `90 days` = `90d`, `2 years` = `2y`) %>% 
+coefs_selected_tab %>% 
 knitr::kable(
-  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model.
-Variables chosen at least 33 times out of 100 were kept in the reduced model as well.
+  caption = "Variables selected by the bootstrap ranking procedure.
+Variables selected at least 10 out of 100 times were used in the main model.
+Variables chosen at least 80 times were kept in the reduced model as well.
 (BMI = body mass index.
@@
   mutate(
-    time = replace(as.character(time), duplicated(time), ""),
     beta = round(beta, 2)

### 2021-02-03 lib/clean_names.R

@@ -6,7 +6,12 @@ clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
   {gsub("ASA",     "ASA class: ", .)} %>%
   {gsub("BMI",     "BMI: ",       .)} %>%
   {gsub("aids",    "AIDS/HIV",    .)} %>%
-  {paste0(toupper(substr(., 1, 1)), substring(., 2))}
+  {gsub("cns",     "CNS",         .)} %>%
+  {gsub("\\.",     " ",           .)} %>%
+  {gsub("  ",      " ",           .)} %>%
+  {gsub("AVN",     "\\(AVN\\)",   .)} %>%
+  {paste0(toupper(substr(., 1, 1)), substring(., 2))} %>%
+  trimws()
 }
 
 

### 2021-02-03 src/tab_coefs_selected.R

@@ -1,0 +1,12 @@
+library(ProjectTemplate)
+load.project()
+
+load("cache/coefs_selected.RData")
+
+coefs_selected_tab <-
+  coefs_selected$brlasso_tbl_selected %>%
+  mutate(
+    variable = clean_names(variable)
+  )
+
+cache("coefs_selected_tab")

### 2021-02-05 diagnostics/README.Rmd
Updated report

--- 2020-12-14 diagnostics/README.Rmd
@@
 bibliography: 
-  # C:\\Users\\erik_\\Documents\\library.bib
-  H:\\zotero_lib.bib
+  C:\\Users\\erik_\\Documents\\library.bib
+  # H:\\zotero_lib.bib
 ---
@@
 Load the exported models to validate (also found "manually" at [Github](https://github.com/eribul/NH_luxation_infektion/blob/master/cache/ext_val_required_data.RData)).
-This is a nested data frame with one row for each prediction period (90 days and 2 years).
-Column `fit` contains `glm` objects stripped from any personal patient data.
 
@@
 # Load the exported model object
-load("../cache/fit_brl_reduced_lean.RData")
+load("../cache/model_reduced_lean.RData")
 ```
 
-Let's inspect the model for 90 days just to get a sence of it:
+Let's inspect the model for 90 days just to get a sense of it:
 
 ```{r}
-fit_brl_reduced_lean$fit[[1]]
+model_reduced_lean
 ```
@@
 
-The outcome variables are simply called `outcome` in each model (hence, adjusted for each model).
-Those are logical/boolean (or 0/1-numeric) indicating whether the patient got PJI within 90 days (2 years) after THA (`TRUE`/1) or not (`FALSE`/0).
-Note that we had previously excluded all patients who died within two years.
-A competing risk model for survival analysis might be better but we were pragmatic in this case.
+The outcome variable is simply called `outcome` . This is a logical/boolean (or 0/1-numeric) variable indicating whether the patient got PJI within 90 days after THA (`TRUE`/1) or not (`FALSE`/0).
 
-We identified PJI within 90 days/2 years as either the primary or secondary reson for reoperation performed within this time frame, as recorded to SHAR, or if a relevant ICD-10/NOMESCO code was recorded during a hospitl visit/admission during this period.
+We identified PJI within 90 days as either the primary or secondary reason for reoperation performed within this time frame, as recorded to SHAR, or if a relevant ICD-10/NOMESCO code was recorded during a hospital visit/admission during this period.
 
@@
 ...
-are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register.
-Individual codes were grouped according to Charlson and Elixhauser as codified by table 2 in \@Quan2005, and as RxRisk V according to table 1 in \@Pratt2018.
+are logical/boolean indicators of comorbidities based on ICD-10 codes from one year prior to THA, as recorded in our National Patient Register.
+Individual codes were grouped according to Charlson and Elixhauser as codified by table 2 in \@Quan2005.
 Those conditions were then further combined according to table 1 in the drafted manuscript:
@@
 ```{r}
-model <- fit_brl_reduced_lean$fit[[1]]
+model <- model_reduced_lean
 ```
@@
 ```{r}
-outcome <- df$outcome_infection_90d
+outcome <- df$outcome
 X       <- ext_val_required_data
@@
 
-# 2 years
-
-Repeat for the 2-year model.
-
 # Bibliography

### 2021-02-05 lib/clean_names.R

@@ -10,6 +10,7 @@ clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
   {gsub("\\.",     " ",           .)} %>%
   {gsub("  ",      " ",           .)} %>%
   {gsub("AVN",     "\\(AVN\\)",   .)} %>%
+  {gsub("Rheumatidisease", "Rheumatic disease",   .)} %>%
   {paste0(toupper(substr(., 1, 1)), substring(., 2))} %>%
   trimws()
 }

### 2021-02-10 reports/article.Rmd

@@
 title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
-author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^
+author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^ **Ute Hahn + ? + decide order)**
 date: '`r Sys.Date()`'
@@
     df_print: kable
+    number_sections: false
 ---
@@
 
-**Correspondance:** [erik.bulow\@vgregion.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden
+**Correspondance:** [erik.bulow\@gu.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; The Swedish Joint Arthroplasty Register, Registercentrum Vastra Gotaland, SE-413 45, Sweden
 
@@
 
-**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and healthcare resources.
+**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and health care resources.
 Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population.
-We thus developed a parsimonious model to predict PJI within 90 days and two years after elective THA.
+We thus developed a parsimonious model to predict PJI within 90 days after elective THA.
 
-**PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+**PARTICIPANTS AND METHODS:** We analyzed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery.
-Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**.
-We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.
+Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from The Danish Hip Arthroplasty Register.
+We assessed model performance by the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.
 
@@
 The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
-Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020].
-Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
+Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
 Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.
@@
 The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total).
-Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables.
-Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
+Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of numeric variables.
+An even more parsimonious model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
 Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.
@@
 We built an online web calculator available at **XXX** to be used in clinical practice.
-`r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.
+`r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder[@Bulow2020b] and shiny were used.
 
@@
 The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
-In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig.
+In addition, **XXX** patients from DHR were included as an external validation cohort (Fig.
 \@ref(fig:flowchart), right panel).
@@
 For example, `r examples$desc` for PJI within 90 days.
+Note, however, that observed proportions above 10 % were rare.
+Predicted probabilities above this limit are therefore subject to extrapolation.
 
@@
 Importantly, some previously identified risk factors for PJI were not included in our models, mostly because we lacked detailed information on laboratory findings and non-prescribed medications.
+
+Although, not a strict limitation, it should be noted that the aim of prediction is different from traditional statistical inference.
+As such, confidence intervals presented for the model coefficients (Table \@ref(tab:coefs)) were based solely on the final logistic regression model.
+Additional uncertainty introduced by the variable selection process was not incorporated.
+Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variables.
+Also, the individual odds ratios should not be interpreted as effect measures of individual relations.
+Some variables are correlated.
+It is therefore not always possible to hold all other variables constant while changing only the relevant one.
 
@@
 
-```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity tp illustrate discriminative abilities of the different models.
-The reduced model perform almost as good as the main model for prediction of PJI within 90 days after surgery, and as good for PJI within 2 years."}
+```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
+The main and reduced models performed almost identival for prediction of PJI within 90 days after surgery."}
 knitr::include_graphics("../graphs/roc.png")
@@
 ```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
-The reduced model performs no different than the main model, but both of these perform better than traditional comorbidity measures alone."}
+The reduced model performs no different than the main model, and both of these perform better than traditional comorbidity measures alone."}
 
@@
 ```{r sep, fig.cap = "The vast majority of patients did not have PJI.
-Hence, the blue bars dominates the histograms (upper panels; note the scales).
-Normalized density plots reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
+Hence, the blue bars dominates the histograms (left panel; note the scales).
+A normalized density plot reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
 knitr::include_graphics("../graphs/separation.png")

### 2021-04-08 diagnostics/README.Rmd
m

@@
 
-Load the exported models to validate (also found "manually" at [Github](https://github.com/eribul/NH_luxation_infektion/blob/master/cache/ext_val_required_data.RData)).
+Load the exported model to validate (also found "manually" at [Github](https://github.com/eribul/NH_luxation_infektion/blob/master/cache/ext_val_required_data.RData)).
 

### 2021-05-26 reports/article.Rmd
new draft after NHs comments

@@
 ---
-title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
-author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^ **Ute Hahn + ? + decide order)**
+title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
+author: Erik Bülow,^1,2^ Ute Hahn^3^, ~~Ina Trolle Andersen,^3^~~**???** Aurélie Mailhac,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3^ Nils P. Hailer^4^ **TODO DECIDE ORDER!!!**
 date: '`r Sys.Date()`'
@@
 
-1.  The Swedish Hip Arthroplasty Register, Gothenburg, Sweden
+1.  The Swedish Arthroplasty Register, Gothenburg, Sweden
 2.  Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-3.  Department of Clinical Medicine - Department of Clinical Epidemiology, Aarhus university, Aarhus, Denmark
-4.  XXX, Denmark
-5.  Department of Orthopaedics, Institute of Surgical Sciences, Uppsala University Hospital, Uppsala, Sweden
+3.  Department of Clinical Medicine, Department of Clinical Epidemiology, Aarhus university, Aarhus, Denmark
+4.  Department Surgical Sciences/Orthopaedics, Uppsala University Hospital, Uppsala, Sweden
 
-**Correspondance:** [erik.bulow\@gu.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; The Swedish Joint Arthroplasty Register, Registercentrum Vastra Gotaland, SE-413 45, Sweden
+**Correspondence:** [erik.bulow@gu.se](mailto:erik.bulow@gu.se); +46 70 08 234 28; The Swedish Arthroplasty Register, Registercentrum Västra Götaland, SE-413 45, Sweden
 
@@
   table1 %>% 
-  filter(what == "n") %>% 
-  select(Total) %>% 
+  filter(what == "Total") %>% 
+  select(`SE Total`) %>% 
+  pluck(1)
+
+N_dk <- 
+  table1 %>% 
+  filter(what == "Total") %>% 
+  select(`DK Total`) %>% 
   pluck(1)
@@
 
-get_auc <- function(model, version = "1") {
+get_auc <- function(model, version = "1", Country = "Sweden") {
   data_auc_ci %>% 
-  filter(Model == model) %>% 
+  filter(Model == model, country == Country) %>% 
   select(ends_with(version)) %>% 
@@
 
-**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and health care resources.
-Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population.
+**OJECTIVE:** Previous attempts at developing risk calculators for periprosthetic joint infection (PJI)  after total hip arthroplasty (THA) are hampered by poor predictive performance of the proposed models or a lack of external validation, and no published model is derived on a European population.
 We thus developed a parsimonious model to predict PJI within 90 days after elective THA.
 
-**PARTICIPANTS AND METHODS:** We analyzed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden.
+**PARTICIPANTS AND METHODS:** This prediction model was derived on `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register and validated on `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register.
+Data on comorbidities and socioeconomic background variables on all patients were obtained from the National Registers and.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery.
-Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from The Danish Hip Arthroplasty Register.
-We assessed model performance by the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.
+We assessed model performance by the area under the receiver operation characteristics curve (AUC), calibration belt plots and the fraction of new information.
 
 **RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort.
-A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally.
-These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("ASA class")`, Charlson `r get_auc("Charlson")`, Elixhauser `r get_auc("Elixhauser")`, and the Rx Risk V `r get_auc("Rx Risk V")` comorbidity indices.
+A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and `r sub("^AUC = ", "", get_auc("Reduced model", "2", Country = "Denmark"))` externally.
+These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("ASA")`, or the Charlson `r get_auc("Charlson")`, Elixhauser `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` comorbidity indices.
 
-**CONCLUSION:** The proposed prediction models for PJI after THA are relatively parsimonious since they are based on easily accessible clinical data, and they are superior to predictions based on ASA class or more complex comorbidity indices.
+**CONCLUSION:** Our parsimonious prediction model for PJI after THA is based on easily accessible clinical data, and it is superior to predictions based on ASA class alone or more complex comorbidity indices.
 Our web-based calculator could simplify the individualized assessment of the risk of developing PJI prior to THA surgery and enable risk-stratified patient management.
+Nonetheless, the remaining unexplained variance indicates that unmeasured factors play an important role in the pathogenesis of PJI.
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
+Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
 Although high ASA class is a risk factor associated with development of PJI, this classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
@@
 Thus, there is an urgent need for improved prediction models of PJI.
-We therefore developed a prediction model of PJI within 90 days or two years after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.
+We therefore developed a prediction model of PJI within 90 days after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.
 
@@
 
-The derivation cohort was obtained from patients recorded in the SHAR `r period`.
-The starting point was chosen because registration of ASA class and body mass index (BMI) in the SHAR was initiated in 2008, and we had no access to data from the national patient register (NPR) after 2015.
-Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
-Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.
+The derivation cohort was obtained from patients recorded in the SHAR `r period` with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) in the SHAR was initiated then.
+Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
+Data on comorbidities for one year prior to index surgery were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
+Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
 The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
@@
 
+Danish THA patients were operated from 2016 until the last of September 2018, and this period was chosen due to availability of data and thus follows after the period used for model derivation in Sweden.
+
+**TODO: ADD INFO (data sources)!**
+
 ## Definitions of comorbidity and outcomes
@@
 Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
-PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-codes recorded in the NPR within 90 days (Tab.
-\@ref(tab:infcodes)), or if the patient was reported to SHAR as re-operated due to infection within the same time frame.
+PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the NPR within 90 days (Tab.
+\@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.
 
@@
 
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Age was normalized prior to modeling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
 BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
 
-Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model.
-We kept the $\lambda$ that minimized the mean cross-validated deviance.
+Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance.
 100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
@@
 The second change point was then used as a break-point.
-Influential predictors with absolute coefficients value above this point, scored a point.
-The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total).
+Influential predictors with absolute coefficients value above this point scored a point and the entire process was repeated 100 times, yielding 10,000 bootstrap samples in total.
 Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of numeric variables.
-An even more parsimonious model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison.
-Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison.
+An even more parsimonious "reduced" model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times was evaluated for comparison.
+Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, gender and BMI (10 models in total).
 The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
@@
 Each developed model was used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
-Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples.
-The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a Danish cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to Denmark.
+Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
+Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
+The model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$.
+This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to the baseline model.[@Harrell2018]
+
+The derived model was then externally validated on the Danish cohort, both with the coefficient estimates from the Swedish cohort and with new estimates based on the Danish cohort.
+The 10 simpler models, with adopted coefficient estimates, were also applied to the Danish cohort for comparison.
+Finally, the new coefficient estimates for the derived model, re-estimated with Danish data, were empirically compared to analogues estimates derived from subsamples of the Swedish data with sample sizes to match the Danish cohort (supplementary, Tab \@ref(tab:comparecoefs) and Fig.
+\@ref(fig:coefsspread)).
 
@@
 
-We built an online web calculator available at **XXX** to be used in clinical practice.
-`r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder[@Bulow2020b] and shiny were used.
+We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder[@Bulow2020b], decoder, furrr, pROC, rms, tidymodels and tidyverse.
 
@@
 
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and **XXX**.
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by **XXX in Denmark**.
 
@@
 We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig.
-\@ref(fig:flowchart), left panel).
-`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of the patients developed a PJI within 90 days (Tab.
+\@ref(fig:flowchartse)).
+`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of these developed a PJI within 90 days (Tab.
 \@ref(tab:tab1)).
 The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
-In addition, **XXX** patients from DHR were included as an external validation cohort (Fig.
-\@ref(fig:flowchart), right panel).
+
+In addition, `r nmbr(N_dk)` patients **(42.1 % males)** with a mean age of **XXX (SD = XXX)** years were included from DHR as an external validation cohort (Fig.
+\@ref(fig:flowchartdk)).
+**2.2 % (n = 410)** of these developed a PJI within 90 days (Tab.
+\@ref(tab:tab1)).
+The proportion of patients with ASA class III was **18 % (n = 3,334)**; **14 % (n = 2,722)** had at least one comorbidity according to the Charlson comorbidity index, **22 % (n = 4,101)** according to Elixhauser index, and **XXX % (n = XXX)** had medications classified according to Rx Risk V.
+
+**TODO:: Add XXX numbers frmo DK. Differences in Charlson/Elix might be discussed?**
 
@@
 
-Patients with `r excl_factors` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling.
+Patients with `r excl_factors` were uncommon, why these comorbidities were excluded as potential predictors prior to further modelling.
 ROC-curves (Fig.
-\@ref(fig:rocs)) and AUC-values (Fig.
-\@ref(fig:aucci)) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced models.
+\@ref(fig:rocs) left panel) and AUC-values (Fig.
+\@ref(fig:aucci) left panel) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced model.
+
+```{r}
+CCI_AUC <- sub("AUC = ", "", get_auc("Charlson"))
+CCI_R2 <-
+  tab_new_info %>% 
+  subset(Model == "Charlson", "R2 (%)", drop = TRUE)
+red_FNI <- 
+  tab_new_info %>% 
+  subset(Model == "Reduced model", "FNI (%)", drop = TRUE) %>% 
+  round()
+```
+
+The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text` (Tab.
+\@ref(tab:coefs)).
+This reduced model had an `r get_auc("Reduced model", "2")`, whereas ASA class `r get_auc("ASA")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` had less discriminative ability (Fig.
+\@ref(fig:rocs) left panel and \@ref(fig:aucci) left panel).
+The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2$ (`r CCI_R2`; Tab.
+\@ref(tab:fni)), and compared to this model, the fraction of new information for the reduced model was `r red_FNI` %.
+
+The reduced model had good internal calibration (Fig.
+\@ref(fig:calibrationse)).
 Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for this outcome (Fig.
@@
 
-The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text` (Tab.
-\@ref(tab:coefs)).
-This reduced model for the predicition of PJI within 90 days had an `r get_auc("Reduced model", "2")`, whereas ASA class `r get_auc("ASA class")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk V")` had less discriminative ability (Fig.
-\@ref(fig:rocs) and \@ref(fig:aucci)).
-The reduced model had good accuracy as visualized by the calibration plot (Fig.
-\@ref(fig:calibration), left panel).
-
 ## External validation
 
-XXX
+The reduced model derived from the Swedish cohort produced a very similar ROC curve when applied to the Danish cohort (Fig.
+\@ref(fig:rocs) right panel), and its AUC value was **0.66 (95% CI: 0.64 to 0.69)**, thus not statistically significantly different from the derivation model.
+This did not change for the model with re-estimated coefficient values (data not shown).
+AUC values for some of the other models estimated on the Danish cohort did differ from the Swedish cohort, with a statistically significantly different estimate obtained for the two models based on the Rx Risk V, but all those models were still less accurate compared to the reduced model (Fig.
+\@ref(fig:aucci) right panel).
+The re-estimated model coefficients of the reduced model were coherent with the estimates based on the Swedish data (supplementary Tab.
+\@ref(tab:comparecoefs)).
 
@@
 
-We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
-This model also performed better than the established comorbidity measures in terms of AUC and calibration.
+We found that a reduced prediction model based on some easily identifiable clinical conditions, gender, and BMI was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
+This model performed better than the established comorbidity measures in terms of AUC, Nagelkerke's $R^2$ and calibration.
 
@@
 \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics.
-For example, `r examples$desc` for PJI within 90 days.
-Note, however, that observed proportions above 10 % were rare.
-Predicted probabilities above this limit are therefore subject to extrapolation.
+For example, `r examples$desc[[1]]` would have a probability of `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
+Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
 
@@
 
-The predictors in our models were chosen based on a strict analytical approach, not on proposed relevance of patient or procedure characteristics.
+The predictors in our models were chosen based on a combination of clinical and analytical reasons, but they should not be interpreted as having direct causal effects on the outcome of PJI.
 The majority of the identified predictor variables are, however, associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.
 
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred a nearly two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] and this predictor variable was statistically significance in our model of PJI within two years.
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred a nearly two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] and this predictor variable was also associated with a more then doubled risk of PJI in our model.
 
-Obesity is associated with an increase in the risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with the risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both models.[@Tan2018] In accordance with observational studies[@Jamsen2012] we found the presence of CNS diseases to be a predictor of PJI within two years, and our rather broad category encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease which were all associated with a risk increase in the cited study.
+Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both models.[@Tan2018] In accordance with observational studies[@Bozic2012] we found the presence of CNS diseases to be a predictor of PJI, and our broad definition of CNS diseases encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease, most of which were associated with a risk increase in the cited study.
 
-Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015]Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators, which agrees with our findings.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+
+High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as is comorbidity in a broader sense, and, more specifically, the presence of cardiovascular comorbidity.[@Ong2009; @Higuera2011] We found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model, but ASA class was one of the predictor variables in our model.
 
 Some previously mentioned risk factors for developing PJI were not selected during our model development.
-High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as are comorbidity in a broader sense, and, more specifically, the presence of cardiovascular comorbidity.[@Ong2009; @Higuera2011] In contrast, we found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model.
-
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a] 20 but this comorbidity did not reach statistical significance in our models.
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a] but this comorbidity did not reach statistical significance in our models.
 
 THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
-We found no indication that the type of fixation was associated with the risk of PJI, and this variable was also not selected as a potential predictor variable in any of the iterations during model development.
+We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.
 
-Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index alone, resulted in poor predictive power.
-This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019]
+Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
+This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our reduced model.
 
@@
 Our study is hampered by the limitations inherent to observational research, such as misclassification, and residual confounding due to known confounders that we had no information on, or due to unknown confounders.
-Importantly, some previously identified risk factors for PJI were not included in our models, mostly because we lacked detailed information on laboratory findings and non-prescribed medications.
 
@@
 Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variables.
-Also, the individual odds ratios should not be interpreted as effect measures of individual relations.
-Some variables are correlated.
-It is therefore not always possible to hold all other variables constant while changing only the relevant one.
+Also, the individual odds ratios should not be interpreted as effect measures of individual relations since some variables are correlated.
 
 The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
-This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative hemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anemia, we had no information on actual laboratory findings.
-Use of TNF-a blockers may be associated with the risk of developing PJI.[@Gilson2010; @Momohara2011] Similarly, intra-articular steroid application may enhance this risk,[@Cancienne2015] but injections given by physicians would not necessarily be identifiable within a Swedish registry setting.
+This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycaemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative haemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
+Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers may be associated with the risk of developing PJI,[@Gilson2010; @Momohara2011] and intra-articular steroid application may also enhance this risk,[@Cancienne2015] but injections given by physicians would not necessarily be identifiable within a Swedish registry setting.
 Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter.
+
+The proportion of explained variation (Nagelkerke's $R^2$ = `r subset(tab_new_info, Model == "Reduced model", "R2 (%)", drop = TRUE)`) was low.
 The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
-
-Taken together, the inclusion of some or all of the variables mentioned above -- unfortunately unavailable to us -- might improve the predictive power of PJI risk assessment models.
+In addition, a potential underlying susceptibility to PJI -— as has been described for patients with osteomyelitis -— could be a factor of considerable influence on the outcome, but information on single nucleotide polymorphisms in genes that have been associated with musculoskeletal infections was unavailable to us.
+Such shortcomings could be addressed by future genome-wide association studies.
 
@@
 
-Our results indicate that the risk of PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
+Our results indicate that the risk of early PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
 We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
@@
 EB developed the statistical model.
-ITA and UH performed the external validation.
+~~ITA~~**???**, UH and AM performed the external validation.
 NPH and EB drafted, and all authors edited and finalized the manuscript.
@@
 
-We would like to thank .
+We would like to thank ...
 
@@
 
-```{r flowchart, fig.cap = "Flowchart with inclusion criteria and number of patients.
-Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation (left).
-Data from XXX were used for external validation (right)."}
+```{r flowchartse, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation"}
 knitr::include_graphics("../graphs/flowchart.png")
+```
+
+```{r flowchartdk, fig.cap = "Flowchart with inclusion criteria and number of patients.
+Data from the Danish Hip Arthroplasty Register were used for external model validation.**TODO: This figure is a screenshot from a PDF.
+Can we have the original figure file with high resolution?**"}
+knitr::include_graphics("../validation/dk_flowchart.png")
 ```
@@
 ```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
-The main and reduced models performed almost identival for prediction of PJI within 90 days after surgery."}
+The main and reduced models performed almost identical for prediction of PJI within 90 days after surgery.
+They both performed better than all other models (left panel).
+The reduced model performed as well when applied to the Danish validation data set (right panel)."}
 knitr::include_graphics("../graphs/roc.png")
@@
 
-```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals.
-The reduced model performs no different than the main model, and both of these perform better than traditional comorbidity measures alone."}
+```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % confidence intervals (De-Long).
+The reduced model performed no different than the main model on the Swedish data, and both of these models performed better than all other models (left panel).
+Similar models were fitted to the Danish cohort (right panel).
+Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as Reduced model (DK) with coefficient values refitted to the Danish cohort."}
 
@@
 ```{r sep, fig.cap = "The vast majority of patients did not have PJI.
-Hence, the blue bars dominates the histograms (left panel; note the scales).
+Hence, the blue bars dominate the histograms (left panel; note the scales).
 A normalized density plot reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
@@
 
-```{r calibration, fig.cap = "This figure illustrates calibration between observed proportions and predicted probabilities with 95 % confidence intervals.
-Calibration is good for predictions up to 5 %, although higher probabilities tend to over-estimate the observed outcome.
-"}
+```{r calibrationse, fig.cap = "This figure illustrates internal calibration of the reduced model, used in the Swedish cohort. Predicted probabilities are similar to observed proportions."}
 knitr::include_graphics("../graphs/calibration.png")
+```
+
+```{r calibrationdk, fig.cap = "This figure illustrates external calibration of the reduced model, with Swedish coefficient values, used in the Danish cohort.
+Predicted probabilities are similar to observed proportions.
+**TODO: Could EB get an export of the R-object to adapt the graphical presentation?**"}
+knitr::include_graphics("../validation/calibration.tiff")
 ```
@@
   knitr::kable(
-    caption = "Categorization of comorbidities from the Charlson, Elixhauser and Rx RIsk V comorbidities. (CNS = central nervous system.)")
+    caption = "Categorization of comorbidities from the Charlson, Elixhauser and Rx Risk V comorbidities. (CNS = central nervous system.)")
 
@@
 
+table1 %>% 
+  mutate(
+    what = replace(what, duplicated(what), "")
+  ) %>% 
 knitr::kable(
-  mutate(table1, what = gsub("Heart infarct", "Myocardial infarction", what)), 
   caption = "Patient characteristics in the model derivation cohort.
@@
 BMI = Body mass index.
-ASA class = American Society for Anesthesiologists classification.
+ASA class = American Society for Anaesthesiologists classification.
 CNS = central nervous system.")
@@
 (BMI = body mass index.
-ASA class = American Society for Anesthesiologists classification.
+ASA class = American Society for Anaesthesiologists classification.
 CNS = central nervous system.)")
+```
+
+##### PAGE BREAK
+
+```{r fni}
+tab_new_info %>% 
+  arrange(desc(`R2 (%)`)) %>% 
+knitr::kable(
+  caption = "Nagelkerke's pseudo-coefficient of determination (R2) as an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
+The univariable Charlson model had the lowest R2 and was therefore used as reference for the fraction of new information (FNI) added by the other models.")
 ```
@@
   caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models.
-BMI = body mass index ("under/normal weight as baseline).
+BMI = body mass index (under/normal weight as baseline).
 CNS = central nervous system.
@@
 
+# Supplementary
+
+## Model coefficients
+
+Variable selection and coefficient estimates were based on the Swedish cohort. 
+The reduced model was then used to predict PJI also in Denmark.
+The coefficient values of the reduced model were also re-fitted to the Danish cohort for comparison.
+We compared the estimates from the Swedish and Danish cohort and found most values to be of similar magnitude.
+The only coefficient with reversed direction was for patients with a diagnose of "sequelae after childhood hip disease".
+This is a rare condition, with very few observed PJI:s within 90 days.
+We also re-estimated the model coefficients 1,000 times based on the Swedish data.
+Each time, we took a random sample of the same size, $N = 18854$, as was observed in the Danish cohort.
+The 2.5th and 97.5th percentiles were then used to form empirical 95 % confidence intervals for each coefficient ((Tab.
+\@ref(tab:comparecoefs)) and Fig.
+\@ref(fig:coefsspread)).
+It is seen that the estimated effect of the diagnose of "sequelae after childhood hip disease" fall outside this empirical CI.
+Applying multiplicity correction by Bonferroni or similar, would eliminate this significance.
+Hence, it seems that both cohorts are similar in respect to association between the studied covariates and the risk of PJI within 90 days of THA.
+
+```{r comparecoefs}
+compare_se_dk_coefs %>% 
+knitr::kable(
+  caption = ""
+)
+```
+
+```{r coefsspread, fig.cap = "Estimated coefficient values for the reduced model.
+Red triangles indicate estimates from the Swedish derivation data set.
+Red circles indicate re-estimated coefficient values for the same model based on the Danish validation cohort.
+Small coulored dots are empirical estimates from 1000 subsamples based on the Swedish data set, of the same sample size as used in the Danish validation cohort.
+Black dots indicate the outer limits of empirical confidence intervals based on the 2.5th and 97.5th percentiles.
+It is seen that the Swedish and Danish estimates are similar.
+The Danish estimates almost always fall within the 95 % empirical CIs.
+The sole exception is for the rare diagnose group of 'sequelae after childhood hip disease'."}
+knitr::include_graphics("../validation/Swedish_model.png")
+```
+
+## Stratified separation
+
+**TODO: If we decide to include those figures, could we get the Danish ggplot2-objects as well? Would you also like to add some descriptive text in that case Ute?**
+
+```{r sepsex, fig.cap = "Predicted probability by the reduced model, stratified by outcome and sex."}
+knitr::include_graphics("../graphs/separation_strata_sex.png")
+```
+
+```{r sepasa, fig.cap = "Predicted probability by the reduced model, stratified by outcome and ASA class."}
+knitr::include_graphics("../graphs/separation_strata_asa.png")
+```
+
+```{r sepbmi, fig.cap = "Predicted probability by the reduced model, stratified by outcome and Body Mass index (BMI)."}
+knitr::include_graphics("../graphs/separation_strata_bmi.png")
+```
+
+```{r sepdiagrp, fig.cap = "Predicted probability by the reduced model, stratified by outcome and diagnose group."}
+knitr::include_graphics("../graphs/separation_strata_diagrp.png")
+```
+
+##### PAGE BREAK
+
 # Bibliography

### 2021-05-26 munge/04-remove_empty_variables.R

@@ -107,7 +107,8 @@ cache("infection_data")
 # Save list of excluded co-morbidities for presentation
 excl_factors <-
   comb_rm$lgl_text %>%
+  tolower() %>%
   {gsub("pancreatii", "pancreatic i", .)} %>%
-  {gsub("AIDS/HIV hiv", "AIDS/HIV", .)}
+  {gsub("aids/hiv hiv", "AIDS/HIV", .)}
 
 cache("excl_factors")

### 2021-05-26 munge/06-compare_models.R

@@ -33,13 +33,19 @@ future::plan("multiprocess", workers = 2)
 
 models <-
   tribble(
-    ~Model,         ~fit,
-    "Reduced model",glmdf(form(coefs_selected$best_coefs_reduced)),
-    "Main model",   glmdf(form(coefs_selected$best_coefs)),
-    "Charlson",     glmdf("outcome ~ CCI_index_quan_original"),
-    "Elixhauser",   glmdf("outcome ~ ECI_index_sum_all"),
-    "Rx Risk V",    glmdf("outcome ~ Rx_index_pratt"),
-    "ASA class",    glmdf("outcome ~ P_ASA"),
+    ~Model,                    ~ fit,
+    "Reduced model",           glmdf(form(coefs_selected$best_coefs_reduced)),
+    "Main model",              glmdf(form(coefs_selected$best_coefs)),
+    "Charlson",                glmdf("outcome ~ CCI_index_quan_original"),
+    "Charlson + Age + Sex",    glmdf("outcome ~ CCI_index_quan_original + P_Age + P_Sex"),
+    "Elixhauser",              glmdf("outcome ~ ECI_index_sum_all"),
+    "Elixhauser + Age + Sex",  glmdf("outcome ~ ECI_index_sum_all + P_Age + P_Sex"),
+    "Rx Risk",                 glmdf("outcome ~ Rx_index_pratt"),
+    "Rx Risk + Age + Sex",     glmdf("outcome ~ Rx_index_pratt"),
+    "ASA",                     glmdf("outcome ~ P_ASA"),
+    "ASA + Age + Sex",         glmdf("outcome ~ P_Age + P_Sex + P_ASA"),
+    "BMI + Age + Sex",         glmdf("outcome ~ P_Age + P_Sex + P_BMI"),
+    "ASA + BMI + Age + Sex",   glmdf("outcome ~ P_Age + P_Sex + P_ASA + P_BMI")
   ) %>%
   mutate(
     tidy          = map(fit, broom::tidy, conf.int = TRUE, exponentiate = TRUE),
@@ -52,7 +58,7 @@ models <-
     AUCci         = furrr::future_map(
                       ROC,
                       pROC::ci.auc,
-                      method = "bootstrap",
+                      method = "delong",
                       boot.stratified = FALSE,
                       .progress = TRUE,
                       .options = furrr_options(seed = TRUE)
[…]

### 2021-05-26 src/examples.R

@@ -33,10 +33,8 @@ examples <-
   mutate(
     p = sprintf("%.1f %%", p * 100),
     desc = c(
-      "a 60 year old female with normal BMI, primary osteoarthritis and no co-morbidities",
-      "a 85 year old male with overweight, secondary osteoarthritis, psoriasis, dementia
-       (CNS disease) and ASA class III"),
-    desc = sprintf("%s would have a probability of %s", desc, p)
+      "a 60-year-old female with normal BMI and primary osteoarthritis without comorbidities",
+      "an 85-year-old male with overweight, secondary osteoarthritis, psoriasis, dementia and ASA class III")
   )
 
 cache("examples")

### 2021-05-26 src/export_for_validation.R

@@ -18,8 +18,7 @@ model_vars <-
 # Export data frame of data needed for external validation
 ext_val_required_data <-
   df %>%
-  select(one_of(model_vars)) %>%
-  group_by(P_DiaGrp) %>%
-  slice(1:3)
+  select(one_of(model_vars))
 
 cache("ext_val_required_data")
+

### 2021-05-26 src/fig_auc.R

@@ -10,12 +10,14 @@ options(digits = 2)
 
 data_auc_ci <-
   models %>%
-  select(Model, starts_with("AUC")) %>%
+  select(country, Model, starts_with("AUC")) %>%
   mutate(
+    Model  = gsub("Reduced model$", "Reduced model (SE)", Model),
     Model  = reorder(Model, AUC_est),
     ci     = sprintf("95 %% CI: %.2f to %.2f", AUC_lo, AUC_hi),
     text1  = sprintf("(AUC = %.2f, %s)", AUC_est, ci),
     text2  = sprintf("AUC = %.2f (%s)", AUC_est, ci),
+    country = factor(country, c("se", "dk"), c("Sweden", "Denmark"))
   )
 
 cache("data_auc_ci")
@@ -27,33 +29,35 @@ annotates <-
 # Figure ------------------------------------------------------------------
 
 ggplot(data_auc_ci, aes(Model, AUC_est)) +
-  #geom_hline(yintercept = 0.7, color = "darkgreen", size = 1, linetype = "dashed") +
+  geom_hline(yintercept = 0.7, color = "grey") +
   geom_pointrange(
     aes(ymin = AUC_lo, ymax = AUC_hi),
-    #size = .5,
+    size = .33,
     position = position_dodge(width = .25)
   ) +
   geom_text(
     aes(Model, value, label = round(value, 2)),
     data = annotates,
     nudge_x = .3,
-    size = 3,
-    #position = position_dodge(width = .85)
+  #  nudge_y = .5,
+    size = 1.5,
+    # position = position_dodge(width = .85)
   ) +
   coord_flip() +
-  theme_minimal(15) +
+  theme_minimal(10) +
   theme(
     axis.title.y = element_blank(),
-    # legend.title = element_blank(),
+    legend.title = element_blank(),
     legend.position = "bottom",
     panel.grid.major.y = element_blank(),
     panel.grid.minor.y = element_blank(),
-    #panel.grid.major.x = element_blank(),
+    panel.grid.major.x = element_blank(),
     panel.grid.minor.x = element_blank(),
-    #axis.text.x = element_blank()
+    # axis.text.x = element_blank()
   ) +
-  ylab("AUC with 95% CI")
+  ylab("AUC with 95% CI") +
+  facet_wrap(~ country, scales = "free_x")
 
-ggsave("graphs/auc_ci.png", width = 15, height = 8, units = "cm")
[…]

### 2021-05-26 src/fig_roc.R

@@ -12,21 +12,41 @@ shar_roc <-
     Model = model_names(Model)
   )
 
+
+load("validation/export_DK.RData")
+dk <-
+  roc_plot_coords %>%
+  mutate(Model = "Reduced model (DK)") %>%
+  rename(
+    specificity = specificities,
+    sensitivity = sensitivities
+  )
+
+ggdata <-
+  bind_rows(
+    Sweden  = shar_roc,
+    Denmark = filter(shar_roc, Model == "Reduced model"),
+    Denmark = dk,
+    .id = "country"
+  ) %>%
+  mutate(country = factor(country, c("Sweden", "Denmark")))
+
 # Figure ------------------------------------------------------------------
 
-fig_roc <-
-  shar_roc %>%
+ggdata %>%
   ggplot(aes(1 - specificity, sensitivity, col = Model)) +
-  geom_path(size = 1) +
+  geom_path() +
   geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
-  theme_minimal(15) +
+  theme_minimal(10) +
   theme(
     legend.position = "bottom", # c(1, 0),
     #legend.justification = c(1, 0),
-    legend.title = element_blank()
+    legend.title = element_blank(),
+    legend.text = element_text(size = 4.5)
   ) +
   scale_x_continuous(labels = scales::percent) +
-  scale_y_continuous(labels = scales::percent)
+  scale_y_continuous(labels = scales::percent) +
+  facet_wrap(~ country)
 
 
-ggsave("graphs/roc.png", fig_roc, height = 15, width = 15, units = "cm")
+ggsave("graphs/roc.png", height = 10, width = 15, units = "cm", dpi = 900)

### 2021-05-26 src/fig_separation_strata.R

@@ -1,0 +1,74 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+load("cache/models.RData")
+load("cache/df.RData")
+
+digs <- options("digits")
+options(digits = 2)
+
+library(ggridges)
+
+
+plotstratified <- function(stratum) {
+  usedat <- tibble(
+    outcome = df$outcome,
+    pred = predict(models$fit[[1]], df, type = "response"),
+    stratum = as.factor(df[[stratum]])
+  )
+  stratlevel <- levels(usedat$stratum)
+  obsprob <- tibble(
+    trueprop = tapply(usedat$outcome, usedat$stratum, mean),
+    meanpred = tapply(usedat$pred,    usedat$stratum, mean),
+    stratum = seq_along(stratlevel)
+  )
+  ggplot(usedat, aes(y = stratum)) +
+    ggridges::geom_density_ridges(aes(x = pred,
+                                      fill = paste(stratum, factor(outcome))),
+                                  alpha = .3,
+                                  scale = .8) +
+    geom_segment(
+      data = obsprob,
+      mapping = aes(
+        x = trueprop,
+        xend = trueprop,
+        y = as.numeric(stratum) - .05,
+        yend = as.numeric(stratum) + .5
+      ),
+      inherit.aes = FALSE,
+      color = "red"
+    ) +
+    geom_segment(
+      data = obsprob,
+      mapping = aes(
+        x = meanpred,
+        xend = meanpred,
+        y = as.numeric(stratum) - .05,
+        yend = as.numeric(stratum) + .5
+      ),
+      inherit.aes = FALSE,
+      color = "black"
+    ) +
+    scale_fill_cyclical(
+      values = c(" violetred", "royalblue"),
+      name = "outcome",
+      guide = "legend",
+      labels = c("No PJI", "PJI")
+    ) +
+    scale_y_discrete(expand = c(0, 0)) +
+    scale_x_continuous(limits = c(0, .1)) +
+    theme_minimal() +
+    theme(
+      axis.title.y = element_blank(),
+      legend.position = "bottom",
+      legend.title = element_blank()
+    ) +
+    xlab('predicted probability')
+}
+plotstratified("P_BMI")
+  ggsave("graphs/separation_strata_bmi.png", width = 10, height = 15, units = "cm")
+plotstratified("P_Sex")
[…]

### 2021-05-26 src/tab_coefs.R

@@ -16,6 +16,7 @@ baseline <-
     )
   ) %>%
   mutate(beta = 0, `OR 95 % CI` = "(ref)", p = "")
+cache("baseline")
 
 
 # Prepare temp table ------------------------------------------------------
@@ -73,6 +74,7 @@ coefs_text <-
       variable == "cns disease" ~ "CNS disease",
       variable == "pancreatiinsufficiency" ~ "pancreatic insufficiency",
       variable == "rheumatidisease" ~ "rheumatic disease",
+      variable == "lung airways disease" ~ "lung and airways disease"
       TRUE ~ variable
     ),
     ordinary = fct | variable == "age"
@@ -83,7 +85,7 @@ coefs_text <-
   summarise(
     text = if_else(
       !ordinary,
-      paste("the precense of", glue::glue_collapse(variable, sep = ", ", last = " or ")),
+      paste("the presence of", glue::glue_collapse(variable, sep = ", ", last = " or ")),
       paste(variable, collapse = ", ")
     )
   ) %>%
@@ -94,3 +96,5 @@ coefs_text <-
   deframe()
 
 cache("coefs_text")
+
+

### 2021-05-26 src/tab_new_info.R

@@ -1,0 +1,34 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+load("cache/models.RData")
+load("cache/df.RData")
+
+# Table of added values according to Harrell:
+# https://www.fharrell.com/post/addvalue/
+
+se_mods_lrm <-
+  models %>%
+    filter(country == "se") %>%
+    transmute(
+      Model,
+      `AUC (95 % CI)` = sprintf("%.2f (%.2f; %.2f)", AUC_est, AUC_lo, AUC_hi),
+      lrm = map(fit, ~ lrm(formula(.), df)),
+      R2  = map_dbl(lrm, ~ .$stats['R2'])
+    )
+
+# PLocka ut R2
+base_R2 <-
+  se_mods_lrm %>%
+  filter(R2 == min(R2)) %>%
+  select(R2) %>%
+  pluck(1)
+
+tab_new_info <-
+  se_mods_lrm %>%
+  mutate(
+    FNI = map_dbl(R2, ~ 1 - base_R2 / .)
+  ) %>%
+  transmute(
+    Model, `AUC (95 % CI)`, `R2 (%)` = R2 * 100, `FNI (%)` = FNI * 100)
+
+cache("tab_new_info")

### 2021-05-26 src/tab_table1.R

@@ -5,10 +5,11 @@ load("cache/df.RData")
 dft1 <-
   df %>%
   mutate(
-    Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original == 4, "4+"),
-    Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all == 3, "3+"),
-    RxRiskV    = Rx_index_pratt,
-    P_BMI      = relevel(P_BMI, "under/normal weight")
+    Charlson   = replace(CCI_index_quan_original, CCI_index_quan_original > 1, "2+"),
+    Elixhauser = replace(ECI_index_sum_all, ECI_index_sum_all > 2, "3+"),
+    RxRiskV    = cut(Rx_index_pratt, c(-Inf, 0, 3, 6, Inf), c("-5-0", "1-3", "4-6", "7+")),
+    P_BMI      = relevel(P_BMI, "under/normal weight"),
+    P_Age      = cut(P_Age, c(0, 49, 59, 69, 79, Inf), c("<50", "50-<60", "60-<70", "70-<80", "80+"))
   ) %>%
   select(
     PJI = outcome,
@@ -29,58 +30,165 @@ dft1 <-
   ) %>%
   setNames(gsub(": ", "", clean_names(names(.))))
 
-fct <- c("Sex", "ASA class", "Elixhauser",
-           "Charlson", "Diagnosis", "Cemented stem", "Cemented cup")
-
 t1 <-
   tableone::CreateTableOne(
     strata = "PJI",
     vars = setdiff(names(dft1), "PJI"),
-    factorVars = fct,
     data = dft1,
     test = FALSE
   ) %>% print(
-    printToggle = FALSE
+    printToggle = FALSE,
+    showAllLevels = TRUE
   ) %>%
   as_tibble(rownames = "what") %>%
   rename(
-    `PJI` = `TRUE`,
-    `No PJI` = `FALSE`
+    `SE PJI` = `TRUE`,
+    `SE No PJI` = `FALSE`
+  ) %>%
+  mutate(
+    what = na_if(what, ""),
+    what = zoo::na.locf(what),
+    what = gsub(":? \\(%\\)", "", what)
   )
 
 
 t1_tot <-
   tableone::CreateTableOne(
     vars = setdiff(names(dft1), "PJI"),
-    factorVars = fct,
     data = dft1,
-    test = FALSE
+    test = FALSE,
   ) %>%
   print(
[…]

### 2021-05-26 validation/subsample_SE_data.R

@@ -1,0 +1,329 @@
+#' ---
+#' title: Fitting Swedish hip replacement model to subsampled data
+#' author: ute
+#' date: 25.3.2021
+#' ---
+#'
+#' Is the discrepancy between the coefficients of Eriks model, when fitted
+#' to Danish data, perhaps just due to the smaller sample size?
+#'
+#' As a first experiment, we could look at the behaviour when fitted to
+#' subsamples of the Swedish data set. The following code collects results
+#' from 1000 subsamples and should not take long to run.
+#'
+#' It would be fantastic if you could share the results with me :-)
+#'
+
+# change here for your own data paths, and adjust names where necessary
+
+# ugly but useful
+if (!endsWith(getwd(), "validation"))
+  setwd("validation")
+
+load("../cache/model_reduced_lean.RData")
+load("../cache/df.RData")
+SE_data <- df
+
+N_SE <- nrow(SE_data)
+N_DK <- 18854L
+
+nsubsamp <- 1000
+
+if (!file.exists("swedish_subsample_fits.Rdata")) {
+  result <- replicate(nsubsamp,
+      unlist(
+          glm(model_reduced_lean$formula,
+              binomial,
+              data = SE_data[sample(N_SE, N_DK, replace = FALSE), ]
+          )[c("coefficients", "deviance", "aic")]
+      )
+  )
+
+  save(result, file="swedish_subsample_fits.Rdata")
+}
+
+load("swedish_subsample_fits.Rdata")
+
+summary(t(result))
+
+
+
+#' # Compare model coefficients
+
+suppressPackageStartupMessages(library(tidyverse))
+
+load("model_dk_superlean.RData")
+
+cfs <-
+  left_join(
+    enframe(coef(model_reduced_lean)),
+    enframe(coef(model_dk_superlean)),
+    suffix = c(".Sweden", ".Denmark"),
+    by = "name"
+  ) %>%
+  rename(coef = name) %>%
+  filter(!grepl("Intercept", coef)) %>%
+  pivot_longer(starts_with("value"), names_to = "orig", names_prefix = "value.") %>%
+  mutate(coef = clean_names(coef))
+
+
+#' # Plot
+
+#' Help function to plot the coefficients
+plot_data <- function(results) {
+  result %>%
+    as_tibble(rownames = "coef") %>%
+    filter(
+      startsWith(coef, "coef"),
[…]

### 2021-05-26 reports/article.Rmd
version to share

@@
 **PARTICIPANTS AND METHODS:** This prediction model was derived on `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register and validated on `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register.
-Data on comorbidities and socioeconomic background variables on all patients were obtained from the National Registers and.
+Data on comorbidities and socioeconomic background variables on all patients were obtained from the National Registers.
 Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery.
@@
 
-**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort.
+**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort and **X.XX %** in Denmark.
 A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and `r sub("^AUC = ", "", get_auc("Reduced model", "2", Country = "Denmark"))` externally.
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male gender are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
+Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
 Although high ASA class is a risk factor associated with development of PJI, this classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
@@
 We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
-The best models were then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).
+The best model was then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).
 
@@
 
-The derivation cohort was obtained from patients recorded in the SHAR `r period` with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) in the SHAR was initiated then.
+The derivation cohort was obtained from patients recorded in the SHAR `r period` with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then.
 Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on comorbidities for one year prior to index surgery were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates.
+Data on comorbidities for one year prior to index surgery were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
@@
 An even more parsimonious "reduced" model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times was evaluated for comparison.
-Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, gender and BMI (10 models in total).
+Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total).
 The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
-This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals.
+This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals based on the whole cohort.
 
@@
 
-**TODO:: Add XXX numbers frmo DK. Differences in Charlson/Elix might be discussed?**
+**TODO:: Add XXX numbers for DK. Differences in Charlson/Elix might be discussed?**
 
@@
 
-We found that a reduced prediction model based on some easily identifiable clinical conditions, gender, and BMI was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
+We found that a reduced prediction model based on some easily identifiable clinical conditions, sex, and BMI was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
 This model performed better than the established comorbidity measures in terms of AUC, Nagelkerke's $R^2$ and calibration.
@@
 
-Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab.
-\@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics.
-For example, `r examples$desc[[1]]` would have a probability of `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
+Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/[1 + \mathrm{exp}(-\hat \alpha - x'\hat \beta)]$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\hat \beta = (\hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k)$ are the estimated intercept and coefficients (Tab.
+\@ref(tab:coefs)) and $x = (x_1, x_2, \ldots, x_k)$ are corresponding patient characteristics.
+For example, `r examples$desc[[1]]` would have a probability of $1/[1 + \mathrm{exp}(6.31 - 60\cdot0.02)] \approx$ `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
 Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
@@
 
-Male gender is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and gender is included as a predictor variable in several previously developed risk calculators, which agrees with our findings.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
+Male sex is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and sex is included as a predictor variable in several previously developed risk calculators, which agrees with our findings.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
 
@@
 Similar models were fitted to the Danish cohort (right panel).
-Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as Reduced model (DK) with coefficient values refitted to the Danish cohort."}
+Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as the reduced model (DK) with coefficient values refitted to the Danish cohort."}
 
@@
 
-```{r sep, fig.cap = "The vast majority of patients did not have PJI.
-Hence, the blue bars dominate the histograms (left panel; note the scales).
-A normalized density plot reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
+```{r sep, fig.cap = "Separation between patients with and without PJI within 90 days versus their predicted probabilities of this event.
+Most patients did not have PJI, wherefore the blue bars dominate the histogram (left panel; note the scales).
+A normalized density plot reveals, however, that patients with PJI had, on average, higher predicted probabilities for this adverse event (right panel)."}
 knitr::include_graphics("../graphs/separation.png")
@@
 
-```{r calibrationse, fig.cap = "This figure illustrates internal calibration of the reduced model, used in the Swedish cohort. Predicted probabilities are similar to observed proportions."}
+```{r calibrationse, fig.cap = "This figure illustrates internal calibration of the reduced model, used in the Swedish cohort.
+Predicted probabilities are similar to observed proportions, as indicated by close proximity of the grey 95 % confidence band to the diagonal line, which indicates perfect calibration.
+Proportions of PJI above 10 % were rarely observed and therefore omitted."}
 knitr::include_graphics("../graphs/calibration.png")
@@
   knitr::kable(
-    caption = "Codes identifying infection if recorded in the national patient register within 90 days or 2 years after THA respectively."
+    caption = "Codes identifying PJI if recorded in the national patient register within 90 days after THA."
   )
@@
 knitr::kable(
-  caption = "Nagelkerke's pseudo-coefficient of determination (R2) as an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
+  caption = "Model performance based on the Swedish derivation cohort.
+The area under the receiver operating characteristics curve (AUC) is a measure of discriminatory ability.
+Nagelkerke's pseudo-coefficient of determination (R2) is an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
 The univariable Charlson model had the lowest R2 and was therefore used as reference for the fraction of new information (FNI) added by the other models.")
@@
 BMI = body mass index (under/normal weight as baseline).
-CNS = central nervous system.
-Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
+CNS = central nervous system.'
 )
@@
 We also re-estimated the model coefficients 1,000 times based on the Swedish data.
-Each time, we took a random sample of the same size, $N = 18854$, as was observed in the Danish cohort.
-The 2.5th and 97.5th percentiles were then used to form empirical 95 % confidence intervals for each coefficient ((Tab.
-\@ref(tab:comparecoefs)) and Fig.
+Each time, we took a random sample of the same size ($N = 18,854$) as was observed in the Danish cohort.
+The 2.5th and 97.5th percentiles were then used to form empirical 95 % confidence intervals for each coefficient (Tab.
+\@ref(tab:comparecoefs) and Fig.
 \@ref(fig:coefsspread)).
-It is seen that the estimated effect of the diagnose of "sequelae after childhood hip disease" fall outside this empirical CI.
+The estimated effect of "sequelae after childhood hip disease" fall outside this empirical CI.
 Applying multiplicity correction by Bonferroni or similar, would eliminate this significance.
@@
 compare_se_dk_coefs %>% 
+  rename(`95 % CI` = CI) %>% 
 knitr::kable(
-  caption = ""
+  caption = "Estimated coefficients for the reduced model based on the Swedish derivation cohort (Swedish), as well as re-estimated coefficient values based on the Danish cohort (Danish).
+Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (N = 18,854) as used in the Danish cohort."
 )
@@
 Small coulored dots are empirical estimates from 1000 subsamples based on the Swedish data set, of the same sample size as used in the Danish validation cohort.
-Black dots indicate the outer limits of empirical confidence intervals based on the 2.5th and 97.5th percentiles.
-It is seen that the Swedish and Danish estimates are similar.
-The Danish estimates almost always fall within the 95 % empirical CIs.
-The sole exception is for the rare diagnose group of 'sequelae after childhood hip disease'."}
+Black dots indicate the outer limits of 95 % empirical confidence intervals.
+The Danish estimates almost always fall within those CIs."}
 knitr::include_graphics("../validation/Swedish_model.png")

### 2021-06-23 reports/article.Rmd
kanske sista version innan inskick (delad med NPH)

@@
 ---
-title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
-author: Erik Bülow,^1,2^ Ute Hahn^3^, ~~Ina Trolle Andersen,^3^~~**???** Aurélie Mailhac,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3^ Nils P. Hailer^4^ **TODO DECIDE ORDER!!!**
+title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark"
+author: Erik Bülow,^1,2^ Ina Trolle Andersen,^3^ Ute Hahn,^3,4^  Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
 date: '`r Sys.Date()`'
@@
   C:\\Users\\erik_\\Documents\\library.bib
-csl: the-lancet.csl
+csl: bmj.csl
 output: 
   bookdown::word_document2:
-    reference_docx: article.dotx
+    reference_docx: article.dotm
     toc: false
@@
 2.  Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-3.  Department of Clinical Medicine, Department of Clinical Epidemiology, Aarhus university, Aarhus, Denmark
-4.  Department Surgical Sciences/Orthopaedics, Uppsala University Hospital, Uppsala, Sweden
+3.  Department of Clinical Epidemiology, Aarhus university hospital, Aarhus, Denmark
+4.  Department of Mathematics, Aarhus University
+5.  Department of Clinical Medicine, Aarhus university
+6.  Department Surgical Sciences/Orthopaedics, Uppsala University Hospital, Uppsala, Sweden
+
+- EB: Statistician, MSc, PhD, erik.bulow@gu.se
+- IT: Statistician, MSc, PhD, ita@clin.au.dk
+- UH: Biostatistician, PhD, associate professor, ute@math.au.dk
+- OR: Orthopaedic consultant, PhD, professor, register director, ola.rolfson@gu.se
+- ABP: Orthopaedic consultant, PhD, DMSc, professor, abp@clin.au.dk
+- NPH: Orthopaedic consultant, PhD, professor, nils.Hailer@surgsci.uu.se
 
@@
 
-# Note!
+**Guarantor:** NPH, as the principal investigator, affirms that this manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as planned have been explained.
 
-Most text in the article is automatically generated! Bold text is not! It is thus possible that this text does not correspond to the rest! It must be updated manually!
+**Word count:** 3,583
 
@@
 }
+dkauc <- function(model) {
+  sub("AUC = ", "", get_auc(model, "2", Country = "Denmark"))
+}
 ```
@@
 
-**OJECTIVE:** Previous attempts at developing risk calculators for periprosthetic joint infection (PJI)  after total hip arthroplasty (THA) are hampered by poor predictive performance of the proposed models or a lack of external validation, and no published model is derived on a European population.
-We thus developed a parsimonious model to predict PJI within 90 days after elective THA.
+**OJECTIVE:** We developed a parsimonious prediction model for prosthesis joint infection (PJI) within 90 days after total hip arthroplasty (THA).
 
-**PARTICIPANTS AND METHODS:** This prediction model was derived on `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register and validated on `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register.
-Data on comorbidities and socioeconomic background variables on all patients were obtained from the National Registers.
-Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery.
-We assessed model performance by the area under the receiver operation characteristics curve (AUC), calibration belt plots and the fraction of new information.
+**DESIGN:** Logistic LASSO regression with bootstrap ranking applied to prospectively collected patient data from multiple linked national registers.
 
-**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort and **X.XX %** in Denmark.
-A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and `r sub("^AUC = ", "", get_auc("Reduced model", "2", Country = "Denmark"))` externally.
-These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("ASA")`, or the Charlson `r get_auc("Charlson")`, Elixhauser `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` comorbidity indices.
+**SETTING:** Model derivation and internal (external) validation for all hospitals performing THA in Sweden (Denmark).
 
-**CONCLUSION:** Our parsimonious prediction model for PJI after THA is based on easily accessible clinical data, and it is superior to predictions based on ASA class alone or more complex comorbidity indices.
-Our web-based calculator could simplify the individualized assessment of the risk of developing PJI prior to THA surgery and enable risk-stratified patient management.
-Nonetheless, the remaining unexplained variance indicates that unmeasured factors play an important role in the pathogenesis of PJI.
+**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register with elective THA 2008-2015, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register with THA 2016--September 2018.
+
+**MAIN OUTCOME MEASURE:** PJI as identified either as reason for re-operation, or as recorded by ICD-10 or NOMESCO codes at later hospital visits.
+
+**RESULTS:** Incidence of PJI was `r crude_rate * 100` %  in Sweden and 2.17 % in Denmark.
+A model with the `r coefs_text[["long"]]`, had an `r get_auc("Reduced model (SE)", "2")` in Sweden, and `r sub("^AUC = ", "", get_auc("Reduced model (SE)", "2", Country = "Denmark"))` in Denmark.
+This was superior to traditional models based on ASA class, Charlson, Elixhauser, or the Rx Risk V comorbidity indices.
+Internal calibration was good for predicted probabilities up to 10 %.
+
+**CONCLUSION:** Our model, based on easily accessible data available before THA, had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices.
+It had good calibration and was externally validated in Denmark.
+We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to enable shared decision making by patients and surgeons.
+
+##### PAGE BREAK
+
+# Summary box
+
+## What is already known on this topic
+
+- Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA).
+- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia.
+- None of those models have gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.
+
+## What this study adds
+- Our model, based on easily accessible data available before THA, had superior discriminatory ability compared to more complex comorbidity indices. 
+- Our model had good calibration, was externally validated in Denmark, and can be used in clinical practice by a simple web calculator (https://erikbulow.shinyapps.io/thamortpred/).
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its severity in terms of increased mortality, the required number of re-operations, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly desirable tool in the pre-operative assessment of patients scheduled for THA, but established measures of comorbidity such as the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices cannot be used in this context.
-Although high ASA class is a risk factor associated with development of PJI, this classification is imprecise, and it was never developed for the purpose of predicting this specific complication.[@Haynes1995; @Berbari2012] The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
-The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Several other risk calculators for the risk of PJI after THA have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminative ability or low accuracy of the prediction models, or the absence of external model validation.
-Thus, there is an urgent need for improved prediction models of PJI.
-We therefore developed a prediction model of PJI within 90 days after THA that is based upon easily accessible data that are available in the setting of clinical decision-making.
+Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA.
+Both the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices can be used for risk profiling, but their use is rather challenging.
+Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI [@Haynes1995; @Berbari2012].
+The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation.
+We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA that is based on easily accessible data that are available in the setting of clinical decision-making.
+Furthermore, we aimed to validate the model both internally in Sweden, as well as externally on a Danish national cohort.
 
@@
 The best model was then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).
+We followed the TRIPOD checklist for the reporting of the study.
 
@@
 
-The derivation cohort was obtained from patients recorded in the SHAR `r period` with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then.
-Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on comorbidities for one year prior to index surgery were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
+The derivation cohort was obtained from SHAR `r period`, with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then.
+Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR.
+Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
-The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
-Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
-Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.
+The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020),[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
+Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
+Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded, as were children under the age of 18.
 
@@
 
-Danish THA patients were operated from 2016 until the last of September 2018, and this period was chosen due to availability of data and thus follows after the period used for model derivation in Sweden.
+Danish THA patients were operated from 2016 until the last of September 2018.
+This period was chosen due to availability of ASA and BMI data and possibility for all patients to have 90 days of follow up.
 
-**TODO: ADD INFO (data sources)!**
+All surgery related variables, ASA and BMI were obtained from the DHR.
+The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs.[@Gundtoft2016] All hospitals performing THA in Denmark (40 in 2019) are included in the register.
+The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals.[@Schmidt2015] Data on education and civil status were obtained from Statistics Denmark.[@zotero-2587]
 
-## Definitions of comorbidity and outcomes
+Similar exclusion criteria were applied for diagnosis, ASA class, BMI and bilateral prosthesis.
+Education and type of hospital were not used in the final model and therefore not required for the Danish patients.
+Most people in the Swedish cohort were of age 35 to 99.
+We therefore chose to validate the model for this age group only.
 
-The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+## Definitions of covariates and outcomes
+
+The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery.
+Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
 \@ref(tab:tabcategorization)).
-Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events.
-PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the NPR within 90 days (Tab.
-\@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.
-
-## Model development
-
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
-BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, $\geq 35$: class II-III obesity).
-
-Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance.
-100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
-
-Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term.
-Regression coefficients were ranked based on their absolute values as a measure of variable importance.
-We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model.
-The second change point was then used as a break-point.
-Influential predictors with absolute coefficients value above this point scored a point and the entire process was repeated 100 times, yielding 10,000 bootstrap samples in total.
-Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of numeric variables.
-An even more parsimonious "reduced" model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times was evaluated for comparison.
-Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total).
-The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
-This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals based on the whole cohort.
-
-## Model validation
-
-Each developed model was used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
-Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
-The model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$.
-This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to the baseline model.[@Harrell2018]
-
-The derived model was then externally validated on the Danish cohort, both with the coefficient estimates from the Swedish cohort and with new estimates based on the Danish cohort.
-The 10 simpler models, with adopted coefficient estimates, were also applied to the Danish cohort for comparison.
-Finally, the new coefficient estimates for the derived model, re-estimated with Danish data, were empirically compared to analogues estimates derived from subsamples of the Swedish data with sample sizes to match the Danish cohort (supplementary, Tab \@ref(tab:comparecoefs) and Fig.
-\@ref(fig:coefsspread)).
-
-## Statistical tools
-
-We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder[@Bulow2020b], decoder, furrr, pROC, rms, tidymodels and tidyverse.
-
-## Ethical approval
-
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13) and by **XXX in Denmark**.
-
-# Results
-
-```{r}
-age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
-
-comorb <- function(x) {
-  sprintf("%.0f %% (n = %s)", mean(x) * 100, nmbr(sum(x)))
-}
-```
-
-## Study participants
-
-We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig.
-\@ref(fig:flowchartse)).
-`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of these developed a PJI within 90 days (Tab.
-\@ref(tab:tab1)).
-The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
-
-In addition, `r nmbr(N_dk)` patients **(42.1 % males)** with a mean age of **XXX (SD = XXX)** years were included from DHR as an external validation cohort (Fig.
-\@ref(fig:flowchartdk)).
-**2.2 % (n = 410)** of these developed a PJI within 90 days (Tab.
-\@ref(tab:tab1)).
-The proportion of patients with ASA class III was **18 % (n = 3,334)**; **14 % (n = 2,722)** had at least one comorbidity according to the Charlson comorbidity index, **22 % (n = 4,101)** according to Elixhauser index, and **XXX % (n = XXX)** had medications classified according to Rx Risk V.
-
-**TODO:: Add XXX numbers for DK. Differences in Charlson/Elix might be discussed?**
-
-## Model development and internal validation
-
-Patients with `r excl_factors` were uncommon, why these comorbidities were excluded as potential predictors prior to further modelling.
-ROC-curves (Fig.
-\@ref(fig:rocs) left panel) and AUC-values (Fig.
-\@ref(fig:aucci) left panel) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced model.
-
-```{r}
-CCI_AUC <- sub("AUC = ", "", get_auc("Charlson"))
-CCI_R2 <-
-  tab_new_info %>% 
-  subset(Model == "Charlson", "R2 (%)", drop = TRUE)
-red_FNI <- 
-  tab_new_info %>% 
-  subset(Model == "Reduced model", "FNI (%)", drop = TRUE) %>% 
-  round()
-```
-
-The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text` (Tab.
-\@ref(tab:coefs)).
-This reduced model had an `r get_auc("Reduced model", "2")`, whereas ASA class `r get_auc("ASA")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` had less discriminative ability (Fig.
-\@ref(fig:rocs) left panel and \@ref(fig:aucci) left panel).
-The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2$ (`r CCI_R2`; Tab.
-\@ref(tab:fni)), and compared to this model, the fraction of new information for the reduced model was `r red_FNI` %.
-
-The reduced model had good internal calibration (Fig.
-\@ref(fig:calibrationse)).
-Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for this outcome (Fig.
-\@ref(fig:sep)).
-
-## External validation
-
-The reduced model derived from the Swedish cohort produced a very similar ROC curve when applied to the Danish cohort (Fig.
-\@ref(fig:rocs) right panel), and its AUC value was **0.66 (95% CI: 0.64 to 0.69)**, thus not statistically significantly different from the derivation model.
-This did not change for the model with re-estimated coefficient values (data not shown).
-AUC values for some of the other models estimated on the Danish cohort did differ from the Swedish cohort, with a statistically significantly different estimate obtained for the two models based on the Rx Risk V, but all those models were still less accurate compared to the reduced model (Fig.
-\@ref(fig:aucci) right panel).
-The re-estimated model coefficients of the reduced model were coherent with the estimates based on the Swedish data (supplementary Tab.
-\@ref(tab:comparecoefs)).
-
-# Discussion
-
-## Principal findings
-
-We found that a reduced prediction model based on some easily identifiable clinical conditions, sex, and BMI was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices.
-This model performed better than the established comorbidity measures in terms of AUC, Nagelkerke's $R^2$ and calibration.
-
-## Clinical usage
-
-Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/[1 + \mathrm{exp}(-\hat \alpha - x'\hat \beta)]$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\hat \beta = (\hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k)$ are the estimated intercept and coefficients (Tab.
-\@ref(tab:coefs)) and $x = (x_1, x_2, \ldots, x_k)$ are corresponding patient characteristics.
-For example, `r examples$desc[[1]]` would have a probability of $1/[1 + \mathrm{exp}(6.31 - 60\cdot0.02)] \approx$ `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
-Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
-
-## Model predictors
-
-The predictors in our models were chosen based on a combination of clinical and analytical reasons, but they should not be interpreted as having direct causal effects on the outcome of PJI.
-The majority of the identified predictor variables are, however, associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.
-
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons, and in agreement with previous observational data we found that both AVN and secondary osteoarthritis conferred a nearly two-fold increased risk of PJI in our prediction models.[@Pedersen2010a; @Dale2012] Rheumatoid arthritis as the diagnosis underlying THA surgery is pointed out as a risk factor for PJI,[@Bozic2012; @Inacio2015a] and this predictor variable was also associated with a more then doubled risk of PJI in our model.
-
-Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013] Obesity is identified as a risk factor for PJI in a risk calculator derived on two independent US institutional cohorts, which agrees with our identification of obesity as a predictor variable in both models.[@Tan2018] In accordance with observational studies[@Bozic2012] we found the presence of CNS diseases to be a predictor of PJI, and our broad definition of CNS diseases encompasses both cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease, most of which were associated with a risk increase in the cited study.
-
-Male sex is associated with an increased risk of developing PJI after arthroplasty surgery,[@Dale2012] and sex is included as a predictor variable in several previously developed risk calculators, which agrees with our findings.[@Wingert2016a; @Tan2018] Patients with liver cirrhosis are described to have a more than doubled risk of suffering from PJI within one year after THA, supporting our finding of the presence of liver diseases among our predictor variables.[@Deleuran2015] Taken together, the above cited findings support that the predictor variables in our models seem to be relevant in the context of predicting PJI.
-
-High ASA class is frequently associated with an increased risk of PJI,[@Berbari2012; @Pulido2008; @Namba2012] as is comorbidity in a broader sense, and, more specifically, the presence of cardiovascular comorbidity.[@Ong2009; @Higuera2011] We found that ASA class alone was an insufficient predictor of PJI, and none of the cardiovascular disease categories remained as a predictor variable in our final main model, but ASA class was one of the predictor variables in our model.
-
-Some previously mentioned risk factors for developing PJI were not selected during our model development.
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013] and it is one of the defined comorbidities in the ACS NSQIP Surgical Risk Calculator,[@Wingert2016a] but this comorbidity did not reach statistical significance in our models.
-
-THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
-We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.
-
-Although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
-This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our reduced model.
-
-## Strengths and limitations
-
-Our study is hampered by the limitations inherent to observational research, such as misclassification, and residual confounding due to known confounders that we had no information on, or due to unknown confounders.
-
-Although, not a strict limitation, it should be noted that the aim of prediction is different from traditional statistical inference.
-As such, confidence intervals presented for the model coefficients (Table \@ref(tab:coefs)) were based solely on the final logistic regression model.
-Additional uncertainty introduced by the variable selection process was not incorporated.
-Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variables.
-Also, the individual odds ratios should not be interpreted as effect measures of individual relations since some variables are correlated.
-
-The presence of diabetes was not among the predictor variables selected in our models, although it was almost double as frequent among patients with PJI than among those without (data not shown).
-This is in disagreement with numerous observational studies describing an increased risk of developing PJI after both THA or total knee arthroplasty,[@Pedersen2010; @Maoz2015; @Everhart2013] and diabetes is also found among the risk factors in previously developed risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012] Peri-operative hyperglycaemia is a predictor of PJI after total joint replacement of the hip or knee, and information on laboratory parameters such as morning blood glucose or HBA1c could have improved the predictive power of our models, but we had no access to such information.[@Mraovic2011; @Iorio2012] In analogy, low preoperative haemoglobin is associated with the risk of developing surgical site infection,[@Rasouli2014a] but although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
-Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers may be associated with the risk of developing PJI,[@Gilson2010; @Momohara2011] and intra-articular steroid application may also enhance this risk,[@Cancienne2015] but injections given by physicians would not necessarily be identifiable within a Swedish registry setting.
-Operating times, or, in the case of knee surgery, tourniquet times, are associated with the risk of PJI after hip or knee joint replacement,[@Maoz2015; @Willis-Owen2010] and procedure time is one of the risk factors in the Mayo PJI risk score,[@Berbari2012] but we had no information on this potentially important parameter.
-
-The proportion of explained variation (Nagelkerke's $R^2$ = `r subset(tab_new_info, Model == "Reduced model", "R2 (%)", drop = TRUE)`) was low.
-The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
-In addition, a potential underlying susceptibility to PJI -— as has been described for patients with osteomyelitis -— could be a factor of considerable influence on the outcome, but information on single nucleotide polymorphisms in genes that have been associated with musculoskeletal infections was unavailable to us.
-Such shortcomings could be addressed by future genome-wide association studies.
-
-## Conclusion
-
-Our results indicate that the risk of early PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
-We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
-
-##### Page break
-
-# Contribution of authors
-
-NPH initiated the study.
-OR managed the ethical review board application in Sweden and ABP in Denmark.
-EB developed the statistical model.
-~~ITA~~**???**, UH and AM performed the external validation.
-NPH and EB drafted, and all authors edited and finalized the manuscript.
-
-# Acknowledgement
-
-We would like to thank ...
-
-##### PAGE BREAK
-
-<!-- FIGURES -->
-
-```{r flowchartse, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation"}
-knitr::include_graphics("../graphs/flowchart.png")
-```
-
-```{r flowchartdk, fig.cap = "Flowchart with inclusion criteria and number of patients.
-Data from the Danish Hip Arthroplasty Register were used for external model validation.**TODO: This figure is a screenshot from a PDF.
-Can we have the original figure file with high resolution?**"}
-knitr::include_graphics("../validation/dk_flowchart.png")
-```
-
-##### PAGE BREAK
-
-```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
-The main and reduced models performed almost identical for prediction of PJI within 90 days after surgery.
-They both performed better than all other models (left panel).
-The reduced model performed as well when applied to the Danish validation data set (right panel)."}
-knitr::include_graphics("../graphs/roc.png")
-```
-
-##### PAGE BREAK
-
-```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % confidence intervals (De-Long).
-The reduced model performed no different than the main model on the Swedish data, and both of these models performed better than all other models (left panel).
-Similar models were fitted to the Danish cohort (right panel).
-Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as the reduced model (DK) with coefficient values refitted to the Danish cohort."}
-
-knitr::include_graphics("../graphs/auc_ci.png")
-```
-
-##### PAGE BREAK
-
-```{r sep, fig.cap = "Separation between patients with and without PJI within 90 days versus their predicted probabilities of this event.
-Most patients did not have PJI, wherefore the blue bars dominate the histogram (left panel; note the scales).
-A normalized density plot reveals, however, that patients with PJI had, on average, higher predicted probabilities for this adverse event (right panel)."}
-knitr::include_graphics("../graphs/separation.png")
-```
-
-##### PAGE BREAK
-
-```{r calibrationse, fig.cap = "This figure illustrates internal calibration of the reduced model, used in the Swedish cohort.
-Predicted probabilities are similar to observed proportions, as indicated by close proximity of the grey 95 % confidence band to the diagonal line, which indicates perfect calibration.
-Proportions of PJI above 10 % were rarely observed and therefore omitted."}
-knitr::include_graphics("../graphs/calibration.png")
-```
-
-```{r calibrationdk, fig.cap = "This figure illustrates external calibration of the reduced model, with Swedish coefficient values, used in the Danish cohort.
-Predicted probabilities are similar to observed proportions.
-**TODO: Could EB get an export of the R-object to adapt the graphical presentation?**"}
-knitr::include_graphics("../validation/calibration.tiff")
-```
-
-<!-- Tables -->
-
-##### PAGE BREAK
 
@@
   knitr::kable(
-    caption = "Categorization of comorbidities from the Charlson, Elixhauser and Rx Risk V comorbidities. (CNS = central nervous system.)")
+    caption = "Categorization of comorbidities from the Charlson and Elixhauser  comorbidities. (CNS = central nervous system.)")
 
@@
 
-##### PAGE BREAK
+Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting the model to spurious events.
+
+PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Tab.
+\@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.
 
@@
   knitr::kable(
-    caption = "Codes identifying PJI if recorded in the national patient register within 90 days after THA."
+    caption = "Codes identifying PJI if recorded in the Swedish and Danish National Patient Registers within 90 days after THA."
   )
@@
 
-##### PAGE BREAK
+## Model development
+
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
+BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; BMI $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, and BMI $\geq 35$: class II-III obesity).
+
+Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance.
+100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
+
+Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term.
+Regression coefficients were ranked based on their absolute values as a measure of variable importance.
+We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model.
+The second change point was then used as a break-point.
+Influential predictors with absolute coefficients value above this point scored a point and the entire process was repeated 100 times, yielding 10,000 bootstrap samples in total.
+Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of age.
+An even more parsimonious "reduced" model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times was evaluated for comparison.
+Odds ratios for the final models were estimated with 95 % confidence intervals based on the whole derivation cohort.
+
+## Reference models
+
+Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total).
+The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
+This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al.[@Pratt2018]
+
+## Model validation
+
+The main, reduced and the 10 reference models were used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
+Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
+Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
+The baseline model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$.
+This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline.[@Harrell2018]
+
+The derived model was then externally validated on the Danish cohort, both with the coefficient estimates from the Swedish cohort and with new estimates based on the Danish cohort.
+The 10 reference models, with adopted coefficient estimates, were also applied to the Danish cohort for comparison.
+Finally, the new coefficient estimates for the derived model, re-estimated with Danish data, were empirically compared to analogues estimates derived from subsamples of the Swedish data with sample sizes to match the Danish cohort (supplementary, Tab \@ref(tab:comparecoefs) and Fig.
+\@ref(fig:coefsspread)).
+
+## Statistical tools
+
+We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse.
+We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).
+
+## Patient and Public Involvement
+
+Patient representatives are elected as ordinary members of the steering committees for SHAR and DHR, where the development of shared decision making has been discussed.
+Our web calculator will be available for (potential) patients and the public.
+
+## Ethical approval
+
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13).
+The study was approved by the Danish Data Protection Agency (journal number 2015-57-0002) and recorded at Aarhus University (journal number 2016-051-000001).
+
+# Results
+
+```{r}
+age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
+
+comorb <- function(x) {
+  sprintf("%.0f %% (n = %s)", mean(x) * 100, nmbr(sum(x)))
+}
+```
+
+## Study participants
+
+We included `r nmbr(N_shar)` patients (`r round(mean(df$P_Sex == "Male") * 100)` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig.
+\@ref(fig:flowchartse)).
+`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of these developed a PJI within 90 days (Tab.
+\@ref(tab:tab1)).
+The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
+
+In addition, `r nmbr(N_dk)` patients (42.1 % males) with a mean age of 70 (SD = 11) years were included in the external validation cohort (Fig.
+\@ref(fig:flowchartdk)).
+2.2 % (n = 410) of these developed a PJI within 90 days (Tab.
+\@ref(tab:tab1)).
+The proportion of patients with ASA class III was 18 % (n = 3,334); 14 % (n = 2,722) had at least one comorbidity according to the Charlson comorbidity index, 22 % (n = 4,101) according to Elixhauser index, and 56 % (n = 10,625) had medications classified according to Rx Risk V.
 
@@
 knitr::kable(
-  caption = "Patient characteristics in the model derivation cohort.
-Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (at least 12 years).
+  caption = "Patient characteristics in Sweden 2008-2015 and Denmark 2016 to September 2018.
+Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (more than 12 years).
+Data on hospital levels were not available for the Danish cohort.
+Some additional cells have been intentionally left blank due to small numbers and privacy concerns for the Danish cohort.
 BMI = Body mass index.
@@
 
-##### PAGE BREAK
+## Model development and internal validation
+
+Patients with `r excl_factors` were uncommon, why these comorbidities were excluded as potential predictors prior to further modelling.
+
+Results from the 100 bootstrap replicates for variable selection were mostly coherent.
+Diagnosis, BMI and some comorbidities where selected each time.
+Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Tab.
+\@ref(tab:brlprop)).
 
@@
 
-##### PAGE BREAK
-
-```{r fni}
-tab_new_info %>% 
-  arrange(desc(`R2 (%)`)) %>% 
-knitr::kable(
-  caption = "Model performance based on the Swedish derivation cohort.
-The area under the receiver operating characteristics curve (AUC) is a measure of discriminatory ability.
-Nagelkerke's pseudo-coefficient of determination (R2) is an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
-The univariable Charlson model had the lowest R2 and was therefore used as reference for the fraction of new information (FNI) added by the other models.")
-```
-
-##### PAGE BREAK
+The reduced model (variables with $n \leq 80$ in Tab. \@ref(tab:brlprop)) included `r coefs_text[["short"]]` with estimated coefficient values in Tab. \@ref(tab:coefs).
 
@@
 BMI = body mass index (under/normal weight as baseline).
-CNS = central nervous system.'
+CNS = central nervous system.
+ASA = American Society for Anesthesiologists'
 )
+```
+
+ROC-curves and AUC-values for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced model.
+
+```{r}
+CCI_AUC <- sub("AUC = ", "", get_auc("Charlson"))
+CCI_R2 <-
+  tab_new_info %>% 
+  subset(Model == "Charlson", "R2 (%)", drop = TRUE)
+red_FNI <- 
+  tab_new_info %>% 
+  subset(Model == "Reduced model", "FNI (%)", drop = TRUE) %>% 
+  round()
+```
+
+The reduced model had an `r get_auc("Reduced model (SE)", "2")`, whereas univariable models with ASA class `r get_auc("ASA")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` had less discriminatory abilities (Fig.
+\@ref(fig:rocs) left panel and \@ref(fig:aucci) left panel).
+The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2 = $ `r CCI_R2`, and compared to this model, the fraction of new information for the reduced model was `r red_FNI` % (Tab.
+\@ref(tab:fni)).
+
+```{r fni}
+tab_new_info %>% 
+  arrange(desc(`R2 (%)`)) %>% 
+knitr::kable(
+  caption = "Model performance based on the Swedish derivation cohort.
+The area under the receiver operating characteristics curve (AUC) is a measure of discriminatory ability.
+Nagelkerke's pseudo-coefficient of determination (R2) is an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
+The univariable Charlson model had the lowest R2 and was therefore used as baseline for the fraction of new information (FNI) added by the other models.")
+```
+
+The reduced model had good internal calibration for predictions up to 10 % (Fig.
+\@ref(fig:calibration)).
+Estimated probability density curves for patients with and without PJI were partially overlapping, but patients with PJI had on average higher predicted probabilities for this outcome (Fig.
+\@ref(fig:sep)).
+
+## External validation
+
+The reduced model derived from the Swedish cohort produced a very similar ROC curve when applied to the Danish cohort (Fig.
+\@ref(fig:rocs) right panel), and its AUC value was 0.66 (95% CI: 0.64 to 0.69).
+This did not change for the model with re-estimated coefficient values based on the Danish cohort (0.67, 95 % CI: 0.65 to 0.7).
+AUC values for some of the reference models, estimated on the Danish cohort, did differ from the Swedish cohort, but all those models were less accurate (had broader 95 % CIs) than the reduced model (Fig.
+\@ref(fig:aucci) right panel).
+The re-estimated model coefficients of the reduced model were coherent with the estimates based on the Swedish data (supplementary Tab.
+\@ref(tab:comparecoefs)).
+
+Calibration was good for predictions up to 3 %, with some tendency to over-estimating higher probabilities, which improved after initial re-calibration of the model intercept to the Danish PJI incidence (Fig.
+\@ref(fig:calibration)).
+
+# Discussion
+
+## Principal findings
+
+A model based on `r coefs_text[["short"]]` was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices, with or without adjustments for age and sex, in terms of AUC, Nagelkerke's $R^2$ and calibration.
+
+## Strengths and weaknesses
+
+Our external validation is a strength, which showed that the model works equally well in Denmark, despite some perceived differences between the countries, partially explained by different coding routines in each country (Tab.
+\@ref(tab:tab1)).
+
+Both Sweden and Denmark have strived for mass higher education.[@Isopahkala-Bouret2018] This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education.[@zotero-2598] Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ.
+ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark,[@Schmidt2015] compared to 47 % in the Swedish cohort.
+Serious comorbidities such as cancer, are therefore reported equally in both countries, while for example arterial hypertension, is less documented in Denmark.
+Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register.[@Schmidt2015] The RxRisk V index is a weighted index, which is difficult to compare without further analysis.
+Prescription patterns might differ by indications based on different national guidelines, different subsidization, and different medications sold over-the-counter.
+Also, the Danish, but not the Swedish, national prescription register, includes prescriptions from nursery homes and other long-term care facilities.
+Taken together, the similarities between the countries might be bigger than perceived by Tab.
+\@ref(tab:tab1).
+
+However, the difference in mode of fixation, with a dominance of cemented components in Sweden, is real, and has been described elsewhere.[@Bunyoz2020]
+
+## Strengths and weaknesses in relation to other studies
+
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a] but this comorbidity was not included in our model.
+It nevertheless appeared in `r subset(coefs_selected_tab, variable == "Cancer", "n", drop = TRUE)` of the 100 bootstrap replicates (Tab.
+\@ref(tab:brlprop)), and could have been included in a less parsimonious model.
+Neither diabetes was included in our model, but has been in previous risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012]
+
+Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c,[@Mraovic2011; @Iorio2012] or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model.
+But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
+Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers,[@Gilson2010; @Momohara2011] intra-articular steroid application[@Cancienne2015] or operating time,[@Maoz2015; @Willis-Owen2010; @Berbari2012] are also associated with the risk of PJI after THA.
+However, we chose to only consider factors available before surgery and are thus able to use the model as a tool for shared decision making using the implemented web calculator.
+This is a strength of our model.
+
+We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
+This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig.
+\@ref(fig:aucci) and Tab.
+\@ref(tab:fni)).
+
+## Meaning of the study
+
+Our model can be presented in a simple formula $\hat p = 1/[1 + \mathrm{exp}(-\hat \alpha - x'\hat \beta)]$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\hat \beta = (\hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k)$ are the estimated intercept and coefficients (Tab.
+\@ref(tab:coefs)) and $x = (x_1, x_2, \ldots, x_k)$ are corresponding patient characteristics.
+For example, `r examples$desc[[1]]` would have a probability of $1/[1 + \mathrm{exp}(6.31 - 60\cdot0.02)] \approx$ `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
+Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
+
+It should be noted that the aim of prediction is different from statistical inference.
+As such, confidence intervals presented for the model coefficients (Table \@ref(tab:coefs)) were based solely on the final logistic regression model.
+Additional uncertainty introduced by the variable selection process was not incorporated.
+Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variable.
+Also, the individual odds ratios should not be interpreted as effect measures of individual relations since some variables are correlated.
+
+The selected variables are nevertheless coherent with previous studies.
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons.[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a] Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018] CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI,[@Bozic2012] as have male sex,[@Dale2012; @Wingert2016a; @Tan2018] liver disease [@Deleuran2015] and ASA class.[@Berbari2012; @Pulido2008; @Namba2012]
+
+THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
+We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.
+
+## Unanswered questions and future research
+
+The proportion of explained variation (Nagelkerke's $R^2$ = `r subset(tab_new_info, Model == "Reduced model", "R2 (%)", drop = TRUE)` %) was low.
+The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
+In addition, a potential underlying susceptibility to PJI -— as has been described for patients with osteomyelitis -— could be a factor of considerable influence on the outcome, but information on single nucleotide polymorphisms in genes that have been associated with musculoskeletal infections was unavailable to us.
+Such shortcomings could be addressed by future genome-wide association studies.
+
+## Conclusion
+
+Our results indicate that the risk of early PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
+We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
+
+##### Page break
+
+# Contribution of authors
+
+NPH initiated the study.
+OR managed the ethical review board application in Sweden and ABP in Denmark.
+EB developed the statistical model.
+IT and UH performed the external validation.
+NPH and EB drafted, and all authors edited and finalized the manuscript.
+
+# Acknowledgement
+
+We would like to thank Aurelie Mailhac, statistician at the Department of Clinical Epidemiology at Aarhus university, for help with the external validation.
+
+# Data sharing
+
+Access to the model derivation cohort is granted after permission from the Västra Götaland Region, as the central personal data controller agency for the SHAR, in accordance with the European General Data Protection Regulation (GDPR), as well as Swedish laws and regulations (https://registercentrum.se/vara-tjaenster/datauttag/p/B1eeV66QE).
+
+# Disclosure
+
+All authors have completed the ICMJE uniform disclosure form at http://icmje.org/downloads/coi_disclosure.docx and declare: no support from any organisation for the submitted work; OR is principal investigator for research partly founded by Pfizer; OR is register director of the Swedish Arthroplasty Register (SAR) and board member of the International Society of Arthroplasty Registries; NPH has received institutional support from Waldemar Link GmbH, and Zimmer Biomet; NPH is president of the Nordic Arthroplasty Register Association, member of the steering committee of SAR, and Co-Editor of Acta Orthopaedica; no other relationships or activities that could appear to have influenced the submitted work.
+
+##### PAGE BREAK
+
+<!-- FIGURES -->
+
+```{r}
+# MOve figure file to folder for easy access
+fc <- 0 # Figure count
+
+cp_fig <- function(file) {
+  fc <<- fc + 1
+  x <- file.copy(file, paste0("bmj_figs/", fc, ".pdf"), TRUE)
+}
+```
+
+```{r flowchartse, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation"}
+
+cp_fig("../graphs/flowchart.pdf")
+knitr::include_graphics("../graphs/flowchart.png")
+```
+
+```{r flowchartdk, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Danish Hip Arthroplasty Register were used for external model validation."}
+
+cp_fig("../validation/flowchartdk_detailed.pdf")
+knitr::include_graphics("../validation/dk_flowchart.png")
+```
+
+##### PAGE BREAK
+
+```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
+The main and reduced models performed almost identical for prediction of PJI within 90 days after surgery.
+They both performed better than all other models (left panel).
+The reduced model performed as well when applied to the Danish validation cohort (right panel)."}
+
+cp_fig("../graphs/roc.pdf")
+knitr::include_graphics("../graphs/roc.png")
+```
+
+##### PAGE BREAK
+
+```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % confidence intervals (De-Long).
+The reduced model performed no different than the main model on the Swedish data, and both of these models performed better than all other models (left panel).
+Similar models were fitted to the Danish cohort (right panel).
+Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as the reduced model (DK) with coefficient values refitted to the Danish cohort."}
+
+cp_fig("../graphs/auc_ci.pdf")
+knitr::include_graphics("../graphs/auc_ci.png")
+```
+
+##### PAGE BREAK
+
+```{r sep, fig.cap = "Separation between patients with and without PJI within 90 days versus their predicted probabilities of this event.
+Most patients did not have PJI, wherefore the pink bars dominate the histogram (left panel; note the scales).
+A normalized density plot reveals, however, that patients with PJI had, on average, higher predicted probabilities for this adverse event (right panel)."}
+
+cp_fig("../graphs/separation.pdf")
+knitr::include_graphics("../graphs/separation.png")
+```
+
+##### PAGE BREAK
+
+```{r calibration, fig.cap = "Calibration for the reduced model.
+Proportions of PJI above 10 % were rarely observed and therefore omitted.
+Internal calibration (green 95 % confidence band) is good; predicted probabilities are similar to observed proportions, as indicated by close proximity to the diagonal line.
+External calibration with the same model (pink) indicates some over-estimation and less accuracy.
+This is expected due to smaller sample size in the Danish cohort, and due to national differences.
+Calibration improves after re-calibration of the model intercept to better resamble the Danish base incidence (blue)."}
+
+cp_fig("../graphs/calibration.pdf")
+knitr::include_graphics("../graphs/calibration.png")
 ```
@@
 The coefficient values of the reduced model were also re-fitted to the Danish cohort for comparison.
-We compared the estimates from the Swedish and Danish cohort and found most values to be of similar magnitude.
+We compared the estimates from the Swedish and Danish cohorts and found most values to be of similar magnitude.
 The only coefficient with reversed direction was for patients with a diagnose of "sequelae after childhood hip disease".
@@
   caption = "Estimated coefficients for the reduced model based on the Swedish derivation cohort (Swedish), as well as re-estimated coefficient values based on the Danish cohort (Danish).
-Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (N = 18,854) as used in the Danish cohort."
+Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (N = 18,854) as used in the Danish cohort.
+The Danish estimates fall within most of the CIs, indicating no support to reject the null hypothesis of no differneces between the countries."
 )
@@
 
-```{r coefsspread, fig.cap = "Estimated coefficient values for the reduced model.
-Red triangles indicate estimates from the Swedish derivation data set.
-Red circles indicate re-estimated coefficient values for the same model based on the Danish validation cohort.
-Small coulored dots are empirical estimates from 1000 subsamples based on the Swedish data set, of the same sample size as used in the Danish validation cohort.
+```{r coefsspread, fig.cap = "Estimated coefficient values for the reduced model applied to the Swedish (red triangles) and Danish cohorts (red dots).
+Small coloured dots are empirical estimates from 1000 subsamples based on the Swedish data set, of the same sample size as used in the Danish validation cohort (outliers not shown).
 Black dots indicate the outer limits of 95 % empirical confidence intervals.
 The Danish estimates almost always fall within those CIs."}
+
+cp_fig("../validation/Swedish_model.pdf")
 knitr::include_graphics("../validation/Swedish_model.png")
-```
-
-## Stratified separation
-
-**TODO: If we decide to include those figures, could we get the Danish ggplot2-objects as well? Would you also like to add some descriptive text in that case Ute?**
-
-```{r sepsex, fig.cap = "Predicted probability by the reduced model, stratified by outcome and sex."}
-knitr::include_graphics("../graphs/separation_strata_sex.png")
-```
-
-```{r sepasa, fig.cap = "Predicted probability by the reduced model, stratified by outcome and ASA class."}
-knitr::include_graphics("../graphs/separation_strata_asa.png")
-```
-
-```{r sepbmi, fig.cap = "Predicted probability by the reduced model, stratified by outcome and Body Mass index (BMI)."}
-knitr::include_graphics("../graphs/separation_strata_bmi.png")
-```
-
-```{r sepdiagrp, fig.cap = "Predicted probability by the reduced model, stratified by outcome and diagnose group."}
-knitr::include_graphics("../graphs/separation_strata_diagrp.png")
 ```

### 2021-06-23 lib/clean_names.R

@@ -9,7 +9,7 @@ clean_names <- function(x, firstupper = TRUE, lvls = TRUE) {
   {gsub("cns",     "CNS",         .)} %>%
   {gsub("\\.",     " ",           .)} %>%
   {gsub("  ",      " ",           .)} %>%
-  {gsub("AVN",     "\\(AVN\\)",   .)} %>%
+  {gsub(" AVN$",   "\\(AVN\\)",   .)} %>%
   {gsub("Rheumatidisease", "Rheumatic disease",   .)} %>%
   {paste0(toupper(substr(., 1, 1)), substring(., 2))} %>%
   trimws()

### 2021-06-23 munge/00-filter.R

@@ -56,7 +56,7 @@ filters <-
 
     tibble(
       step = 3,
-      excl = "Diagnosis: Fracture/tumour/unspecified/unknown.",
+      excl = "Diagnosis: Fracture/tumour/unspecified/unknown",
       incl = "Elective THA",
       expr = list(quo(!P_DiaGrp %in% c(
         "Akut trauma, höftfraktur",
@@ -193,7 +193,7 @@ nodes <-
   ) %>%
   mutate(
     shape     = "rectangle",
-    width     = if_else(startsWith(name, "incl"), 1.75, 3.25),
+    width     = if_else(startsWith(name, "incl"), 1.75, 3.5),
     color     = "Black",
     fillcolor = "White",
     fontcolor = "Black"
@@ -239,5 +239,6 @@ graph <-
   add_global_graph_attrs("fixedsize", "FALSE", "node") # Nice edge height
 
 export_graph(graph, "graphs/flowchart.png", "png", width = 1024)
+export_graph(graph, "graphs/flowchart.pdf", "pdf")
 # render_graph(graph)
 

### 2021-06-23 src/compare_NPR_se_dk.R

@@ -1,0 +1,26 @@
+suppressMessages({library(ProjectTemplate); load.project()})
+
+load("cache/df_icd10_orig.RData")
+
+# VI har mer komorbiditet i Sv jmfrt med DK.
+# Enligt [@Schmidt2015]:
+# > Since the adaption of the tenth revision of the International Classification
+# of Diseases (ICD-10) in 1994, 23% of hospital contacts have had one or more
+# secondary diagnoses recorded. The median number of secondary diagnoses per
+# contact in this period was 1 (interquartile range: 1-2 diagnoses).
+
+df_icd10 <-
+  df_icd10_orig %>%
+  mutate(hospital = as.Date(code_date, origin = "1970-01-01")) %>%
+  select(LopNr, code, hospital) %>%
+  as.data.table(key = c("LopNr", "code", "hospital"))
+
+# Andel av de som har sjukhusvistelse som har fler än en diagnos
+df_icd10[, .N, .(LopNr, hospital)][, .(
+  prop   = mean(N > 1),
+  median = median(N),
+  q1     = quantile(N, .25),
+  q3     = quantile(N, .75)
+)]
+# 47 % > 23 % DK
+# median och IQR samma; 1, 1-2

### 2021-06-23 src/fig_auc.R

@@ -13,12 +13,24 @@ data_auc_ci <-
   select(country, Model, starts_with("AUC")) %>%
   mutate(
     Model  = gsub("Reduced model$", "Reduced model (SE)", Model),
-    Model  = reorder(Model, AUC_est),
     ci     = sprintf("95 %% CI: %.2f to %.2f", AUC_lo, AUC_hi),
     text1  = sprintf("(AUC = %.2f, %s)", AUC_est, ci),
     text2  = sprintf("AUC = %.2f (%s)", AUC_est, ci),
     country = factor(country, c("se", "dk"), c("Sweden", "Denmark"))
   )
+
+
+# Levels based on the Swedish rank order of the models
+# lvls <- levels(with(filter(data_auc_ci, country == "se"), reorder(Model, AUC_est)))
+lvls <-
+  levels(
+    with(
+      filter(data_auc_ci, country == "Sweden" | Model == "Reduced model (DK)"),
+      reorder(Model, AUC_est, order = TRUE)
+    )
+  )
+data_auc_ci$Model <- factor(data_auc_ci$Model, lvls, ordered = TRUE)
+
 
 cache("data_auc_ci")
 
@@ -59,5 +71,6 @@ ggplot(data_auc_ci, aes(Model, AUC_est)) +
   facet_wrap(~ country, scales = "free_x")
 
 ggsave("graphs/auc_ci.png", width = 15, height = 10, units = "cm")
+ggsave("graphs/auc_ci.pdf", width = 15, height = 10, units = "cm")
 
 options(digits = digs$digits)

### 2021-06-23 src/fig_calibration.R

@@ -10,9 +10,9 @@ get_beltdata <- function(cb) {
   )
 }
 
-df_calibration <-
+df_calibration_se <-
   models %>%
-  filter(Model == "Reduced model") %>%
+  filter(Model == "Reduced model", country == "se") %>%
   mutate(
     cal_belt = furrr::future_map(
       obspred, ~ givitiR::givitiCalibrationBelt(
@@ -24,23 +24,36 @@ df_calibration <-
   select(Model, cal_belt) %>%
   unnest(cal_belt)
 
+
+# Danish ------------------------------------------------------------------
+
+source("validation/calibration_DK.txt")
+
+
 # Make figure -------------------------------------------------------------
-calplot <- function(xlim = c(0, 0.1), ylim = c(0, 0.1)) {
-  ggplot(df_calibration, aes(x, ymin = L, ymax = U)) +
-  geom_ribbon(alpha = .3) +
-  geom_abline(aes(intercept = 0, slope = 1)) +
-  theme_minimal(15) +
-  scale_x_continuous(labels = scales::percent_format(1)) +
-  scale_y_continuous(labels = scales::percent_format(1)) +
-  coord_cartesian(xlim, ylim) +
-  xlab("Predicted probability") +
-  ylab("Observed proportion") +
-  theme(
-    legend.position = c(1, 0),
-    legend.justification = c(1, 0),
-    legend.title = element_blank()
-  )
-}
+bind_rows(
+  "Internal (SE)" = df_calibration_se,
+  "External  (DK)" = get_beltdata(calibration),
+  "Re-calibrated (DK)" = get_beltdata(calibration2),
+  .id = "country"
+) %>%
+ggplot(aes(x, ymin = L, ymax = U, fill = country)) +
+geom_ribbon(alpha = .5) +
+geom_abline(aes(intercept = 0, slope = 1)) +
+theme_minimal(15) +
+scale_x_continuous(labels = scales::percent_format(1), limits = c(0, .1), breaks = seq(0, .1, .02)) +
+scale_y_continuous(labels = scales::percent_format(1), limits = c(0, .1), breaks = seq(0, .1, .02)) +
+xlab("Predicted probability") +
+ylab("Observed proportion") +
+theme(
+  legend.position = c(1, 0),
+  legend.justification = c(1, 0),
+  legend.title = element_blank(),
+  panel.grid.minor = element_blank()
+)
 
 
[…]

### 2021-06-23 src/fig_roc.R

@@ -50,3 +50,4 @@ ggdata %>%
 
 
 ggsave("graphs/roc.png", height = 10, width = 15, units = "cm", dpi = 900)
+ggsave("graphs/roc.pdf", height = 10, width = 17, units = "cm")

### 2021-06-23 src/fig_separation.R

@@ -83,3 +83,9 @@ ggsave(
   gridExtra::grid.arrange(fig_separation_hist, fig_separation_density, nrow = 1),
   height = 10, width = 20, units = "cm"
 )
+
+ggsave(
+  "graphs/separation.pdf",
+  gridExtra::grid.arrange(fig_separation_hist, fig_separation_density, nrow = 1),
+  height = 10, width = 22, units = "cm"
+)

### 2021-06-23 src/tab_coefs.R

@@ -68,10 +68,10 @@ coefs_text <-
     variable = tolower(variable),
     variable = case_when(
       variable == "bmi" ~ "body mass index (BMI)",
-      variable == "diagnosis" ~ "the underlaying diagnosis for THA",
+      variable == "diagnosis" ~ "the underlying diagnosis for THA",
       # variable == "sex" ~ "gender",
       variable == "asa class" ~ "American Society for Anesthesiologists (ASA) class",
-      variable == "cns disease" ~ "CNS disease",
+      variable == "cns disease" ~ "central nervous system (CNS) disease",
       variable == "pancreatiinsufficiency" ~ "pancreatic insufficiency",
       variable == "rheumatidisease" ~ "rheumatic disease",
       variable == "lung airways disease" ~ "lung and airways disease",
@@ -94,6 +94,15 @@ coefs_text <-
   arrange(!ordinary) %>%
   summarise(text = paste(text, collapse = ", and ")) %>%
   deframe()
+
+coefs_text <- c(
+  long = coefs_text,
+  short =
+    gsub("\\(|\\)|body mass index|American Society for Anesthesiologists|central nervous system|", "", coefs_text) %>%
+    {gsub("  ", " ", .)}
+)
+
+
 
 cache("coefs_text")
 

### 2021-06-23 src/tab_define_groups.R

@@ -16,6 +16,7 @@ tab_categorization <-
 
   # To test without Rx Risk V
   select(-Rx) %>%
-  filter(!(is.na(Charlson) & is.na(Elixhauser)))
+  filter(!(is.na(Charlson) & is.na(Elixhauser))) %>%
+  mutate(across(c(Charlson, Elixhauser), ~ gsub("Aids/hiv", "AIDS/HIV", .)))
 
 cache("tab_categorization")

### 2021-06-23 src/tab_table1.R

@@ -107,6 +107,8 @@ dk1 <-
   docxtractr::read_docx("validation/DK_table1_20210325.docx") %>%
   docxtractr::docx_extract_tbl()
 
+
+
 names(dk1) <- c("what", "No_PJI_N", "No_PJI_prop", "PJI_N", "PJI_prop", "Tot_N", "Tot_prop")
 dk1 <- dk1[-(1:2), ] # Remove headers and age (which should not be categorized)
 
@@ -116,10 +118,14 @@ dk1 <- mutate(dk1, across(-what, as.numeric))
 # Identifiera namn på faktorer till skillnad från levels
 dk1$factors <- dk1$what %in% dk1$what[which(is.na(dk1$Tot_N)) -1]
 
-
+# Save empty row for Elixhauser == 0
+elix0 <- filter(dk1, what == "0") %>% slice(1)
 
 dk2 <-
   dk1 %>%
+  distinct() %>%
+  # 0-raden för Elixhauser försvinner och behöver läggas tillbaka
+  add_row(elix0, .after = which(.$what == "Elixhauser")[1]) %>%
   # FLytta ner icke-NA-värden till resp level
   mutate(across(c(-what, -factors), zoo::na.locf)) %>%
   # Ta bort värdena från faktor-rubriken
@@ -175,11 +181,11 @@ dk2 <-
 # Slå samman --------------------------------------------------------------
 
 # Check that no levels differ
-setdiff(table1$what, dk2$what)
-setdiff(table1$level, dk2$level)
+setdiff(se_table1$what, dk2$what)
+setdiff(se_table1$level, dk2$level)
 
-setdiff(dk2$what, table1$what)
-setdiff(dk2$level, table1$level)
+setdiff(dk2$what, se_table1$what)
+setdiff(dk2$level, se_table1$level)
 
 table1 <-
   left_join(se_table1, dk2) %>%

### 2021-06-23 validation/subsample_SE_data.R

@@ -78,7 +78,7 @@ plot_data <- function(results) {
       !grepl("Intercept", coef),
     ) %>%
     pivot_longer(-coef) %>%
-    filter(value > -5) %>%
+    # filter(value > -5) %>%
     mutate(
       coef = gsub("coefficients\\.", "", coef),
       coef = clean_names(coef),
@@ -95,22 +95,29 @@ ci_data <- function(d) {
 plot_help <- function(d, ci, title) {
   d %>%
     ggplot(aes(coef, value, color = coef)) +
-    geom_point(position = position_dodge2(1), alpha = .33, show.legend = FALSE) +
-    geom_boxplot(alpha = .5, size = 1, show.legend = FALSE) +
-    geom_point(aes(shape = orig), data = cfs, size = 5, alpha = 1, color = "red") +
-    geom_point(data = ci, size = 5, alpha = 1, color = "black") +
-    geom_hline(aes(yintercept = 0), linetype = "dotted", color = "orange", size = 3) +
-    coord_flip() +
+    geom_point(position = position_dodge2(1), size = .5, alpha = .1, show.legend = FALSE) +
+    # geom_boxplot(alpha = .5, size = 1, show.legend = FALSE) +
+    geom_point(aes(shape = orig), data = cfs, size = 2, alpha = 1, color = "red") +
+    geom_point(data = ci, size = 2, alpha = 1, color = "black") +
+    geom_hline(aes(yintercept = 0), linetype = "dotted", color = "orange") +
+    coord_flip(, c(-1, 2)) +
     theme_minimal() +
-    theme(legend.position = c(0, 1), legend.justification = c(0, 1))
-    ggtitle(title)
+    theme(
+      legend.position = c(0, 1),
+      legend.justification = c(0, 1),
+      legend.title = element_blank(),
+      axis.title.y = element_blank()
+    ) +
+    ylab("Estimated coefficient values")
+    # ggtitle(title)
 }
 
 plot_result <- function(result, title) {
   d  <- plot_data(result)
   ci <- ci_data(d)
   p  <- plot_help(d, ci, title)
-  ggsave(paste0(title, ".png"), height = 20, width = 40, units = "cm")
+  ggsave(paste0(title, ".png"), height = 10, width = 20, units = "cm")
+  ggsave(paste0(title, ".pdf"), height = 10, width = 20, units = "cm")
   list(d = d, ci = ci, p = p)
 }
 

### 2021-06-29 reports/article.Rmd
changes from ITA

@@
 title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark"
-author: Erik Bülow,^1,2^ Ina Trolle Andersen,^3^ Ute Hahn,^3,4^  Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
+author: Erik Bülow,^1,2^ Ute Hahn,^3,4^ Ina Trolle Andersen,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
 date: '`r Sys.Date()`'

### 2021-06-30 reports/article.Rmd
with comments from NPH

@@
 - EB: Statistician, MSc, PhD, erik.bulow@gu.se
-- IT: Statistician, MSc, PhD, ita@clin.au.dk
+- ITA: Statistician, MSc, PhD, ita@clin.au.dk
 - UH: Biostatistician, PhD, associate professor, ute@math.au.dk
@@
 
-**Word count:** 3,583
+**Word count:** XXX
 
@@
 
-**OJECTIVE:** We developed a parsimonious prediction model for prosthesis joint infection (PJI) within 90 days after total hip arthroplasty (THA).
+**OJECTIVE:** To develop a parsimonious risk prediction model for periprosthetic joint infection (PJI) within 90 days after total hip arthroplasty (THA).
 
-**DESIGN:** Logistic LASSO regression with bootstrap ranking applied to prospectively collected patient data from multiple linked national registers.
+**DESIGN:** Cohort study.
 
-**SETTING:** Model derivation and internal (external) validation for all hospitals performing THA in Sweden (Denmark).
+**SETTING:** Register data from Sweden for model derivation and (from Denmark for external) validation.
 
-**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register with elective THA 2008-2015, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register with THA 2016--September 2018.
+**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register operated with elective THA.
 
-**MAIN OUTCOME MEASURE:** PJI as identified either as reason for re-operation, or as recorded by ICD-10 or NOMESCO codes at later hospital visits.
+**MAIN OUTCOME MEASURE:** Logistic LASSO regression with bootstrap ranking was used to develop a risk prediction model for PJI within 90 days based on the Swedish cohort, followed by external validation on the Danish cohort.
 
 **RESULTS:** Incidence of PJI was `r crude_rate * 100` %  in Sweden and 2.17 % in Denmark.
-A model with the `r coefs_text[["long"]]`, had an `r get_auc("Reduced model (SE)", "2")` in Sweden, and `r sub("^AUC = ", "", get_auc("Reduced model (SE)", "2", Country = "Denmark"))` in Denmark.
+A model with the `r coefs_text[["abstract"]]` had an area under the curve (AUC) of `r gsub("AUC = ", "", get_auc("Reduced model (SE)", "2"))` in Sweden and `r sub("^AUC = ", "", get_auc("Reduced model (SE)", "2", Country = "Denmark"))` in Denmark.
 This was superior to traditional models based on ASA class, Charlson, Elixhauser, or the Rx Risk V comorbidity indices.
@@
 
-**CONCLUSION:** Our model, based on easily accessible data available before THA, had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices.
-It had good calibration and was externally validated in Denmark.
-We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to enable shared decision making by patients and surgeons.
+**CONCLUSION:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated.
+The model had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices and had good calibration.
+We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to facilitate shared decision making by patients and surgeons.
 
@@
 - Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA).
-- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia.
-- None of those models have gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.
+- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia but none of those models has gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.
 
 ## What this study adds
-- Our model, based on easily accessible data available before THA, had superior discriminatory ability compared to more complex comorbidity indices. 
-- Our model had good calibration, was externally validated in Denmark, and can be used in clinical practice by a simple web calculator (https://erikbulow.shinyapps.io/thamortpred/).
+- Our prediction model was derived on a Swedish cohort and externally validated on a Danish cohort
+- Our model is based on easily accessible data that are available before THA
+- Compared to more complex comorbidity indices or to ASA class, our model had superior discriminatory ability and good calibration
+- We provide a simple web calculator for easy use in clinical practice (https://erikbulow.shinyapps.io/thamortpred/).
 
@@
 Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
-Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
+Data on marital status and the educational level of patients were obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
 The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020),[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
@@
 \@ref(fig:rocs) left panel and \@ref(fig:aucci) left panel).
-The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2 = $ `r CCI_R2`, and compared to this model, the fraction of new information for the reduced model was `r red_FNI` % (Tab.
+The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2 =$ `r CCI_R2`, and compared to this model, the fraction of new information for the reduced model was `r red_FNI` % (Tab.
 \@ref(tab:fni)).
@@
 Also, the Danish, but not the Swedish, national prescription register, includes prescriptions from nursery homes and other long-term care facilities.
-Taken together, the similarities between the countries might be bigger than perceived by Tab.
+Taken together, the similarities between the countries might be more pronounced than indicated by Tab.
 \@ref(tab:tab1).
 
-However, the difference in mode of fixation, with a dominance of cemented components in Sweden, is real, and has been described elsewhere.[@Bunyoz2020]
+However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere.[@Bunyoz2020]
 
@@
 EB developed the statistical model.
-IT and UH performed the external validation.
+UH and ITA performed the external validation.
 NPH and EB drafted, and all authors edited and finalized the manuscript.

### 2021-06-30 src/tab_coefs.R

@@ -99,7 +99,8 @@ coefs_text <- c(
   long = coefs_text,
   short =
     gsub("\\(|\\)|body mass index|American Society for Anesthesiologists|central nervous system|", "", coefs_text) %>%
-    {gsub("  ", " ", .)}
+    {gsub("  ", " ", .)},
+  abstract = gsub("and the presence of [A-Za-z ,\\(\\)]*$", "and the presence of five defined comorbidities", coefs_text)
 )
 
 

### 2021-10-15 reports/article.Rmd
Submission to Osteoarthritis & Cartilage. Fix #61.

@@
   C:\\Users\\erik_\\Documents\\library.bib
-csl: bmj.csl
+csl: oac.csl
 output: 
@@
 
-**Word count:** XXX
+**Word count:** XXX **(MAX 4000!!!)**
 
@@
 
-**DESIGN:** Cohort study.
-
-**SETTING:** Register data from Sweden for model derivation and (from Denmark for external) validation.
-
-**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register operated with elective THA.
-
-**MAIN OUTCOME MEASURE:** Logistic LASSO regression with bootstrap ranking was used to develop a risk prediction model for PJI within 90 days based on the Swedish cohort, followed by external validation on the Danish cohort.
+**DESIGN:** Cohort study with register data.
+`r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register used for model derivation and internal validation, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register used for external validation.
+Logistic LASSO regression with bootstrap ranking was used for model development.
 
@@
 
-**CONCLUSION:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated.
+**CONCLUSIONS:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated.
 The model had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices and had good calibration.
@@
 
-##### PAGE BREAK
+# Key words
 
-# Summary box
+Total hip arthroplasty, prosthesis joint infection, prediction, model development, model validation
 
-## What is already known on this topic
+# Running title
 
-- Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA).
-- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia but none of those models has gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.
-
-## What this study adds
-- Our prediction model was derived on a Swedish cohort and externally validated on a Danish cohort
-- Our model is based on easily accessible data that are available before THA
-- Compared to more complex comorbidity indices or to ASA class, our model had superior discriminatory ability and good calibration
-- We provide a simple web calculator for easy use in clinical practice (https://erikbulow.shinyapps.io/thamortpred/).
+Prediction model for PJI after THA
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA.
+Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life[@Kapadia2016].
+Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors[@Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015].
+Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA.
 Both the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices can be used for risk profiling, but their use is rather challenging.
-Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI [@Haynes1995; @Berbari2012].
-The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
-The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation.
+Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI[ @Berbari2012].
+The Charlson and Elixhauser comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
+The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification[@Pratt2018].
+Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first[@Edelstein2015; @Wingert2016a].
+Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance[@Wingert2016a; @Bozic2013a; @Inacio2015a].
+This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation.
 We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA that is based on easily accessible data that are available in the setting of clinical decision-making.
@@
 The derivation cohort was obtained from SHAR `r period`, with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then.
-Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR.
+Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration[@Ludvigsson2009; @Cnudde2016].
+The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR.
 Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
 Data on marital status and the educational level of patients were obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
-The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020),[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
-Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
+The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020)[@Karrholm2018], the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.
+Death dates were linked from the national population register.
+Only the last operated hip was considered for patients with bilateral THA[@Bulow2020].
+Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
 Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded, as were children under the age of 18.
@@
 All surgery related variables, ASA and BMI were obtained from the DHR.
-The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs.[@Gundtoft2016] All hospitals performing THA in Denmark (40 in 2019) are included in the register.
-The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals.[@Schmidt2015] Data on education and civil status were obtained from Statistics Denmark.[@zotero-2587]
+The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs[@Gundtoft2016].
+All hospitals performing THA in Denmark (40 in 2019) are included in the register.
+The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals[@Schmidt2015].
+Data on education and civil status were obtained from Statistics Denmark[@zotero-2587].
 
@@
 
-The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery.
-Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
+The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each[@Quan2005].
+Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery.
+Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Supplementary Tab.
 \@ref(tab:tabcategorization)).
-
-```{r tabcategorization}
-load("../cache/tab_categorization.RData")
-
-tab_categorization %>%
-  ungroup() %>% 
-  mutate(
-    `Comorbidities by groups` = 
-      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
-  ) %>% 
-  knitr::kable(
-    caption = "Categorization of comorbidities from the Charlson and Elixhauser  comorbidities. (CNS = central nervous system.)")
-
-```
 
@@
 
-PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Tab.
+PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Supplementary Tab.
 \@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.
-
-```{r infcodes}
-library(coder)
-
-icd10_codes <- 
-  summary(
-      hip_ae_hailer, 
-      "icd10se", 
-      cc_args = list(regex = "icd10", stop = TRUE)
-    )$codes_vct$Infection %>% 
-    paste(collapse = ", ")
-
-nomesco_codes <- 
-  summary(
-      hip_ae_hailer, 
-      "kva", 
-      cc_args = list(regex = "kva", stop = TRUE)
-    )$codes_vct$Infection %>% 
-    paste(collapse = ", ")
-
-tibble(
-  classification = c("ICD-10", "NOMESCO"),
-  codes = c(icd10_codes, nomesco_codes)
-) %>% 
-  knitr::kable(
-    caption = "Codes identifying PJI if recorded in the Swedish and Danish National Patient Registers within 90 days after THA."
-  )
-
-```
 
@@
 
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
+Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO)[@Guo2015; @Baranowski2020; @Garland2021].
+Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
 BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; BMI $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, and BMI $\geq 35$: class II-III obesity).
@@
 Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance.
-100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
+100 bootstrap samples were then drawn from the observed data set.
 
@@
 The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
-This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al.[@Pratt2018]
+This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al[@Pratt2018].
 
@@
 The main, reduced and the 10 reference models were used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
-Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
+Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals.
+The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
+Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration[@Nattino2016].
+Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
 The baseline model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$.
-This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline.[@Harrell2018]
+This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline[@Harrell2018].
 
@@
 
-We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse.
+We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder[@Bulow2020b], decoder, furrr, pROC, rms, tidymodels and tidyverse.
 We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).
@@
 
-```{r tab1, warning=FALSE}
-source("../lib/clean_names.R")
-
-table1 %>% 
-  mutate(
-    what = replace(what, duplicated(what), "")
-  ) %>% 
-knitr::kable(
-  caption = "Patient characteristics in Sweden 2008-2015 and Denmark 2016 to September 2018.
-Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (more than 12 years).
-Data on hospital levels were not available for the Danish cohort.
-Some additional cells have been intentionally left blank due to small numbers and privacy concerns for the Danish cohort.
-BMI = Body mass index.
-ASA class = American Society for Anaesthesiologists classification.
-CNS = central nervous system.")
-```
-
 ## Model development and internal validation
@@
 Diagnosis, BMI and some comorbidities where selected each time.
-Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Tab.
+Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Supplementary Tab.
 \@ref(tab:brlprop)).
 
-```{r brlprop}
-coefs_selected_tab %>% 
-knitr::kable(
-  caption = "Variables selected by the bootstrap ranking procedure.
-Variables selected at least 10 out of 100 times were used in the main model.
-Variables chosen at least 80 times were kept in the reduced model as well.
-(BMI = body mass index.
-ASA class = American Society for Anaesthesiologists classification.
-CNS = central nervous system.)")
-```
-
 The reduced model (variables with $n \leq 80$ in Tab. \@ref(tab:brlprop)) included `r coefs_text[["short"]]` with estimated coefficient values in Tab. \@ref(tab:coefs).
-
-```{r coefs}
-tbl_coefs_present %>% 
-  mutate(
-    beta = round(beta, 2)
-  ) %>% 
-knitr::kable(
-  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models.
-BMI = body mass index (under/normal weight as baseline).
-CNS = central nervous system.
-ASA = American Society for Anesthesiologists'
-)
-```
 
@@
 \@ref(tab:fni)).
-
-```{r fni}
-tab_new_info %>% 
-  arrange(desc(`R2 (%)`)) %>% 
-knitr::kable(
-  caption = "Model performance based on the Swedish derivation cohort.
-The area under the receiver operating characteristics curve (AUC) is a measure of discriminatory ability.
-Nagelkerke's pseudo-coefficient of determination (R2) is an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
-The univariable Charlson model had the lowest R2 and was therefore used as baseline for the fraction of new information (FNI) added by the other models.")
-```
 
@@
 
-Both Sweden and Denmark have strived for mass higher education.[@Isopahkala-Bouret2018] This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education.[@zotero-2598] Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ.
-ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark,[@Schmidt2015] compared to 47 % in the Swedish cohort.
+Both Sweden and Denmark have strived for mass higher education[@Isopahkala-Bouret2018].
+This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education[@zotero-2598].
+Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ.
+ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark[@Schmidt2015], compared to 47 % in the Swedish cohort.
 Serious comorbidities such as cancer, are therefore reported equally in both countries, while for example arterial hypertension, is less documented in Denmark.
-Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register.[@Schmidt2015] The RxRisk V index is a weighted index, which is difficult to compare without further analysis.
+Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register[@Schmidt2015].
+The RxRisk V index is a weighted index, which is difficult to compare without further analysis.
 Prescription patterns might differ by indications based on different national guidelines, different subsidization, and different medications sold over-the-counter.
@@
 
-However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere.[@Bunyoz2020]
+However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere[@Bunyoz2020].
 
@@
 
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a] but this comorbidity was not included in our model.
+Cancer seems to confer an increased risk of PJI in several observational studies[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a], but this comorbidity was not included in our model.
 It nevertheless appeared in `r subset(coefs_selected_tab, variable == "Cancer", "n", drop = TRUE)` of the 100 bootstrap replicates (Tab.
 \@ref(tab:brlprop)), and could have been included in a less parsimonious model.
-Neither diabetes was included in our model, but has been in previous risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012]
+Neither diabetes was included in our model, but has been in previous risk calculators[@Wingert2016a; @Tan2018; @Inacio2015].
+On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery[@Namba2012].
 
-Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c,[@Mraovic2011; @Iorio2012] or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model.
+Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c[@Mraovic2011; @Iorio2012], or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model.
 But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
-Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers,[@Gilson2010; @Momohara2011] intra-articular steroid application[@Cancienne2015] or operating time,[@Maoz2015; @Willis-Owen2010; @Berbari2012] are also associated with the risk of PJI after THA.
+Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers[@Gilson2010; @Momohara2011], intra-articular steroid application[@Cancienne2015] or operating time[@Maoz2015; @Willis-Owen2010; @Berbari2012], are also associated with the risk of PJI after THA.
 However, we chose to only consider factors available before surgery and are thus able to use the model as a tool for shared decision making using the implemented web calculator.
@@
 
-We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
-This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig.
-\@ref(fig:aucci) and Tab.
+We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty[@Rasouli2014a; @Inacio2015a], prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
+This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA[@Bulow2017; @Bulow2019].
+However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig.
+\@ref(fig:aucci) and Supplementary Tab.
 \@ref(tab:fni)).
@@
 The selected variables are nevertheless coherent with previous studies.
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons.[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a] Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018] CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI,[@Bozic2012] as have male sex,[@Dale2012; @Wingert2016a; @Tan2018] liver disease [@Deleuran2015] and ASA class.[@Berbari2012; @Pulido2008; @Namba2012]
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a].
+Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018].
+CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI[@Bozic2012], as have male sex[@Dale2012; @Wingert2016a; @Tan2018], liver disease[@Deleuran2015] and ASA class[@Berbari2012; @Pulido2008; @Namba2012].
 
-THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
+THA fixation using cement without antibiotics confers an increased risk of PJI[@Dale2012], but such cement brands were not in clinical use in Sweden during the studied period.
 We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.
@@
 
-# Contribution of authors
+# Acknowledgement
+
+We would like to thank Aurelie Mailhac, statistician at the Department of Clinical Epidemiology at Aarhus university, for help with the external validation.
+
+# Author contributions
 
@@
 
-# Acknowledgement
+# Role of funding sources
 
-We would like to thank Aurelie Mailhac, statistician at the Department of Clinical Epidemiology at Aarhus university, for help with the external validation.
+No external funding
 
-# Data sharing
-
-Access to the model derivation cohort is granted after permission from the Västra Götaland Region, as the central personal data controller agency for the SHAR, in accordance with the European General Data Protection Regulation (GDPR), as well as Swedish laws and regulations (https://registercentrum.se/vara-tjaenster/datauttag/p/B1eeV66QE).
-
-# Disclosure
+# Conflict of interest
 
@@
 
+<!-- TABLES -->
+
+```{r tab1, warning=FALSE}
+source("../lib/clean_names.R")
+
+table1 %>% 
+  mutate(
+    what = replace(what, duplicated(what), "")
+  ) %>% 
+knitr::kable(
+  caption = "Patient characteristics in Sweden 2008-2015 and Denmark 2016 to September 2018.
+Educational levels were classified as low (up to 9 years), middle (10-12 years) and high (more than 12 years).
+Data on hospital levels were not available for the Danish cohort.
+Some additional cells have been intentionally left blank due to small numbers and privacy concerns for the Danish cohort.
+BMI = Body mass index.
+ASA class = American Society for Anaesthesiologists classification.
+CNS = central nervous system.")
+```
+
+```{r coefs}
+tbl_coefs_present %>% 
+  mutate(
+    beta = round(beta, 2)
+  ) %>% 
+knitr::kable(
+  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models.
+BMI = body mass index (under/normal weight as baseline).
+CNS = central nervous system.
+ASA = American Society for Anesthesiologists'
+)
+```
+
+##### PAGE BREAK
+
 # Supplementary
+
+## Supplementary tables
+
+```{r tabcategorization}
+load("../cache/tab_categorization.RData")
+
+tab_categorization %>%
+  ungroup() %>% 
+  mutate(
+    `Comorbidities by groups` = 
+      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
+  ) %>% 
+  knitr::kable(
+    caption = "Categorization of comorbidities from the Charlson and Elixhauser  comorbidities. (CNS = central nervous system.)")
+
+```
+
+```{r infcodes}
+library(coder)
+
+icd10_codes <- 
+  summary(
+      hip_ae_hailer, 
+      "icd10se", 
+      cc_args = list(regex = "icd10", stop = TRUE)
+    )$codes_vct$Infection %>% 
+    paste(collapse = ", ")
+
+nomesco_codes <- 
+  summary(
+      hip_ae_hailer, 
+      "kva", 
+      cc_args = list(regex = "kva", stop = TRUE)
+    )$codes_vct$Infection %>% 
+    paste(collapse = ", ")
+
+tibble(
+  classification = c("ICD-10", "NOMESCO"),
+  codes = c(icd10_codes, nomesco_codes)
+) %>% 
+  knitr::kable(
+    caption = "Codes identifying PJI if recorded in the Swedish and Danish National Patient Registers within 90 days after THA."
+  )
+
+```
+
+```{r brlprop}
+coefs_selected_tab %>% 
+knitr::kable(
+  caption = "Variables selected by the bootstrap ranking procedure.
+Variables selected at least 10 out of 100 times were used in the main model.
+Variables chosen at least 80 times were kept in the reduced model as well.
+(BMI = body mass index.
+ASA class = American Society for Anaesthesiologists classification.
+CNS = central nervous system.)")
+```
+
+```{r fni}
+tab_new_info %>% 
+  arrange(desc(`R2 (%)`)) %>% 
+knitr::kable(
+  caption = "Model performance based on the Swedish derivation cohort.
+The area under the receiver operating characteristics curve (AUC) is a measure of discriminatory ability.
+Nagelkerke's pseudo-coefficient of determination (R2) is an estimate of the proportion of explained variance for each model (the correlation between observed and predicted values).
+The univariable Charlson model had the lowest R2 and was therefore used as baseline for the fraction of new information (FNI) added by the other models.")
+```
 

### 2022-01-15 reports/article.Rmd
initial submission to clinep

--- 2021-06-30 reports/article.Rmd
@@
   C:\\Users\\erik_\\Documents\\library.bib
-csl: bmj.csl
+csl: clinep.csl
 output: 
@@
 
-- EB: Statistician, MSc, PhD, erik.bulow@gu.se
-- ITA: Statistician, MSc, PhD, ita@clin.au.dk
-- UH: Biostatistician, PhD, associate professor, ute@math.au.dk
-- OR: Orthopaedic consultant, PhD, professor, register director, ola.rolfson@gu.se
-- ABP: Orthopaedic consultant, PhD, DMSc, professor, abp@clin.au.dk
-- NPH: Orthopaedic consultant, PhD, professor, nils.Hailer@surgsci.uu.se
+- EB: Statistician, MSc, PhD, 0000-0002-9973-456X, erik.bulow@gu.se
+- ITA: Statistician, MSc, PhD, 0000-0002-8140-1548, ita@clin.au.dk
+- UH: Biostatistician, PhD, associate professor, 000-0001-5059-6293, ute@math.au.dk
+- OR: Orthopaedic consultant, PhD, professor, register director, 0000-0001-6534-1242, ola.rolfson@gu.se
+- ABP: Orthopaedic consultant, PhD, DMSc, professor, 0000-0002-3288-9401, abp@clin.au.dk
+- NPH: Orthopaedic consultant, PhD, professor, 0000-0002-3233-2638, nils.Hailer@surgsci.uu.se
 
@@
 **Guarantor:** NPH, as the principal investigator, affirms that this manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as planned have been explained.
-
-**Word count:** XXX
 
@@
 We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to facilitate shared decision making by patients and surgeons.
-
-##### PAGE BREAK
-
-# Summary box
-
-## What is already known on this topic
-
-- Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA).
-- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia but none of those models has gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.
-
-## What this study adds
-- Our prediction model was derived on a Swedish cohort and externally validated on a Danish cohort
-- Our model is based on easily accessible data that are available before THA
-- Compared to more complex comorbidity indices or to ASA class, our model had superior discriminatory ability and good calibration
-- We provide a simple web calculator for easy use in clinical practice (https://erikbulow.shinyapps.io/thamortpred/).
 
@@
 We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).
-
-## Patient and Public Involvement
-
-Patient representatives are elected as ordinary members of the steering committees for SHAR and DHR, where the development of shared decision making has been discussed.
-Our web calculator will be available for (potential) patients and the public.
 
@@
 
-## Conclusion
+# Conclusion
 
@@
 
+# Abbreviations
+
+ACS NSQIP	Universal American College of Surgeons National Surgical Quality Improvement Program
+ASA		American Society for Anesthesiologists
+ATC		Anatomical Therapeutic Chemical
+AUC		area under the curve 
+BMI		body mass index
+CI		Confidence interval
+DHR		The Danish Hip Arthroplasty Register
+DK		Denmark
+FNI		fraction of new information
+ICD		International Classification of Diseases
+LASSO		least absolute shrinkage and selection operator
+LISA		The longitudinal integration database for health insurance and labour market studies
+NOMESCO		Nordic Medico-Procedural Committee
+NPR		The National Patient Register
+PJI		periprosthetic joint infection
+R2		Nagelkerke's pseudo-coefficient of determination
+ROC		Receiver operating characteristic
+SD		Standard deviation
+SE		Sweden
+SHAR		The Swedish Hip Arthroplasty Register
+THA		total hip arthroplasty
+WHO		World Health Organization
+
 ##### Page break
-
-# Contribution of authors
-
-NPH initiated the study.
-OR managed the ethical review board application in Sweden and ABP in Denmark.
-EB developed the statistical model.
-UH and ITA performed the external validation.
-NPH and EB drafted, and all authors edited and finalized the manuscript.
 
@@
 
-# Disclosure
+# Funding
+
+None
+
+# Conflict of interest
 
 All authors have completed the ICMJE uniform disclosure form at http://icmje.org/downloads/coi_disclosure.docx and declare: no support from any organisation for the submitted work; OR is principal investigator for research partly founded by Pfizer; OR is register director of the Swedish Arthroplasty Register (SAR) and board member of the International Society of Arthroplasty Registries; NPH has received institutional support from Waldemar Link GmbH, and Zimmer Biomet; NPH is president of the Nordic Arthroplasty Register Association, member of the steering committee of SAR, and Co-Editor of Acta Orthopaedica; no other relationships or activities that could appear to have influenced the submitted work.
+
+# Contribution of authors
+
+NPH initiated the study.
+OR managed the ethical review board application in Sweden and ABP in Denmark.
+EB developed the statistical model.
+UH and ITA performed the external validation.
+NPH and EB drafted, and all authors edited and finalized the manuscript.
 
@@
   caption = "Estimated coefficients for the reduced model based on the Swedish derivation cohort (Swedish), as well as re-estimated coefficient values based on the Danish cohort (Danish).
-Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (N = 18,854) as used in the Danish cohort.
+Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (n = 18,854) as used in the Danish cohort.
 The Danish estimates fall within most of the CIs, indicating no support to reject the null hypothesis of no differneces between the countries."

### 2022-01-28 src/fig_roc.R
resubmission to CLinical Epidemiology after review response

@@ -7,7 +7,7 @@ load("cache/models.RData")
 shar_roc <-
   models %>%
   unnest(roc_curve) %>%
-  select(Model, specificity, sensitivity) %>%
+  select(Model, specificity, sensitivity, AUC_est) %>%
   mutate(
     Model = model_names(Model)
   )
@@ -29,25 +29,31 @@ ggdata <-
     Denmark = dk,
     .id = "country"
   ) %>%
-  mutate(country = factor(country, c("Sweden", "Denmark")))
+  mutate(
+    country = factor(country, c("Sweden", "Denmark")),
+    Model = sprintf("%s (%.2f)", Model, AUC_est)
+  )
 
-# Figure ------------------------------------------------------------------
+
+
+# Sweden ------------------------------------------------------------------
 
 ggdata %>%
+  filter(country == "Sweden") %>%
   ggplot(aes(1 - specificity, sensitivity, col = Model)) +
   geom_path() +
   geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
-  theme_minimal(10) +
+  theme_minimal(12) +
   theme(
-    legend.position = "bottom", # c(1, 0),
-    #legend.justification = c(1, 0),
+    legend.position = "bottom",
     legend.title = element_blank(),
-    legend.text = element_text(size = 4.5)
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank()
   ) +
+  guides(col = guide_legend(ncol = 2)) +
   scale_x_continuous(labels = scales::percent) +
-  scale_y_continuous(labels = scales::percent) +
-  facet_wrap(~ country)
+  scale_y_continuous(labels = scales::percent)
 
 
-ggsave("graphs/roc.png", height = 10, width = 15, units = "cm", dpi = 900)
-ggsave("graphs/roc.pdf", height = 10, width = 17, units = "cm")
+ggsave("graphs/roc_se.png", height = 15, width = 15, units = "cm", dpi = 900)
+ggsave("graphs/roc_se.pdf", height = 15, width = 15, units = "cm")

### 2022-01-28 src/fig_separation.R

@@ -11,63 +11,23 @@ obspred <-
   )
 
 
-# Histogram ---------------------------------------------------------------
-
-lims <- c(.002, .2)
-
-fig_separation_hist <-
+g <-
   obspred %>%
   ggplot(aes(pred)) +
-  geom_histogram(
-    aes(fill = obs),
-    alpha = .5, position = "identity",
-    bins = 30
-  ) +
-  theme_minimal(15) +
-  theme(
-    legend.position = "none",
-    panel.grid      = element_blank(),
-    axis.text.x     = element_text(angle = 30),
-    axis.ticks      = element_line(size = 1)
-  ) +
-  xlab("Predicted probability [log]") +
-  ylab(expression(paste(sqrt(n)))) +
-  scale_x_log10(
-    limits = lims,
-    labels = scales::percent
-  ) +
-  scale_y_sqrt(
-    labels = function(x) format(x, big.mark = ",")
-  ) +
-  expand_limits(x = 0)
-
-# Density -----------------------------------------------------------------
-
-fig_separation_density <-
-  obspred %>%
-  ggplot(aes(pred)) +
-  geom_histogram(
-    aes(
-      y = ..density..,
-      fill = obs
-    ),
-    alpha = .5, position = "identity",
-    bins = 30
-  ) +
   geom_density(aes(col = obs), bw = .1) +
-  theme_minimal(15) +
+  theme_minimal(12) +
   theme(
-    legend.position = c(.9,.9),
+    legend.position = c(.1,.9),
     legend.title    = element_blank(),
     panel.grid      = element_blank(),
     axis.text.x     = element_text(angle = 30),
     axis.ticks      = element_line(size = 1),
     axis.ticks.y    = element_line(color = "white"),
-    axis.text.y     = element_text(color = "white", size = 17)     # To align with upper panel
+    axis.text.y     = element_blank()     # To align with upper panel
   ) +
   ylab("Density") +
   scale_x_log10(
-    name = "Predicted probability [log]",
+    name = "Predicted probability [log(10)]",
     limits = lims,
     labels = scales::percent
   ) +
@@ -76,16 +36,5 @@ fig_separation_density <-
   scale_color_discrete(labels = c("no PJI", "PJI"))
 
 
-# Combine plots -----------------------------------------------------------
-
-ggsave(
-  "graphs/separation.png",
[…]

### 2022-01-28 reports/article.Rmd

@@
 ---
-title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark"
-author: Erik Bülow,^1,2^ Ute Hahn,^3,4^ Ina Trolle Andersen,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
+title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty"
+author: Erik Bülow,^1,2^ Ute Hahn,^3,4^ Ina Trolle Andersen,^3^ Ola Rolfson,^1,2^  Alma B. Pedersen,^3,5^ Nils P. Hailer^6^
 date: '`r Sys.Date()`'
@@
 
-1.  The Swedish Arthroplasty Register, Gothenburg, Sweden
+1.  The Swedish Arthroplasty Register, Centre of Registers Västra Götaland, Gothenburg, Sweden
 2.  Department of Orthopaedics, Institute of Clinical Sciences, The Sahlgrenska Academy, University of Gothenburg, Gothenburg, Sweden
-3.  Department of Clinical Epidemiology, Aarhus university hospital, Aarhus, Denmark
-4.  Department of Mathematics, Aarhus University
-5.  Department of Clinical Medicine, Aarhus university
-6.  Department Surgical Sciences/Orthopaedics, Uppsala University Hospital, Uppsala, Sweden
+3.  Department of Clinical Epidemiology, Aarhus University Hospital, Aarhus, Denmark
+4.  Department of Mathematics, Aarhus University, Aarhus, Denmark
+5.  Department of Clinical Medicine, Aarhus University, Aarhus, Denmark
+6.  Department of Surgical Sciences/Orthopaedics, Uppsala University Hospital, Uppsala, Sweden
 
-- EB: Statistician, MSc, PhD, 0000-0002-9973-456X, erik.bulow@gu.se
-- ITA: Statistician, MSc, PhD, 0000-0002-8140-1548, ita@clin.au.dk
-- UH: Biostatistician, PhD, associate professor, 000-0001-5059-6293, ute@math.au.dk
-- OR: Orthopaedic consultant, PhD, professor, register director, 0000-0001-6534-1242, ola.rolfson@gu.se
-- ABP: Orthopaedic consultant, PhD, DMSc, professor, 0000-0002-3288-9401, abp@clin.au.dk
-- NPH: Orthopaedic consultant, PhD, professor, 0000-0002-3233-2638, nils.Hailer@surgsci.uu.se
-
-**Correspondence:** [erik.bulow@gu.se](mailto:erik.bulow@gu.se); +46 70 08 234 28; The Swedish Arthroplasty Register, Registercentrum Västra Götaland, SE-413 45, Sweden
+**Corresponding author:** Erik Bülow; erik.bulow@gu.se; +46 70 08 234 28; The Swedish Arthroplasty Register, Centre of Registers Västra Götaland, SE-413 45 Gothenburg, Sweden
 
@@
 
-**OJECTIVE:** To develop a parsimonious risk prediction model for periprosthetic joint infection (PJI) within 90 days after total hip arthroplasty (THA).
+**PURPOSE:** To develop a parsimonious risk prediction model for periprosthetic joint infection (PJI) within 90 days after total hip arthroplasty (THA).
 
-**DESIGN:** Cohort study.
-
-**SETTING:** Register data from Sweden for model derivation and (from Denmark for external) validation.
-
-**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register operated with elective THA.
-
-**MAIN OUTCOME MEASURE:** Logistic LASSO regression with bootstrap ranking was used to develop a risk prediction model for PJI within 90 days based on the Swedish cohort, followed by external validation on the Danish cohort.
+**PATIENTS AND METHODS:** 
+We used logistic LASSO regression with bootstrap ranking to develop a risk prediction model for PJI within 90 days based on a Swedish cohort of `r nmbr(N_shar)` patients with elective THA 2008-2015. 
+The model was externally validated on a Danish cohort with `r nmbr(N_dk)` patients.
 
 **RESULTS:** Incidence of PJI was `r crude_rate * 100` %  in Sweden and 2.17 % in Denmark.
-A model with the `r coefs_text[["abstract"]]` had an area under the curve (AUC) of `r gsub("AUC = ", "", get_auc("Reduced model (SE)", "2"))` in Sweden and `r sub("^AUC = ", "", get_auc("Reduced model (SE)", "2", Country = "Denmark"))` in Denmark.
+A model with `r coefs_text[["abstract"]]` had an area under the curve (AUC) of `r gsub("AUC = ", "", get_auc("Reduced model (SE)", "2"))` in Sweden and `r sub("^AUC = ", "", get_auc("Reduced model (SE)", "2", Country = "Denmark"))` in Denmark.
 This was superior to traditional models based on ASA class, Charlson, Elixhauser, or the Rx Risk V comorbidity indices.
@@
 
-**CONCLUSION:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated.
+**CONCLUSION:** A new PJI prediction model based on easily accessible data available before THA, was developed and externally validated.
 The model had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices and had good calibration.
 We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to facilitate shared decision making by patients and surgeons.
+
+**KEY WORDS:** Prediction model, total hip arthroplasty, orthopaedics, clinical decision-making tool, web calculator, external validation
 
@@
 
-Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA.
-Both the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices can be used for risk profiling, but their use is rather challenging.
-Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI [@Haynes1995; @Berbari2012].
-The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA.
-The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation.
-We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA that is based on easily accessible data that are available in the setting of clinical decision-making.
+Periprosthetic joint infection (PJI) is one of the most devastating early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, often persistently impaired quality of life, and socioeconomic consequences.[@Zimmerli2006; @Kapadia2016; @Shohat2019; @Malizos2016] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015; @Lespasio2020; @Ren2021; @Resende2021; @Sodhi2020]
+
+Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation.
+
+We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA, a commonly used period to clinically define early PJI.[@Lespasio2020] We based the model on easily accessible data that are available pre-operatively in the setting of clinical decision-making.
 Furthermore, we aimed to validate the model both internally in Sweden, as well as externally on a Danish national cohort.
@@
 
-We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration.
-The best model was then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).
-We followed the TRIPOD checklist for the reporting of the study.
+The derivation cohort was based on patients from an updated linkage database including the Swedish Hip Arthroplasty Register (SHAR).
+The included registers and linkage process has been described previously.[@Cnudde2016] We included all primary elective THAs performed in Sweden 2008-2015.
+Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
+Patients with missing information on educational levels or the type of hospital were also excluded, as were persons under the age of 18.
 
-## Derivation cohort (Sweden)
-
-The derivation cohort was obtained from SHAR `r period`, with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then.
-Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR.
-Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates.
-Data on marital status and the educational level of patients were obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden.
-The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020),[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register.
-Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50.
-Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded, as were children under the age of 18.
-
-## External validation cohort (Denmark)
-
+In addition to the Swedish derivation cohort, a Danish sample was used for external model validation.
 Danish THA patients were operated from 2016 until the last of September 2018.
-This period was chosen due to availability of ASA and BMI data and possibility for all patients to have 90 days of follow up.
-
-All surgery related variables, ASA and BMI were obtained from the DHR.
-The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs.[@Gundtoft2016] All hospitals performing THA in Denmark (40 in 2019) are included in the register.
-The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals.[@Schmidt2015] Data on education and civil status were obtained from Statistics Denmark.[@zotero-2587]
-
-Similar exclusion criteria were applied for diagnosis, ASA class, BMI and bilateral prosthesis.
-Education and type of hospital were not used in the final model and therefore not required for the Danish patients.
-Most people in the Swedish cohort were of age 35 to 99.
-We therefore chose to validate the model for this age group only.
+This period was chosen due to availability of information on ASA and BMI, and the possibility for all patients to have a minimum of 90 days of follow up.
+As in Sweden, data were linked from several national registers based on each patients unique identity number:[@Schmidt2014b] DHR as a quality register for THA,[@Gundtoft2016] the Danish National Patient Register with comorbidity data from all hospitals,[@Schmidt2015] and educational level and civil status from Statistics Denmark.[@zotero-2587] As patients with PJI in the Swedish cohort were of age 35 to 99 years, we chose to validate the model for this age span only.
 
@@
 
-The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery.
-Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab.
-\@ref(tab:tabcategorization)).
+Comorbidities were first identified by ICD-10 codes from each countries' national patient register during the year before surgery.
+Individual conditions were grouped, first according to Elixhauser and Charlson, and then again into broader, clinically relevant, categories.
+Comorbidities observed for less than 10 patients with or without PJI, however,  were not considered as potential predictors, to reduce the risk of overfitting the model to spurious events.
 
-```{r tabcategorization}
-load("../cache/tab_categorization.RData")
+PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital episodes within 90 days after THA (Table  \@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to PJI within the same time frame.
 
-tab_categorization %>%
-  ungroup() %>% 
-  mutate(
-    `Comorbidities by groups` = 
-      gsub("Heart infarct", "Myocardial infarction", `Comorbidities by groups`)
-  ) %>% 
-  knitr::kable(
-    caption = "Categorization of comorbidities from the Charlson and Elixhauser  comorbidities. (CNS = central nervous system.)")
+## Model development
 
+We based a statistical variable selection procedure on bootstrap ranking combined with logistic least absolute shrinkage and selection operator (LASSO) regression, in a way previously described and illustrated by Garland et al.[@Garland2021] The distilled predictors were then used in a multiple main effects logistic regression model, our "main model".
+A more parsimonious "reduced model" was build by keeping only the most influential predictors.
+Those two models were then compared to univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices,[@Kildemoes2011; @Pratt2018] as well as multivariable models combining those measures with age, sex and BMI.
+
+## Model validation
+
+Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals (CI) for each model.
+Predicted probabilities and observed proportions were also plotted with 95 % confidence bands to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model, as was the  fraction of new information for each model compared to the model with lowest $R^2$.[@Harrell2018]
+
+The model was externally validated on the Danish cohort, both with the coefficient estimates from the Swedish cohort, and with new estimates based on the Danish cohort.
+The reference models, with adopted coefficient estimates, were also applied to the Danish cohort for comparison.
+
+## Statistical tools
+
+We used R version 4.1 (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse.
+We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/) and made all R scripts publicly available (XXXXXXXXXXXXX).
+
+## Ethical approval
+
+Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13).
+The study was also approved by the Danish Data Protection Agency (journal number 2015-57-0002) and recorded at Aarhus University (journal number 2016-051-000001).
+
+# Results
+
+```{r}
+age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
+
+comorb <- function(x) {
+  sprintf("%.0f %% (n = %s)", mean(x) * 100, nmbr(sum(x)))
+}
 ```
 
-Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting the model to spurious events.
+We included `r nmbr(N_shar)` patients in the derivation cohort from Sweden (Figure \@ref(fig:flowchartse)).
+`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of these developed a PJI within 90 days.
+In addition, `r nmbr(N_dk)` patients were included in the external validation cohort from Denmark (Figure \@ref(fig:flowchartdk)).
+2.2 % (n = 410) of these developed a PJI within 90 days (Table  \@ref(tab:tab1)).
 
-PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Tab.
-\@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.
+```{r}
+CCI_AUC <- sub("AUC = ", "", get_auc("Charlson"))
+CCI_R2 <-
+  tab_new_info %>% 
+  subset(Model == "Charlson", "R2 (%)", drop = TRUE)
+red_FNI <- 
+  tab_new_info %>% 
+  subset(Model == "Reduced model", "FNI (%)", drop = TRUE) %>% 
+  round()
+```
+
+ROC-curves (Figure \@ref(fig:rocs)) and AUC-values (Figure \@ref(fig:aucci)) were very similar for the main model and a reduced model, based on `r coefs_text[["short"]]`.
+For simplicity, we thus focused on the reduced model, with estimated coefficient values listed in Table \@ref(tab:coefs).
+
+The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2 =$ `r CCI_R2`, and compared to this model, the fraction of new information for the reduced model was `r red_FNI` % (Table  \@ref(tab:fni)).
+The reduced model had good internal calibration for predictions up to 10 % (Figure \@ref(fig:calibration)).
+Estimated probability density curves for patients with and without PJI were partially overlapping, but patients with PJI had on average higher predicted probabilities for this outcome (Figure \@ref(fig:sep)).
+
+The reduced model derived from the Swedish cohort produced a very similar ROC curve when applied to the Danish cohort.
+Calibration was good for predictions up to 3 %, with a tendency to over-estimate higher probabilities, which improved after re-calibration of the model intercept to the Danish PJI incidence (Figure \@ref(fig:calibration)).
+
+# Discussion
+
+## Principal findings
+
+We propose a PJI prediction model based on `r coefs_text[["short"]]`.
+This model was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices, with or without adjustments for age and sex, in terms of AUC, Nagelkerke's $R^2$ and calibration.
+
+## Strengths and weaknesses
+
+Our external model validation is a strength, supporting that the model works equally well in a different population.
+The validation was both geographical (different country) and temporal (different period).
+Discriminatory ability and calibration was good in the validation cohort despite apparent dissimilarities between the countries (Table \@ref(tab:tab1)).
+Some of the dissimilarities can partially be explained by different coding routines.
+ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark,[@Schmidt2015] compared to 47 % in the Swedish cohort.
+Serious comorbidities such as cancer, are reported equally well in both countries, while for example arterial hypertension, is less well documented in Denmark.
+Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less well reported to the national patient register.[@Schmidt2015] The RxRisk V index is a weighted index, which is difficult to compare without further analysis.
+Taken together, the similarities between the countries might be more pronounced than indicated by Table \@ref(tab:tab1).
+However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden, and much more widespread usage of uncemented components in Denmark, is tangible and has been described elsewhere.[@Bunyoz2020]
+
+## Strengths and weaknesses in relation to other studies
+
+The standard peri-operative prophylaxis in THA surgery in Sweden is Cloxacillin (2g x 3) on the day of surgery.
+Those with confirmed or suspected allergy to Cloxacillin (roughly 10% of all patients) receive Clindamycin (600 mg x 2).
+This treatment might affect the outcome of interest, and therefore the possibility to generalize the results to countries with other treatment regimens.
+
+Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a] but this comorbidity was not included in our model.
+Neither was diabetes included in our model, although it is included in previous risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012]
+
+Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c,[@Mraovic2011; @Iorio2012] or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model.
+But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
+Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers,[@Gilson2010; @Momohara2011] intra-articular steroid application[@Cancienne2015] or prolonged operating time,[@Maoz2015; @Willis-Owen2010; @Berbari2012] are also associated with the risk of PJI after THA.
+However, we decided to only consider factors that are available in an out-patient setting prior to surgery and our model can thus be used as a tool for shared decision making.
+This is facilitated by use of the implemented web calculator which we believe is a strength of our model.
+Another strength is the closed form regression formula presented below which is more transparent than a black box model sometimes associated with machine learning and artificial intelligence.
+
+## Implications of the study
+
+Our model can be presented by a logistic regression formula, where the probability of PJI within 90 days is estimated by $1/[1 + \mathrm{exp}(6.31 -\sum x_i\hat\beta_i)]$ where $\sum x_i\hat\beta_i$ indicates the sum of individual patient characteristics multiplied by the estimated coefficient values from Table \@ref(tab:coefs).
+For example, `r examples$desc[[1]]` would have a probability of $1/[1 + \mathrm{exp}(6.31 - 60\cdot0.02)] \approx$ `r examples$p[[1]]` of suffering from a PJI within 90 days.
+For comparison, `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
+Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
+
+It should be noted that the aim of prediction is different from statistical inference.
+As such, CIs in table \@ref(tab:coefs) were based solely on the final logistic regression model, ignoring additional uncertainty from the preceding variable selection.
+Hence, those CIs should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variable.
+Also, the individual odds ratios should not be interpreted as effect measures of individual relations since some variables were correlated.
+The selected variables are nevertheless coherent with previous studies.
+Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons.[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a] Obesity is associated with an increased risk of re-operations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after THA.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018] CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI,[@Bozic2012] as have male sex,[@Dale2012; @Wingert2016a; @Tan2018] liver disease [@Deleuran2015] and ASA class.[@Berbari2012; @Pulido2008; @Namba2012]
+
+## Unanswered questions and future research
+
+The proportion of explained variation (Nagelkerke's $R^2$ = `r subset(tab_new_info, Model == "Reduced model", "R2 (%)", drop = TRUE)` % from Table \@ref(tab:fni)) was low.
+The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
+In addition, a potential underlying susceptibility to PJI -— as has been described for patients with osteomyelitis -— could be a factor of considerable influence on the outcome.
+
+It should also be noticed that our study design was observational, and can only yield (statistical) correlations between PJI and the included variables.
+Additional studies are therefore required to establish causal pathways and to identify possible modifiable risk factors.[@deMast2022]
+
+# Conclusion
+
+Our results indicate that the risk of early PJI after THA can be pre-operatively assessed by a parsimonious prediction model based on patient demographics and a short list of well-defined comorbidities.
+We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and patients in need of THA.
+
+# Abbreviations
+
+- ASA:		American Society for Anesthesiologists
+- AUC:		area under the curve 
+- BMI:		body mass index
+- CI:		Confidence interval
+- DHR:		The Danish Hip Arthroplasty Register
+- DK:		Denmark
+- FNI:		fraction of new information
+- ICD:		International Classification of Diseases
+- LASSO:		least absolute shrinkage and selection operator
+- NOMESCO:		Nordic Medico-Procedural Committee
+- NPR:		The National Patient Register
+- PJI:		periprosthetic joint infection
+- ROC:		Receiver operating characteristic
+- SD:		Standard deviation
+- SE:		Sweden
+- SHAR:		The Swedish Hip Arthroplasty Register
+- THA:		total hip arthroplasty
+
+##### Page break
+
+# Acknowledgement
+
+We would like to thank Aurelie Mailhac, statistician at the Department of Clinical Epidemiology at Aarhus University, for help with the external validation.
+
+# Data sharing
+
+Access to the model derivation cohort is granted after permission from the Västra Götaland Region, as the central personal data controller agency for the SHAR, in accordance with the European General Data Protection Regulation (GDPR), as well as Swedish laws and regulations (https://registercentrum.se/vara-tjaenster/datauttag/p/B1eeV66QE).
+
+# Funding
+
+None
+
+# Conflict of interest
+
+All authors have completed the ICMJE uniform disclosure form at http://icmje.org/downloads/coi_disclosure.docx and declare: no support from any organisation for the submitted work; OR is principal investigator for research partly founded by Pfizer; OR is register director of the Swedish Arthroplasty Register (SAR) and board member of the International Society of Arthroplasty Registries; NPH has received institutional support from Waldemar Link GmbH, and Zimmer Biomet; NPH is president of the Nordic Arthroplasty Register Association, member of the steering committee of SAR, and Co-Editor of Acta Orthopaedica; no other relationships or activities that could appear to have influenced the submitted work.
+
+# Contribution of authors
+
+All authors made a significant contribution to the work reported, whether that is in the conception, study design, execution, acquisition of data, analysis and interpretation, or in all these areas; took part in drafting,  revising or critically reviewing the article; gave final approval of the version to be published; have agreed on the journal to which the article has been submitted; and agree to be accountable for all aspects of the work
+
+##### PAGE BREAK
+
+<!-- FIGURES -->
+
+```{r}
+# MOve figure file to folder for easy access
+fc <- 0 # Figure count
+
+cp_fig <- function(file) {
+  fc <<- fc + 1
+  x <- file.copy(file, paste0("clinep_figs/Figure ", fc, ".pdf"), TRUE)
+}
+```
+
+```{r flowchartse, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation"}
+
+cp_fig("../graphs/flowchart.pdf")
+knitr::include_graphics("../graphs/flowchart.png")
+```
+
+```{r flowchartdk, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Danish Hip Arthroplasty Register were used for external model validation."}
+
+cp_fig("../validation/flowchartdk_detailed.pdf")
+knitr::include_graphics("../validation/dk_flowchart.png")
+```
+
+##### PAGE BREAK
+
+```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
+The main and reduced models performed almost identical for prediction of PJI within 90 days after surgery.
+They both performed better than all other models.
+Area under the curve (AUC) are stated for each curve within parenthesis."}
+
+cp_fig("../graphs/roc_se.pdf")
+knitr::include_graphics("../graphs/roc_se.png")
+```
+
+##### PAGE BREAK
+
+```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % confidence intervals.
+The reduced model performed no different than the main model on the Swedish data, and both of these models performed better than all other models (left panel).
+Similar models were fitted to the Danish cohort (right panel).
+The reduced model (SE) with coefficients based on the Swedish data, performed almost as good as the reduced model (DK) with coefficient values refitted to the Danish cohort."}
+
+cp_fig("../graphs/auc_ci.pdf")
+knitr::include_graphics("../graphs/auc_ci.png")
+```
+
+##### PAGE BREAK
+
+```{r sep, fig.cap = "Patients with observed PJI had, on average, higher predicted probabilities for this adverse event (the x-axis is log-transformed for visual clarity and each curve is normalized on the y-axis which therefore has no direct interpretation)."}
+
+cp_fig("../graphs/separation.pdf")
+knitr::include_graphics("../graphs/separation.png")
+```
+
+##### PAGE BREAK
+
+```{r calibration, fig.cap = "Calibration for the reduced model.
+Proportions of PJI above 10 % were rarely observed and are therefore omitted.
+Internal calibration (95 % confidence band between the green lines) is good; predicted probabilities are similar to observed proportions, as indicated by close proximity to the diagonal line.
+External calibration with the same model (red) indicates some over-estimation and less accuracy.
+This is expected due to smaller sample size in the Danish cohort, and due to national differences.
+Calibration improves after re-calibration of the model intercept to better resemble the Danish base incidence (blue)."}
+
+cp_fig("../graphs/calibration.pdf")
+knitr::include_graphics("../graphs/calibration.png")
+```
+
+##### PAGE BREAK
+
+<!-- TABLES -->
 
@@
 
-## Model development
-
-Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables.
-BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; BMI $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, and BMI $\geq 35$: class II-III obesity).
-
-Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance.
-100 bootstrap samples were then drawn from the observed data set.[@Austin2004]
-
-Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term.
-Regression coefficients were ranked based on their absolute values as a measure of variable importance.
-We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model.
-The second change point was then used as a break-point.
-Influential predictors with absolute coefficients value above this point scored a point and the entire process was repeated 100 times, yielding 10,000 bootstrap samples in total.
-Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of age.
-An even more parsimonious "reduced" model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times was evaluated for comparison.
-Odds ratios for the final models were estimated with 95 % confidence intervals based on the whole derivation cohort.
-
-## Reference models
-
-Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total).
-The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery.
-This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al.[@Pratt2018]
-
-## Model validation
-
-The main, reduced and the 10 reference models were used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation).
-Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed.
-Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model.
-The baseline model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$.
-This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline.[@Harrell2018]
-
-The derived model was then externally validated on the Danish cohort, both with the coefficient estimates from the Swedish cohort and with new estimates based on the Danish cohort.
-The 10 reference models, with adopted coefficient estimates, were also applied to the Danish cohort for comparison.
-Finally, the new coefficient estimates for the derived model, re-estimated with Danish data, were empirically compared to analogues estimates derived from subsamples of the Swedish data with sample sizes to match the Danish cohort (supplementary, Tab \@ref(tab:comparecoefs) and Fig.
-\@ref(fig:coefsspread)).
-
-## Statistical tools
-
-We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse.
-We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).
-
-## Ethical approval
-
-Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (360-13).
-The study was approved by the Danish Data Protection Agency (journal number 2015-57-0002) and recorded at Aarhus University (journal number 2016-051-000001).
-
-# Results
-
-```{r}
-age <- sprintf("%.0f (SD = %.0f)", mean(df$P_Age), sd(df$P_Age))
-
-comorb <- function(x) {
-  sprintf("%.0f %% (n = %s)", mean(x) * 100, nmbr(sum(x)))
-}
-```
-
-## Study participants
-
-We included `r nmbr(N_shar)` patients (`r round(mean(df$P_Sex == "Male") * 100)` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig.
-\@ref(fig:flowchartse)).
-`r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of these developed a PJI within 90 days (Tab.
-\@ref(tab:tab1)).
-The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V.
-
-In addition, `r nmbr(N_dk)` patients (42.1 % males) with a mean age of 70 (SD = 11) years were included in the external validation cohort (Fig.
-\@ref(fig:flowchartdk)).
-2.2 % (n = 410) of these developed a PJI within 90 days (Tab.
-\@ref(tab:tab1)).
-The proportion of patients with ASA class III was 18 % (n = 3,334); 14 % (n = 2,722) had at least one comorbidity according to the Charlson comorbidity index, 22 % (n = 4,101) according to Elixhauser index, and 56 % (n = 10,625) had medications classified according to Rx Risk V.
-
 ```{r tab1, warning=FALSE}
@@
   ) %>% 
+  select(-matches("No PJI")) %>% 
 knitr::kable(
@@
 
-## Model development and internal validation
-
-Patients with `r excl_factors` were uncommon, why these comorbidities were excluded as potential predictors prior to further modelling.
-
-Results from the 100 bootstrap replicates for variable selection were mostly coherent.
-Diagnosis, BMI and some comorbidities where selected each time.
-Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Tab.
-\@ref(tab:brlprop)).
-
-```{r brlprop}
-coefs_selected_tab %>% 
-knitr::kable(
-  caption = "Variables selected by the bootstrap ranking procedure.
-Variables selected at least 10 out of 100 times were used in the main model.
-Variables chosen at least 80 times were kept in the reduced model as well.
-(BMI = body mass index.
-ASA class = American Society for Anaesthesiologists classification.
-CNS = central nervous system.)")
-```
-
-The reduced model (variables with $n \leq 80$ in Tab. \@ref(tab:brlprop)) included `r coefs_text[["short"]]` with estimated coefficient values in Tab. \@ref(tab:coefs).
-
 ```{r coefs}
@@
 knitr::kable(
-  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models.
+  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced model.
 BMI = body mass index (under/normal weight as baseline).
@@
 ```
-
-ROC-curves and AUC-values for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced model.
-
-```{r}
-CCI_AUC <- sub("AUC = ", "", get_auc("Charlson"))
-CCI_R2 <-
-  tab_new_info %>% 
-  subset(Model == "Charlson", "R2 (%)", drop = TRUE)
-red_FNI <- 
-  tab_new_info %>% 
-  subset(Model == "Reduced model", "FNI (%)", drop = TRUE) %>% 
-  round()
-```
-
-The reduced model had an `r get_auc("Reduced model (SE)", "2")`, whereas univariable models with ASA class `r get_auc("ASA")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk")` had less discriminatory abilities (Fig.
-\@ref(fig:rocs) left panel and \@ref(fig:aucci) left panel).
-The univariable model with the Charlson comorbidity index had the lowest AUC and Nagelkerke's $R^2 =$ `r CCI_R2`, and compared to this model, the fraction of new information for the reduced model was `r red_FNI` % (Tab.
-\@ref(tab:fni)).
 
@@
 
-The reduced model had good internal calibration for predictions up to 10 % (Fig.
-\@ref(fig:calibration)).
-Estimated probability density curves for patients with and without PJI were partially overlapping, but patients with PJI had on average higher predicted probabilities for this outcome (Fig.
-\@ref(fig:sep)).
-
-## External validation
-
-The reduced model derived from the Swedish cohort produced a very similar ROC curve when applied to the Danish cohort (Fig.
-\@ref(fig:rocs) right panel), and its AUC value was 0.66 (95% CI: 0.64 to 0.69).
-This did not change for the model with re-estimated coefficient values based on the Danish cohort (0.67, 95 % CI: 0.65 to 0.7).
-AUC values for some of the reference models, estimated on the Danish cohort, did differ from the Swedish cohort, but all those models were less accurate (had broader 95 % CIs) than the reduced model (Fig.
-\@ref(fig:aucci) right panel).
-The re-estimated model coefficients of the reduced model were coherent with the estimates based on the Swedish data (supplementary Tab.
-\@ref(tab:comparecoefs)).
-
-Calibration was good for predictions up to 3 %, with some tendency to over-estimating higher probabilities, which improved after initial re-calibration of the model intercept to the Danish PJI incidence (Fig.
-\@ref(fig:calibration)).
-
-# Discussion
-
-## Principal findings
-
-A model based on `r coefs_text[["short"]]` was better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices, with or without adjustments for age and sex, in terms of AUC, Nagelkerke's $R^2$ and calibration.
-
-## Strengths and weaknesses
-
-Our external validation is a strength, which showed that the model works equally well in Denmark, despite some perceived differences between the countries, partially explained by different coding routines in each country (Tab.
-\@ref(tab:tab1)).
-
-Both Sweden and Denmark have strived for mass higher education.[@Isopahkala-Bouret2018] This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education.[@zotero-2598] Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ.
-ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark,[@Schmidt2015] compared to 47 % in the Swedish cohort.
-Serious comorbidities such as cancer, are therefore reported equally in both countries, while for example arterial hypertension, is less documented in Denmark.
-Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register.[@Schmidt2015] The RxRisk V index is a weighted index, which is difficult to compare without further analysis.
-Prescription patterns might differ by indications based on different national guidelines, different subsidization, and different medications sold over-the-counter.
-Also, the Danish, but not the Swedish, national prescription register, includes prescriptions from nursery homes and other long-term care facilities.
-Taken together, the similarities between the countries might be more pronounced than indicated by Tab.
-\@ref(tab:tab1).
-
-However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere.[@Bunyoz2020]
-
-## Strengths and weaknesses in relation to other studies
-
-Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a] but this comorbidity was not included in our model.
-It nevertheless appeared in `r subset(coefs_selected_tab, variable == "Cancer", "n", drop = TRUE)` of the 100 bootstrap replicates (Tab.
-\@ref(tab:brlprop)), and could have been included in a less parsimonious model.
-Neither diabetes was included in our model, but has been in previous risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012]
-
-Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c,[@Mraovic2011; @Iorio2012] or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model.
-But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings.
-Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers,[@Gilson2010; @Momohara2011] intra-articular steroid application[@Cancienne2015] or operating time,[@Maoz2015; @Willis-Owen2010; @Berbari2012] are also associated with the risk of PJI after THA.
-However, we chose to only consider factors available before surgery and are thus able to use the model as a tool for shared decision making using the implemented web calculator.
-This is a strength of our model.
-
-We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power.
-This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig.
-\@ref(fig:aucci) and Tab.
-\@ref(tab:fni)).
-
-## Meaning of the study
-
-Our model can be presented in a simple formula $\hat p = 1/[1 + \mathrm{exp}(-\hat \alpha - x'\hat \beta)]$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\hat \beta = (\hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k)$ are the estimated intercept and coefficients (Tab.
-\@ref(tab:coefs)) and $x = (x_1, x_2, \ldots, x_k)$ are corresponding patient characteristics.
-For example, `r examples$desc[[1]]` would have a probability of $1/[1 + \mathrm{exp}(6.31 - 60\cdot0.02)] \approx$ `r examples$p[[1]]` of suffering from a PJI within 90 days, whereas `r examples$desc[[2]]` would have a probability of `r examples$p[[2]]`.
-Note, however, that observed proportions above 10 % were rare, and predicted probabilities above this limit are subject to extrapolation.
-
-It should be noted that the aim of prediction is different from statistical inference.
-As such, confidence intervals presented for the model coefficients (Table \@ref(tab:coefs)) were based solely on the final logistic regression model.
-Additional uncertainty introduced by the variable selection process was not incorporated.
-Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variable.
-Also, the individual odds ratios should not be interpreted as effect measures of individual relations since some variables are correlated.
-
-The selected variables are nevertheless coherent with previous studies.
-Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons.[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a] Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018] CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI,[@Bozic2012] as have male sex,[@Dale2012; @Wingert2016a; @Tan2018] liver disease [@Deleuran2015] and ASA class.[@Berbari2012; @Pulido2008; @Namba2012]
-
-THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period.
-We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.
-
-## Unanswered questions and future research
-
-The proportion of explained variation (Nagelkerke's $R^2$ = `r subset(tab_new_info, Model == "Reduced model", "R2 (%)", drop = TRUE)` %) was low.
-The question of what other variables might be needed to further refine prediction models remains open, but information that is notoriously difficult to obtain within the setting of large register studies, such as smoking status, might be of value.
-In addition, a potential underlying susceptibility to PJI -— as has been described for patients with osteomyelitis -— could be a factor of considerable influence on the outcome, but information on single nucleotide polymorphisms in genes that have been associated with musculoskeletal infections was unavailable to us.
-Such shortcomings could be addressed by future genome-wide association studies.
-
-# Conclusion
-
-Our results indicate that the risk of early PJI after THA can be pre-operatively assessed by a parsimonious prediction model.
-We hope that this model, with its accompanying web calculator, will facilitate shared decision-making between physicians and their patients in need of THA.
-
-# Abbreviations
-
-ACS NSQIP	Universal American College of Surgeons National Surgical Quality Improvement Program
-ASA		American Society for Anesthesiologists
-ATC		Anatomical Therapeutic Chemical
-AUC		area under the curve 
-BMI		body mass index
-CI		Confidence interval
-DHR		The Danish Hip Arthroplasty Register
-DK		Denmark
-FNI		fraction of new information
-ICD		International Classification of Diseases
-LASSO		least absolute shrinkage and selection operator
-LISA		The longitudinal integration database for health insurance and labour market studies
-NOMESCO		Nordic Medico-Procedural Committee
-NPR		The National Patient Register
-PJI		periprosthetic joint infection
-R2		Nagelkerke's pseudo-coefficient of determination
-ROC		Receiver operating characteristic
-SD		Standard deviation
-SE		Sweden
-SHAR		The Swedish Hip Arthroplasty Register
-THA		total hip arthroplasty
-WHO		World Health Organization
-
-##### Page break
-
-# Acknowledgement
-
-We would like to thank Aurelie Mailhac, statistician at the Department of Clinical Epidemiology at Aarhus university, for help with the external validation.
-
-# Data sharing
-
-Access to the model derivation cohort is granted after permission from the Västra Götaland Region, as the central personal data controller agency for the SHAR, in accordance with the European General Data Protection Regulation (GDPR), as well as Swedish laws and regulations (https://registercentrum.se/vara-tjaenster/datauttag/p/B1eeV66QE).
-
-# Funding
-
-None
-
-# Conflict of interest
-
-All authors have completed the ICMJE uniform disclosure form at http://icmje.org/downloads/coi_disclosure.docx and declare: no support from any organisation for the submitted work; OR is principal investigator for research partly founded by Pfizer; OR is register director of the Swedish Arthroplasty Register (SAR) and board member of the International Society of Arthroplasty Registries; NPH has received institutional support from Waldemar Link GmbH, and Zimmer Biomet; NPH is president of the Nordic Arthroplasty Register Association, member of the steering committee of SAR, and Co-Editor of Acta Orthopaedica; no other relationships or activities that could appear to have influenced the submitted work.
-
-# Contribution of authors
-
-NPH initiated the study.
-OR managed the ethical review board application in Sweden and ABP in Denmark.
-EB developed the statistical model.
-UH and ITA performed the external validation.
-NPH and EB drafted, and all authors edited and finalized the manuscript.
-
-##### PAGE BREAK
-
-<!-- FIGURES -->
-
-```{r}
-# MOve figure file to folder for easy access
-fc <- 0 # Figure count
-
-cp_fig <- function(file) {
-  fc <<- fc + 1
-  x <- file.copy(file, paste0("bmj_figs/", fc, ".pdf"), TRUE)
-}
-```
-
-```{r flowchartse, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Swedish Hip Arthroplasty Register were used for model derivation and internal validation"}
-
-cp_fig("../graphs/flowchart.pdf")
-knitr::include_graphics("../graphs/flowchart.png")
-```
-
-```{r flowchartdk, fig.cap = "Flowchart with inclusion criteria and number of patients. Data from the Danish Hip Arthroplasty Register were used for external model validation."}
-
-cp_fig("../validation/flowchartdk_detailed.pdf")
-knitr::include_graphics("../validation/dk_flowchart.png")
-```
-
-##### PAGE BREAK
-
-```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models.
-The main and reduced models performed almost identical for prediction of PJI within 90 days after surgery.
-They both performed better than all other models (left panel).
-The reduced model performed as well when applied to the Danish validation cohort (right panel)."}
-
-cp_fig("../graphs/roc.pdf")
-knitr::include_graphics("../graphs/roc.png")
-```
-
-##### PAGE BREAK
-
-```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % confidence intervals (De-Long).
-The reduced model performed no different than the main model on the Swedish data, and both of these models performed better than all other models (left panel).
-Similar models were fitted to the Danish cohort (right panel).
-Coefficient values based on the Swedish data, Reduced model (SE), performed almost as good as the reduced model (DK) with coefficient values refitted to the Danish cohort."}
-
-cp_fig("../graphs/auc_ci.pdf")
-knitr::include_graphics("../graphs/auc_ci.png")
-```
-
-##### PAGE BREAK
-
-```{r sep, fig.cap = "Separation between patients with and without PJI within 90 days versus their predicted probabilities of this event.
-Most patients did not have PJI, wherefore the pink bars dominate the histogram (left panel; note the scales).
-A normalized density plot reveals, however, that patients with PJI had, on average, higher predicted probabilities for this adverse event (right panel)."}
-
-cp_fig("../graphs/separation.pdf")
-knitr::include_graphics("../graphs/separation.png")
-```
-
-##### PAGE BREAK
-
-```{r calibration, fig.cap = "Calibration for the reduced model.
-Proportions of PJI above 10 % were rarely observed and therefore omitted.
-Internal calibration (green 95 % confidence band) is good; predicted probabilities are similar to observed proportions, as indicated by close proximity to the diagonal line.
-External calibration with the same model (pink) indicates some over-estimation and less accuracy.
-This is expected due to smaller sample size in the Danish cohort, and due to national differences.
-Calibration improves after re-calibration of the model intercept to better resamble the Danish base incidence (blue)."}
-
-cp_fig("../graphs/calibration.pdf")
-knitr::include_graphics("../graphs/calibration.png")
-```
-
-##### PAGE BREAK
-
-# Supplementary
-
-## Model coefficients
-
-Variable selection and coefficient estimates were based on the Swedish cohort. 
-The reduced model was then used to predict PJI also in Denmark.
-The coefficient values of the reduced model were also re-fitted to the Danish cohort for comparison.
-We compared the estimates from the Swedish and Danish cohorts and found most values to be of similar magnitude.
-The only coefficient with reversed direction was for patients with a diagnose of "sequelae after childhood hip disease".
-This is a rare condition, with very few observed PJI:s within 90 days.
-We also re-estimated the model coefficients 1,000 times based on the Swedish data.
-Each time, we took a random sample of the same size ($N = 18,854$) as was observed in the Danish cohort.
-The 2.5th and 97.5th percentiles were then used to form empirical 95 % confidence intervals for each coefficient (Tab.
-\@ref(tab:comparecoefs) and Fig.
-\@ref(fig:coefsspread)).
-The estimated effect of "sequelae after childhood hip disease" fall outside this empirical CI.
-Applying multiplicity correction by Bonferroni or similar, would eliminate this significance.
-Hence, it seems that both cohorts are similar in respect to association between the studied covariates and the risk of PJI within 90 days of THA.
-
-```{r comparecoefs}
-compare_se_dk_coefs %>% 
-  rename(`95 % CI` = CI) %>% 
-knitr::kable(
-  caption = "Estimated coefficients for the reduced model based on the Swedish derivation cohort (Swedish), as well as re-estimated coefficient values based on the Danish cohort (Danish).
-Empirical confidence intervals (95 % CI) based on 1,000 resamples from the Swedish cohort of the same sample size (n = 18,854) as used in the Danish cohort.
-The Danish estimates fall within most of the CIs, indicating no support to reject the null hypothesis of no differneces between the countries."
-)
-```
-
-```{r coefsspread, fig.cap = "Estimated coefficient values for the reduced model applied to the Swedish (red triangles) and Danish cohorts (red dots).
-Small coloured dots are empirical estimates from 1000 subsamples based on the Swedish data set, of the same sample size as used in the Danish validation cohort (outliers not shown).
-Black dots indicate the outer limits of 95 % empirical confidence intervals.
-The Danish estimates almost always fall within those CIs."}
-
-cp_fig("../validation/Swedish_model.pdf")
-knitr::include_graphics("../validation/Swedish_model.png")
-```
-
 ##### PAGE BREAK

### 2022-01-28 reports/article.Rmd
manual changes made to Word files before submission

@@
 We used R version 4.1 (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse.
-We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/) and made all R scripts publicly available (XXXXXXXXXXXXX).
+We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/) and made all R scripts publicly available (https://doi.org/10.5281/zenodo.5914599).
 
